---
title: "gene_expression_de_genotype_cellcycle"
author: "Melanie Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc()                             # free up memory and report the memory usage
options(
  max.print        = .Machine$integer.max,
  scipen            = 999,
  stringsAsFactors  = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42)
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)

# Single cell
library(Seurat)

# Plotting
library(ggrepel)
```

# Set project directories and file paths

```{r}
project_name     <- "20251104_sophieWiszniak_ncc_pa"
base_repo_dir    <- "/home/melanie-smith/workDir/sophieWiszniak"
project_repo_dir <- file.path(base_repo_dir, project_name)
out_dir          <- file.path(project_repo_dir, "outDir", "07-gene_expression_de_genotype_cellcycle")

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

# Load required object

- This object has been filtered for doublets and >10% MT genes

```{r}
re_embed_interesting_cells <- readRDS(
  file.path(
    project_repo_dir, "outDir", "06-cluster_marker_genes",
    "final_interesting_cells_0.6.rds"
  )
)
```

# Collapse cell cycle phase into cycling / not_cycling

G1 = not_cycling; S or G2M = cycling

```{r}
re_embed_interesting_cells$cycling_status <- case_when(
  re_embed_interesting_cells@meta.data$Phase == "G1" ~ "not_cycling",
  re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M") ~ "cycling",
  TRUE ~ NA_character_  # catches anything unexpected — check for NAs after
)

cat("\nAny unassigned cells:", sum(is.na(re_embed_interesting_cells$cycling_status)), "\n")
```

# Create composite identity variable

Combines cell type + timepoint + genotype + cycling status into a single
grouping variable for pairwise DE comparisons.

```{r}
re_embed_interesting_cells$celltype.cond.cycle <- paste(
  re_embed_interesting_cells$new_celltypes,
  re_embed_interesting_cells$orig.ident,
  re_embed_interesting_cells$cycling_status,
  sep = "_"
)

Idents(re_embed_interesting_cells) <- "celltype.cond.cycle"

# Preview group sizes — important for catching groups too small for DE
cat("\nCell counts per group:\n")
table(re_embed_interesting_cells$celltype.cond.cycle) %>%
  as.data.frame() %>%
  arrange(Var1) %>%
  as_tibble() %>%
  print(n = Inf)
```

# Define comparison parameters

```{r}
cell_types      <- c("Mesenchyme", "Myocardium", "Endocardium", "Epicardium", "VSMC")
timepoints      <- c("e11", "e12")
cycle_states    <- c("cycling", "not_cycling")

# Significance thresholds — defined once, used consistently throughout
fc_threshold  <- 0.25  # minimum absolute avg_log2FC to call significance
padj_threshold <- 0.05  # adjusted p-value cutoff
top_fc        <- 0.5   # avg_log2FC threshold for labelling top genes on volcano
top_padj      <- 0.01  # adjusted p-value threshold for labelling top genes

# Minimum cells required in each group to attempt DE
min_cells <- 10
```

# Run DE and plot: KO vs control, per cell type, timepoint, and cycling status

For each comparison, `ident.1 = KO` and `ident.2 = control`, so:

- **Positive avg_log2FC** = higher expression in KO
- **Negative avg_log2FC** = higher expression in control

Results tables (CSV) and volcano plots (PNG) are saved per comparison.

```{r, message=FALSE, warning=FALSE}
all_de_results <- list()

for (ct in cell_types) {
  for (tp in timepoints) {
    for (cs in cycle_states) {

      comparison_label <- paste0(ct, "_", tp, "_KO_vs_control_", cs)
      message("Running DE: ", comparison_label)

      ident_ko      <- paste(ct, tp, "ko",      cs, sep = "_")
      ident_control <- paste(ct, tp, "control", cs, sep = "_")

      # --- Check both groups exist ---
      available_idents <- levels(Idents(re_embed_interesting_cells))
      if (!ident_ko %in% available_idents || !ident_control %in% available_idents) {
        warning(
          "Skipping ", comparison_label,
          " — one or both groups not found.\n",
          "  Looking for: ", ident_ko, " and ", ident_control
        )
        next
      }

      # --- Check minimum cell numbers ---
      n_ko      <- sum(Idents(re_embed_interesting_cells) == ident_ko)
      n_control <- sum(Idents(re_embed_interesting_cells) == ident_control)

      if (n_ko < min_cells || n_control < min_cells) {
        warning(
          "Skipping ", comparison_label,
          " — too few cells (KO: ", n_ko, ", control: ", n_control,
          "). Minimum required: ", min_cells
        )
        next
      }

      # --- Run DE ---
      de_result <- FindMarkers(
        re_embed_interesting_cells,
        ident.1         = ident_ko,
        ident.2         = ident_control,
        test.use        = "wilcox",
        min.pct         = 0.1,
        logfc.threshold = 0.25,
        verbose         = FALSE
      )

      # --- Annotate results ---
      de_annotated <- de_result %>%
        rownames_to_column("gene") %>%
        mutate(
          neg_log10_padj = -log10(p_val_adj),
          sig = case_when(
            p_val_adj < padj_threshold & avg_log2FC >  fc_threshold ~ "Up in KO",
            p_val_adj < padj_threshold & avg_log2FC < -fc_threshold ~ "Up in control",
            TRUE ~ "Not significant"
          )
        )

      all_de_results[[comparison_label]] <- de_annotated

      # --- Save DE table ---
      write_csv(
        de_annotated,
        file.path(out_dir, paste0(comparison_label, "_de_results.csv"))
      )

      # Print top hits to notebook
      cat("\n### Top DE genes:", comparison_label,
          "(n KO =", n_ko, "| n control =", n_control, ")\n")
      de_annotated %>%
        filter(sig != "Not significant") %>%
        arrange(desc(abs(avg_log2FC))) %>%
        head(20) %>%
        print()

      # --- Top genes to label on volcano ---
      top_genes <- de_annotated %>%
        filter(p_val_adj < top_padj, abs(avg_log2FC) > top_fc)

      # --- Volcano plot ---
      volcano <- ggplot(de_annotated, aes(x = avg_log2FC, y = neg_log10_padj)) +
        geom_point(aes(color = sig), alpha = 0.5, size = 1) +
        ggrepel::geom_text_repel(
          data         = top_genes,
          aes(label    = gene),
          size         = 3,
          max.overlaps = 20
        ) +
        scale_color_manual(
          values = c(
            "Up in KO"        = "firebrick",
            "Up in control"   = "steelblue",
            "Not significant" = "grey80"
          )
        ) +
        geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", color = "grey50") +
        geom_hline(yintercept = -log10(padj_threshold),         linetype = "dashed", color = "grey50") +
        labs(
          title    = paste0(ct, " ", tp, " (", cs, "): KO vs control"),
          subtitle = paste0(
            "n KO = ", n_ko, "  |  n control = ", n_control,
            "  |  Up in KO: ",      sum(de_annotated$sig == "Up in KO"),
            "  |  Up in control: ", sum(de_annotated$sig == "Up in control")
          ),
          x     = "Average log2 fold change (KO / control)",
          y     = "-log10 adjusted p-value",
          color = NULL
        ) +
        theme_classic()

      # --- Save volcano ---
      ggsave(
        volcano,
        filename = file.path(out_dir, paste0(comparison_label, "_volcano.png")),
        width    = 10,
        height   = 8,
        units    = "in",
        dpi      = 300
      )

      # Print to notebook
      print(volcano)
    }
  }
}

message("All comparisons complete. Outputs saved to: ", out_dir)
```

# Plot: cycling vs not_cycling logFC scatter for all cell types × timepoints

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
for (ct in cell_types) {
  for (tp in timepoints) {

    key_cycling     <- paste0(ct, "_", tp, "_KO_vs_control_cycling")
    key_not_cycling <- paste0(ct, "_", tp, "_KO_vs_control_not_cycling")

    # Check both comparisons exist
    if (!key_cycling %in% names(all_de_results) || !key_not_cycling %in% names(all_de_results)) {
      warning(
        "Skipping ", ct, " ", tp,
        " — one or both DE result sets not found.\n",
        "  Looking for: ", key_cycling, " and ", key_not_cycling
      )
      next
    }

    cycling_de     <- all_de_results[[key_cycling]]
    not_cycling_de <- all_de_results[[key_not_cycling]]

    # Join on gene — inner join keeps only genes tested in both groups
    joined <- inner_join(
      cycling_de     %>% dplyr::select(gene, lfc_cycling     = avg_log2FC, sig_cycling     = sig),
      not_cycling_de %>% dplyr::select(gene, lfc_not_cycling = avg_log2FC, sig_not_cycling = sig),
      by = "gene"
    ) %>%
      dplyr::mutate(
        highlight = case_when(
          sig_cycling != "Not significant" & sig_not_cycling != "Not significant" ~ "Sig in both",
          sig_cycling != "Not significant" & sig_not_cycling == "Not significant" ~ "Sig in cycling only",
          sig_cycling == "Not significant" & sig_not_cycling != "Not significant" ~ "Sig in not_cycling only",
          TRUE ~ "Not significant"
        )
      )

    # Genes to label — significant in at least one group with strong FC
label_genes <- joined %>%
  dplyr::filter(highlight != "Not significant") %>%
  dplyr::filter(abs(lfc_cycling) > 0.5 | abs(lfc_not_cycling) > 0.5)

    p <- ggplot(joined, aes(x = lfc_not_cycling, y = lfc_cycling)) +
      geom_point(aes(color = highlight), alpha = 0.5, size = 1.5) +
      geom_text_repel(
        data         = label_genes,
        aes(label    = gene),
        size         = 3,
        max.overlaps = 20
      ) +
      scale_color_manual(
        values = c(
          "Sig in both"             = "firebrick",
          "Sig in cycling only"     = "darkorange",
          "Sig in not_cycling only" = "steelblue",
          "Not significant"         = "grey80"
        )
      ) +
      geom_hline(yintercept = 0, linetype = "dashed",  colour = "grey50") +
      geom_vline(xintercept = 0, linetype = "dashed",  colour = "grey50") +
      geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey30") +
      labs(
        title    = paste0(ct, " ", tp, ": KO vs control"),
        subtitle = paste0(
          "Genes tested in both groups: ", nrow(joined),
          "  |  Sig in both: ",             sum(joined$highlight == "Sig in both"),
          "  |  Cycling only: ",            sum(joined$highlight == "Sig in cycling only"),
          "  |  Not_cycling only: ",        sum(joined$highlight == "Sig in not_cycling only")
        ),
        x     = "avg_log2FC (not cycling)",
        y     = "avg_log2FC (cycling)",
        color = NULL
      ) +
      theme_classic()

    # Save PNG
    out_file <- file.path(
      out_dir,
      paste0(ct, "_", tp, "_KO_vs_control_cycling_scatter.png")
    )

    ggsave(
      p,
      filename = out_file,
      width    = 10,
      height   = 8,
      units    = "in",
      dpi      = 300
    )

    # Print to notebook
    print(p)
    message("Saved: ", basename(out_file))
  }
}

message("All scatter plots complete. Saved to: ", out_dir)
```

# Summary: comparisons run and DE gene counts

```{r}
summary_table <- map_dfr(names(all_de_results), function(name) {
  df <- all_de_results[[name]]
  tibble(
    comparison    = name,
    total_tested  = nrow(df),
    up_in_ko      = sum(df$sig == "Up in KO"),
    up_in_control = sum(df$sig == "Up in control"),
    not_sig       = sum(df$sig == "Not significant")
  )
})

# Add column: genes significant in both cycling and not_cycling for each celltype x timepoint
sig_in_both <- map_dfr(cell_types, function(ct) {
  map_dfr(timepoints, function(tp) {
    
    key_cycling     <- paste0(ct, "_", tp, "_KO_vs_control_cycling")
    key_not_cycling <- paste0(ct, "_", tp, "_KO_vs_control_not_cycling")
    
    # Return NA if either comparison wasn't run
    if (!key_cycling %in% names(all_de_results) || !key_not_cycling %in% names(all_de_results)) {
      return(tibble(comparison = c(key_cycling, key_not_cycling), n_sig_in_both = NA_integer_))
    }
    
    cycling_sig     <- all_de_results[[key_cycling]]     %>% dplyr::filter(sig != "Not significant") %>% dplyr::pull(gene)
    not_cycling_sig <- all_de_results[[key_not_cycling]] %>% dplyr::filter(sig != "Not significant") %>% dplyr::pull(gene)
    
    n <- length(intersect(cycling_sig, not_cycling_sig))
    
    # Return one row per comparison so it joins cleanly onto summary_table
    tibble(
      comparison    = c(key_cycling, key_not_cycling),
      n_sig_in_both = n
    )
  })
})

summary_table <- summary_table %>%
  dplyr::left_join(sig_in_both, by = "comparison")

print(summary_table)
write_csv(summary_table, file.path(out_dir, "00-DE_summary_all_comparisons.csv"))
```

# Session info

```{r}
sessionInfo()
```
