---
title: "gene_expression_de_genotype_cellcycle"
author: "Melanie Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc()                             # free up memory and report the memory usage
options(
  max.print        = .Machine$integer.max,
  scipen            = 999,
  stringsAsFactors  = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42)
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)

# Single cell
library(Seurat)

# Plotting
library(ggrepel)
```

# Set project directories and file paths

```{r}
project_name     <- "20251104_sophieWiszniak_ncc_pa"
base_repo_dir    <- "/home/melanie-smith/workDir/sophieWiszniak"
project_repo_dir <- file.path(base_repo_dir, project_name)
out_dir          <- file.path(project_repo_dir, "outDir", "07-gene_expression_de_genotype_cellcycle")

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

# Load required object

- This object has been filtered for doublets and >10% MT genes

```{r}
re_embed_interesting_cells <- readRDS(
  file.path(
    project_repo_dir, "outDir", "06-cluster_marker_genes",
    "final_interesting_cells_0.6.rds"
  )
)
```

# Collapse cell cycle phase into cycling / not_cycling

G1 = not_cycling; S or G2M = cycling

```{r}
re_embed_interesting_cells$cycling_status <- case_when(
  re_embed_interesting_cells@meta.data$Phase == "G1" ~ "not_cycling",
  re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M") ~ "cycling",
  TRUE ~ NA_character_  # catches anything unexpected — check for NAs after
)

cat("\nAny unassigned cells:", sum(is.na(re_embed_interesting_cells$cycling_status)), "\n")
```

# Create composite identity variable

Combines cell type + timepoint + genotype + cycling status into a single
grouping variable for pairwise DE comparisons.

```{r}
re_embed_interesting_cells$celltype.cond.cycle <- paste(
  re_embed_interesting_cells$new_celltypes,
  re_embed_interesting_cells$orig.ident,
  re_embed_interesting_cells$cycling_status,
  sep = "_"
)

Idents(re_embed_interesting_cells) <- "celltype.cond.cycle"

# Preview group sizes — important for catching groups too small for DE
cat("\nCell counts per group:\n")
table(re_embed_interesting_cells$celltype.cond.cycle) %>%
  as.data.frame() %>%
  arrange(Var1) %>%
  as_tibble() %>%
  print(n = Inf)
```

# Define comparison parameters

```{r}
cell_types      <- c("Mesenchyme", "Myocardium", "Endocardium", "Epicardium", "VSMC")
timepoints      <- c("e11", "e12")
cycle_states    <- c("cycling", "not_cycling")

# Significance thresholds — defined once, used consistently throughout
fc_threshold  <- 0.25  # minimum absolute avg_log2FC to call significance
padj_threshold <- 0.05  # adjusted p-value cutoff
top_fc        <- 0.5   # avg_log2FC threshold for labelling top genes on volcano
top_padj      <- 0.01  # adjusted p-value threshold for labelling top genes

# Minimum cells required in each group to attempt DE
min_cells <- 10
```

# Run DE and plot: KO vs control, per cell type, timepoint, and cycling status

For each comparison, `ident.1 = KO` and `ident.2 = control`, so:

- **Positive avg_log2FC** = higher expression in KO
- **Negative avg_log2FC** = higher expression in control

Results tables (CSV) and volcano plots (PNG) are saved per comparison.

```{r, message=FALSE, warning=FALSE}
all_de_results <- list()

for (ct in cell_types) {
  for (tp in timepoints) {
    for (cs in cycle_states) {

      comparison_label <- paste0(ct, "_", tp, "_KO_vs_control_", cs)
      message("Running DE: ", comparison_label)

      ident_ko      <- paste(ct, tp, "ko",      cs, sep = "_")
      ident_control <- paste(ct, tp, "control", cs, sep = "_")

      # --- Check both groups exist ---
      available_idents <- levels(Idents(re_embed_interesting_cells))
      if (!ident_ko %in% available_idents || !ident_control %in% available_idents) {
        warning(
          "Skipping ", comparison_label,
          " — one or both groups not found.\n",
          "  Looking for: ", ident_ko, " and ", ident_control
        )
        next
      }

      # --- Check minimum cell numbers ---
      n_ko      <- sum(Idents(re_embed_interesting_cells) == ident_ko)
      n_control <- sum(Idents(re_embed_interesting_cells) == ident_control)

      if (n_ko < min_cells || n_control < min_cells) {
        warning(
          "Skipping ", comparison_label,
          " — too few cells (KO: ", n_ko, ", control: ", n_control,
          "). Minimum required: ", min_cells
        )
        next
      }

      # --- Run DE ---
      de_result <- FindMarkers(
        re_embed_interesting_cells,
        ident.1         = ident_ko,
        ident.2         = ident_control,
        test.use        = "wilcox",
        min.pct         = 0.1,
        logfc.threshold = 0.25,
        verbose         = FALSE
      )

      # --- Annotate results ---
      de_annotated <- de_result %>%
        rownames_to_column("gene") %>%
        mutate(
          neg_log10_padj = -log10(p_val_adj),
          sig = case_when(
            p_val_adj < padj_threshold & avg_log2FC >  fc_threshold ~ "Up in KO",
            p_val_adj < padj_threshold & avg_log2FC < -fc_threshold ~ "Up in control",
            TRUE ~ "Not significant"
          )
        )

      all_de_results[[comparison_label]] <- de_annotated

      # --- Save DE table ---
      write_csv(
        de_annotated,
        file.path(out_dir, paste0(comparison_label, "_de_results.csv"))
      )

      # Print top hits to notebook
      cat("\n### Top DE genes:", comparison_label,
          "(n KO =", n_ko, "| n control =", n_control, ")\n")
      de_annotated %>%
        filter(sig != "Not significant") %>%
        arrange(desc(abs(avg_log2FC))) %>%
        head(20) %>%
        print()

      # --- Top genes to label on volcano ---
      top_genes <- de_annotated %>%
        filter(p_val_adj < top_padj, abs(avg_log2FC) > top_fc)

      # --- Volcano plot ---
      volcano <- ggplot(de_annotated, aes(x = avg_log2FC, y = neg_log10_padj)) +
        geom_point(aes(color = sig), alpha = 0.5, size = 1) +
        ggrepel::geom_text_repel(
          data         = top_genes,
          aes(label    = gene),
          size         = 3,
          max.overlaps = 20
        ) +
        scale_color_manual(
          values = c(
            "Up in KO"        = "firebrick",
            "Up in control"   = "steelblue",
            "Not significant" = "grey80"
          )
        ) +
        geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", color = "grey50") +
        geom_hline(yintercept = -log10(padj_threshold),         linetype = "dashed", color = "grey50") +
        labs(
          title    = paste0(ct, " ", tp, " (", cs, "): KO vs control"),
          subtitle = paste0(
            "n KO = ", n_ko, "  |  n control = ", n_control,
            "  |  Up in KO: ",      sum(de_annotated$sig == "Up in KO"),
            "  |  Up in control: ", sum(de_annotated$sig == "Up in control")
          ),
          x     = "Average log2 fold change (KO / control)",
          y     = "-log10 adjusted p-value",
          color = NULL
        ) +
        theme_classic()

      # --- Save volcano ---
      ggsave(
        volcano,
        filename = file.path(out_dir, paste0(comparison_label, "_volcano.png")),
        width    = 10,
        height   = 8,
        units    = "in",
        dpi      = 300
      )

      # Print to notebook
      print(volcano)
    }
  }
}

message("All comparisons complete. Outputs saved to: ", out_dir)
```

# Summary: comparisons run and DE gene counts

```{r}
summary_table <- map_dfr(names(all_de_results), function(name) {
  df <- all_de_results[[name]]
  tibble(
    comparison    = name,
    total_tested  = nrow(df),
    up_in_ko      = sum(df$sig == "Up in KO"),
    up_in_control = sum(df$sig == "Up in control"),
    not_sig       = sum(df$sig == "Not significant")
  )
})

print(summary_table)

write_csv(summary_table, file.path(out_dir, "00-DE_summary_all_comparisons.csv"))
```

# Session info

```{r}
sessionInfo()
```
