---
title: "gene_expression_de_genotype"
author: "Melanie Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc()                             # free up memory and report the memory usage
options(
  max.print      = .Machine$integer.max,
  scipen         = 999,
  stringsAsFactors = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42)
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)

# Single cell
library(Seurat)

# Plotting
library(patchwork)
library(ggrepel)
```

# Set project directories and file paths

```{r}
project_name    <- "20251104_sophieWiszniak_ncc_pa"
base_repo_dir   <- "/home/melanie-smith/workDir/sophieWiszniak"
project_repo_dir <- file.path(base_repo_dir, project_name)
out_dir         <- file.path(project_repo_dir, "outDir", "07-gene_expression_de_genotype")

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

# Load required object

- This object has been filtered for doublets and >10% MT genes

```{r}
re_embed_interesting_cells <- readRDS(
  file.path(
    project_repo_dir, "outDir", "06-cluster_marker_genes",
    "final_interesting_cells_0.6.rds"
  )
)
```

# Create composite identity variable

Combines cell type, timepoint, and genotype into a single grouping variable
for pairwise DE comparisons.

```{r}
re_embed_interesting_cells$celltype.cond <- paste(
  re_embed_interesting_cells$new_celltypes,
  re_embed_interesting_cells$orig.ident,
  sep = "_"
)

Idents(re_embed_interesting_cells) <- "celltype.cond"
```

# Define comparison parameters

```{r}
cell_types <- c("Mesenchyme", "Myocardium", "Endocardium", "Epicardium", "VSMC")
timepoints <- c("e11", "e12")

# Significance thresholds — defined once and used consistently throughout
fc_threshold   <- 0.25  # minimum absolute avg_log2FC to call significance
padj_threshold <- 0.05  # adjusted p-value cutoff for significance calls
top_fc         <- 0.5   # avg_log2FC threshold for labelling top genes on volcano
top_padj       <- 0.01  # adjusted p-value threshold for labelling top genes on volcano
```

# Run DE and plot: KO vs control, per cell type per timepoint

For each combination, `ident.1 = KO` and `ident.2 = control`, so:

- **Positive avg_log2FC** = higher expression in KO
- **Negative avg_log2FC** = higher expression in control

Results and volcano plots are saved per comparison.

```{r, message=FALSE, warning=FALSE}
# Accumulate all DE results for optional downstream use
all_de_results <- list()

for (ct in cell_types) {
  for (tp in timepoints) {

    comparison_label <- paste0(ct, "_", tp, "_KO_vs_control")
    message("Running DE: ", comparison_label)

    ident_ko      <- paste(ct, tp, "ko",      sep = "_")
    ident_control <- paste(ct, tp, "control", sep = "_")

    # Check both groups exist in the object before running
    available_idents <- levels(Idents(re_embed_interesting_cells))
    if (!ident_ko %in% available_idents || !ident_control %in% available_idents) {
      warning(
        "Skipping ", comparison_label,
        " — one or both groups not found in object.\n",
        "  Looking for: ", ident_ko, " and ", ident_control
      )
      next
    }

    # --- Run DE ---
    de_result <- FindMarkers(
      re_embed_interesting_cells,
      ident.1          = ident_ko,
      ident.2          = ident_control,
      test.use         = "wilcox",
      min.pct          = 0.1,
      logfc.threshold  = 0.25,
      verbose          = FALSE
    )

    # --- Annotate results ---
    de_annotated <- de_result %>%
      rownames_to_column("gene") %>%
      mutate(
        neg_log10_padj = -log10(p_val_adj),
        sig = case_when(
          p_val_adj < padj_threshold & avg_log2FC >  fc_threshold ~ "Up in KO",
          p_val_adj < padj_threshold & avg_log2FC < -fc_threshold ~ "Up in control",
          TRUE ~ "Not significant"
        )
      )

    # Store for later use
    all_de_results[[comparison_label]] <- de_annotated

    # --- Save DE table ---
    write_csv(
      de_annotated,
      file.path(out_dir, paste0(comparison_label, "_de_results.csv"))
    )

    # Print top hits to notebook
    cat("\n### Top DE genes:", comparison_label, "\n")
    de_annotated %>%
      filter(sig != "Not significant") %>%
      arrange(desc(abs(avg_log2FC))) %>%
      head(20) %>%
      print()

    # --- Identify top genes to label on volcano ---
    top_genes <- de_annotated %>%
      filter(p_val_adj < top_padj, abs(avg_log2FC) > top_fc)

    # --- Volcano plot ---
    volcano <- ggplot(de_annotated, aes(x = avg_log2FC, y = neg_log10_padj)) +
      geom_point(aes(color = sig), alpha = 0.5, size = 1) +
      ggrepel::geom_text_repel(
        data        = top_genes,
        aes(label   = gene),
        size        = 3,
        max.overlaps = 20
      ) +
      scale_color_manual(
        values = c(
          "Up in KO"        = "firebrick",
          "Up in control"   = "steelblue",
          "Not significant" = "grey80"
        )
      ) +
      geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", color = "grey50") +
      geom_hline(yintercept = -log10(padj_threshold),         linetype = "dashed", color = "grey50") +
      labs(
        title    = paste0(ct, " ", tp, ": KO vs control"),
        subtitle = paste0(
          "Up in KO: ",      sum(de_annotated$sig == "Up in KO"),
          "  |  Up in control: ", sum(de_annotated$sig == "Up in control")
        ),
        x     = "Average log2 fold change (KO / control)",
        y     = "-log10 adjusted p-value",
        color = NULL
      ) +
      theme_classic()

    # --- Save volcano ---
    ggsave(
      volcano,
      filename = file.path(out_dir, paste0(comparison_label, "_volcano.png")),
      width    = 10,
      height   = 8,
      units    = "in",
      dpi      = 300
    )

    # Print to notebook
    print(volcano)
  }
}

message("All comparisons complete. Outputs saved to: ", out_dir)
```

# Session info

```{r}
sessionInfo()
```
