---
title:  "bootstrap cell type proportions"
author: "Melanie Smith"
date:   "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r clean-up}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(loupeR)

# test cell type proportions
library(speckle)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)
library(ggrepel)
library(ggdist)
library(ComplexHeatmap)
library(circlize)
library(UpSetR)
library(glue)

```

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle", "bootstrap_propeller")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

```

# create the slim object
- I've found this analyses difficult to get through without accumulating enough RAM to crash RStudio.  
- Here I will create a slim version of my Seurat object so that it and all subsequent objects are taking up less space.  

```{r}
# Slim down while keeping exactly what FindMarkers needs
slim_re_embed_interesting_cells <- DietSeurat(
  object = re_embed_interesting_cells,
  
  # Keep only the normalized data layer (this is what FindMarkers uses)
  layers = "data",                    # keep only the required layer
  
  # Keep only the main assay
  assays = "RNA",                     # keep only the required assay
  
  # Keep only the reductions you actually use
  dimreducs = c("pca", "umap"),       # keep only the required reductions
  
  # Drop everything else
  graphs = FALSE
)

# remove the large object
rm(re_embed_interesting_cells)
# clean up
gc()
```


## Set up the cell-type and phase colours for consistenc
- setting the colours here to make sure they are consistent across all scripts.  

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",  
                   "Epicardium"    =  "#FF7F00",
                   "Mesenchyme"    =  "#4DAF4A", 
                   "Endocardium"   = "#377EB8",
                   "Other/Unknown" = "grey70")

phase_cols <- c(
  "G1" = "#1f77b4",
  "G2M" = "#2ca02c",
  "S" = "#ff7f0e"
  )
```

# Quick sanity check

```{r}
print(slim_re_embed_interesting_cells)
# which ident is currently active
table(Idents(slim_re_embed_interesting_cells))
stopifnot(!anyDuplicated(colnames(slim_re_embed_interesting_cells)))
```

# Collaps cell cycle phase into cycling vs not-cycling

```{r}
# set the cell cycle phase as a factor to ensure plots are always in the same order
slim_re_embed_interesting_cells$Phase <- factor(
  slim_re_embed_interesting_cells$Phase,
  levels = c("G1", "G2M", "S")  # match the order in phase_cols exactly
)

## collapse the S and G2M phase allocations
slim_re_embed_interesting_cells[[]] <- slim_re_embed_interesting_cells[[]] %>%
  dplyr::mutate(
    cycling = dplyr::if_else(
      Phase %in% c("S", "G2M"),
      "cycling",
      "not_cycling"
    )
  )

# lets make these a factor so they always plot in the same order
slim_re_embed_interesting_cells$cycling <- factor(
  slim_re_embed_interesting_cells$cycling,
  levels = c("not_cycling", "cycling")
)

# quick sanity check
table(slim_re_embed_interesting_cells$Phase,
      slim_re_embed_interesting_cells$cycling)
  #     cycling not_cycling
  # G1        0       12436
  # G2M    5796           0
  # S      6095           0
```

# =============================================================================
# Bootstrap propeller: cycling vs not_cycling proportions
# Testing: e11 vs e12 within each genotype x cell type
# =============================================================================

```{r bootstrap-setup}
# parameters
N_BOOTSTRAPS    <- 100
N_PSEUDOREPS    <- 3   # pseudo-replicates per stage per bootstrap
set.seed(42)

# define the levels to loop over
genotypes  <- unique(slim_re_embed_interesting_cells$genotype)
celltypes  <- unique(slim_re_embed_interesting_cells$new_celltypes)
stages     <- unique(slim_re_embed_interesting_cells$stage)

# initialise the results list
# structure: bootstrap_results_timepoint[["genotype"]][["celltype"]]
bootstrap_results_timepoint <- setNames(
  lapply(genotypes, function(g) setNames(vector("list", length(celltypes)), celltypes)),
  genotypes
)
```

```{r bootstrap-propeller, message=FALSE, warning=FALSE}
# helper: assign pseudo-replicate labels to a cell barcode vector
# samples with replacement within each stage separately, then assigns
# cells into n_reps groups of equal size (remainder cells are dropped
# to keep groups balanced)
assign_pseudo_replicates_timepoint <- function(meta_subset, n_reps, stages) {

  pseudo_rep_col <- rep(NA_character_, nrow(meta_subset))

  for (stg in stages) {

    idx <- which(meta_subset$stage == stg)

    if (length(idx) < n_reps) {
      warning(glue::glue(
        "Fewer than {n_reps} cells for stage {stg} — skipping this stage in this bootstrap."
      ))
      next
    }

    # sample with replacement to get the same number of cells as we started with
    sampled_idx <- sample(idx, size = length(idx), replace = TRUE)

    # split evenly into n_reps groups (drop remainder to keep balanced groups)
    n_keep       <- floor(length(sampled_idx) / n_reps) * n_reps
    sampled_idx  <- sampled_idx[seq_len(n_keep)]
    rep_labels   <- rep(
      paste0(stg, "_rep", seq_len(n_reps)),
      each = floor(length(sampled_idx) / n_reps)
    )

    # assign back using the *sampled* positions (not original idx)
    # we expand meta_subset rows to match the bootstrap sample
    pseudo_rep_col[sampled_idx] <- rep_labels
  }

  return(pseudo_rep_col)
}

# ---- main bootstrap loop ----------------------------------------------------

for (geno in genotypes) {

  message(glue::glue("\n--- Genotype: {geno} ---"))

  # subset to this genotype once (outside the celltype loop for efficiency)
  geno_cells <- slim_re_embed_interesting_cells[
    , slim_re_embed_interesting_cells$genotype == geno
  ]

  for (ct in celltypes) {

    message(glue::glue("  Cell type: {ct}"))

    # subset to this cell type
    ct_cells <- geno_cells[
      , geno_cells$new_celltypes == ct
    ]

    n_cells <- ncol(ct_cells)

    if (n_cells < (N_PSEUDOREPS * length(stages) * 2)) {
      warning(glue::glue(
        "Too few cells for {geno} / {ct} (n = {n_cells}). Skipping."
      ))
      bootstrap_results_timepoint[[geno]][[ct]] <- list(
        raw_results = NULL,
        summary     = data.frame(skipped = TRUE, n_cells = n_cells)
      )
      next
    }

    # pull the metadata we need once
    ct_meta <- ct_cells@meta.data %>%
      dplyr::select(stage, cycling) %>%
      tibble::rownames_to_column("barcode")

    # storage for this genotype x cell type across bootstraps
    raw_boot_list <- vector("list", N_BOOTSTRAPS)

    for (b in seq_len(N_BOOTSTRAPS)) {

      # ---- 1. assign pseudo-replicates (separately per stage) ----------------
      pseudo_reps <- assign_pseudo_replicates_timepoint(
        meta_subset = ct_meta,
        n_reps      = N_PSEUDOREPS,
        stages      = stages
      )

      ct_meta$pseudo_rep <- pseudo_reps

      # keep only cells that received a pseudo-replicate label (balanced set)
      meta_boot <- ct_meta %>%
        dplyr::filter(!is.na(pseudo_rep))

      if (nrow(meta_boot) < (N_PSEUDOREPS * length(stages))) {
        raw_boot_list[[b]] <- NULL
        next
      }

      # ---- 2. run propeller --------------------------------------------------
      prop_result <- tryCatch(
        {
          speckle::propeller(
            clusters  = meta_boot$cycling,        # cycling vs not_cycling
            sample    = meta_boot$pseudo_rep,     # pseudo-replicates
            group     = meta_boot$stage,          # testing e11 vs e12
            transform = "asin"
          )
        },
        error = function(e) {
          message(glue::glue("    Bootstrap {b} failed: {conditionMessage(e)}"))
          NULL
        }
      )

      if (!is.null(prop_result)) {
        # tag which bootstrap this came from
        prop_result$bootstrap_id <- b
        prop_result$genotype     <- geno
        prop_result$cell_type    <- ct
      }

      raw_boot_list[[b]] <- prop_result

    } # end bootstrap loop

    # ---- 3. compile raw results and append summary --------------------------

    # bind all non-null bootstrap results into one data frame
    raw_results_df <- dplyr::bind_rows(
      Filter(Negate(is.null), raw_boot_list)
    )

    # summary across bootstraps (per cycling category row returned by propeller)
    summary_df <- raw_results_df %>%
      dplyr::group_by(BaselineProp.clusters) %>%    # propeller row = cycling level
      dplyr::summarise(
        n_bootstraps_run       = dplyr::n(),
        n_bootstraps_sig_fdr05 = sum(FDR < 0.05,      na.rm = TRUE),
        prop_sig_fdr05         = mean(FDR < 0.05,     na.rm = TRUE),
        median_FDR             = median(FDR,           na.rm = TRUE),
        median_PropRatio       = median(PropRatio,     na.rm = TRUE),
        CI_FDR_lower           = quantile(FDR, 0.025, na.rm = TRUE),
        CI_FDR_upper           = quantile(FDR, 0.975, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        genotype  = geno,
        cell_type = ct
      )

    # store in the named list
    bootstrap_results_timepoint[[geno]][[ct]] <- list(
      raw_results = raw_results_df,
      summary     = summary_df
    )

    message(glue::glue(
      "    Done. {nrow(raw_results_df)} propeller results stored across bootstraps."
    ))

  } # end cell type loop

} # end genotype loop

message("\nAll bootstraps complete.")
```

```{r bootstrap-save-timepoint}
# save the full named list to disk
saveRDS(
  bootstrap_results_timepoint,
  file = file.path(out_dir, "bootstrap_propeller_cycling_timepoint_results.rds")
)

message(glue::glue("Results saved to: {out_dir}"))
```

```{r bootstrap-quick-peek-timepoint}
# quick sanity check — peek at one combination
example_geno <- genotypes[1]
example_ct   <- celltypes[1]

message(glue::glue("Preview: {example_geno} / {example_ct}"))

# summary table
print(bootstrap_results_timepoint[[example_geno]][[example_ct]]$summary)

# first few rows of raw results
head(bootstrap_results_timepoint[[example_geno]][[example_ct]]$raw_results)
```

# Visualise bootstrap results

```{r extract-raw-results-timepoint}
# extract and combine all raw results into a single long data frame
all_raw_results_timepoint <- purrr::map_dfr(
  genotypes, function(geno) {
    purrr::map_dfr(
      celltypes, function(ct) {
        res <- bootstrap_results_timepoint[[geno]][[ct]]$raw_results
        if (is.null(res)) return(NULL)
        res
      }
    )
  }
)

# ensure cell_type and BaselineProp.clusters are factors for consistent plotting
all_raw_results_timepoint <- all_raw_results_timepoint %>%
  dplyr::mutate(
    cell_type             = factor(cell_type, levels = celltypes),
    BaselineProp.clusters = factor(
      BaselineProp.clusters,
      levels = c("not_cycling", "cycling")
    )
  )

# quick sanity check
glimpse(all_raw_results_timepoint)

summary_table_stage <- all_raw_results_timepoint %>%
  dplyr::group_by(
    BaselineProp.clusters,
    genotype,
    cell_type
  ) %>%
  dplyr::summarise(
    n_bootstraps     = dplyr::n(),
    mean_PropRatio   = mean(PropRatio, na.rm = TRUE),
    median_PropRatio = median(PropRatio, na.rm = TRUE),
    sd_PropRatio     = sd(PropRatio, na.rm = TRUE),
    CI_lower         = quantile(PropRatio, 0.025, na.rm = TRUE),
    CI_upper         = quantile(PropRatio, 0.975, na.rm = TRUE),
    mean_log2PR      = mean(log2(PropRatio), na.rm = TRUE),
    .groups = "drop"
  )
```

```{r plot-pvalue-boxplots-timepoint, fig.width = 10, fig.height = 5}

bootstrap_theme <- theme_bw() +
  theme(
    strip.background   = element_rect(fill = "grey92", colour = "grey70"),
    strip.text         = element_text(size = 11, face = "bold"),
    axis.title         = element_text(size = 11),
    axis.text.x        = element_text(size = 10),
    axis.text.y        = element_text(size = 10),
    legend.title       = element_text(size = 11),
    legend.text        = element_text(size = 10),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank()
  )

# shared plot builder function
make_bootstrap_plot <- function(data, title) {
  
  ggplot(
    data,
    aes(
      x    = log2(PropRatio),
      y    = cell_type,
      fill = BaselineProp.clusters
    )
  ) +
    
    ggdist::stat_halfeye(
      adjust        = 0.7,
      width         = 0.6,
      justification = -0.2,
      .width        = c(0.5, 0.95),
      point_colour  = NA,
      alpha         = 0.8,
      position      = position_dodge(width = 0.7)
    ) +

    # ggdist::stat_dots(
    #   aes(group = BaselineProp.clusters),
    #   side          = "left",
    #   justification = 1.1,
    #   binwidth      = 0.05,
    #   alpha         = 0.75,
    #   dotsize       = 0.25,
    #   position      = position_dodge(width = 0.7)
    # ) +
    
    scale_fill_manual(
      values = c(
        "cycling"     = "#E69F00",
        "not_cycling" = "#56B4E9"
      ),
      labels = c(
        "cycling"     = "Cycling",
        "not_cycling" = "Not cycling"
      ),
      name = "Cell cycle state"
    ) +
    
    geom_vline(
      xintercept = 0,
      linetype   = "dashed",
      colour     = "red"
    ) +
    
    labs(
      title = title,
      x     = expression(Log[2]~"Proportion Ratio (100 bootstraps)"),
      y     = "Cell type"
      ) +
    
    bootstrap_theme
}

# ---- control figure ---------------------------------------------------------
p_control <- all_raw_results_timepoint %>%
  dplyr::filter(genotype == "control") %>%
  make_bootstrap_plot(
    title = "Bootstrap propeller proportions — control (e11 vs e12)"
  )

# save the figure
ggsave(plot = p_control, filename = file.path(out_dir, "bootstrap_timepoint_control.png"))

# ---- Mib1_KO figure ---------------------------------------------------------
p_ko <- all_raw_results_timepoint %>%
  dplyr::filter(genotype == "Mib1_KO") %>%
  make_bootstrap_plot(
    title = "Bootstrap propeller proportions — Mib1_KO (e11 vs e12)"
  )
# save the figure
ggsave(plot = p_ko, filename = file.path(out_dir, "bootstrap_timepoint_ko.png"))

# ---- display ----------------------------------------------------------------
p_control
p_ko
```

# =============================================================================
# Bootstrap propeller: cycling vs not_cycling proportions
# Testing: Mib1-KO - control within each timepoint x cell type
# =============================================================================

```{r bootstrap-setup-stage}
# set the genotype factor levels
slim_re_embed_interesting_cells$genotype <- factor(
  slim_re_embed_interesting_cells$genotype,
  levels = c("Mib1_KO", "control")
  )

# parameters
N_BOOTSTRAPS    <- 100
N_PSEUDOREPS    <- 3   # pseudo-replicates per stage per bootstrap
set.seed(42)

# define the levels to loop over
genotypes  <- unique(slim_re_embed_interesting_cells$genotype)
celltypes  <- unique(slim_re_embed_interesting_cells$new_celltypes)
stages     <- unique(slim_re_embed_interesting_cells$stage)

# initialise the results list
# structure: bootstrap_results[["stage"]][["celltype"]]
bootstrap_results_genotype <- setNames(
  lapply(stages, function(s) setNames(vector("list", length(celltypes)), celltypes)),
  stages
)
```

```{r bootstrap-propeller-stage, message=FALSE, warning=FALSE}
# helper: assign pseudo-replicate labels to a cell barcode vector
# samples with replacement within each stage separately, then assigns
# cells into n_reps groups of equal size (remainder cells are dropped
# to keep groups balanced)
assign_pseudo_replicates_genotype <- function(meta_subset, n_reps, genotypes) {
  
  pseudo_rep_col <- rep(NA_character_, nrow(meta_subset))
  
  for (geno in genotypes) {
    
    idx <- which(meta_subset$genotype == geno)
    
    if (length(idx) < n_reps) {
      warning(glue::glue(
        "Fewer than {n_reps} cells for genotype {geno} — skipping this stage in this bootstrap."
      ))
      next
    }
    
    # sample with replacement to get the same number of cells as we started with
    sampled_idx <- sample(idx, size = length(idx), replace = TRUE)
    
    # split evenly into n_reps groups (drop remainder to keep balanced groups)
    n_keep       <- floor(length(sampled_idx) / n_reps) * n_reps
    sampled_idx  <- sampled_idx[seq_len(n_keep)]
    rep_labels   <- rep(
      paste0(geno, "_rep", seq_len(n_reps)),
      each = floor(length(sampled_idx) / n_reps)
    )
    
    # assign back using the *sampled* positions (not original idx)
    # we expand meta_subset rows to match the bootstrap sample
    pseudo_rep_col[sampled_idx] <- rep_labels
  }
  
  return(pseudo_rep_col)
}

# ---- main bootstrap loop ----------------------------------------------------

for (stg in stages) {
  
  message(glue::glue("\n--- Stage: {stg} ---"))
  
  # subset to this stage once (outside the celltype loop for efficiency)
  stg_cells <- slim_re_embed_interesting_cells[
    , slim_re_embed_interesting_cells$stage == stg
  ]
  
  for (ct in celltypes) {
    
    message(glue::glue("  Cell type: {ct}"))
    
    # subset to this cell type
    ct_cells <- stg_cells[
      , stg_cells$new_celltypes == ct
    ]
    
    n_cells <- ncol(ct_cells)
    
    if (n_cells < (N_PSEUDOREPS * length(genotypes) * 2)) {
      warning(glue::glue(
        "Too few cells for {stg} / {ct} (n = {n_cells}). Skipping."
      ))
      bootstrap_results_genotype[[stg]][[ct]] <- list(
        raw_results = NULL,
        summary     = data.frame(skipped = TRUE, n_cells = n_cells)
      )
      next
    }
    
    # pull the metadata we need once
    ct_meta <- ct_cells@meta.data %>%
      dplyr::select(genotype, cycling) %>%
      tibble::rownames_to_column("barcode")
    
    # storage for this genotype x cell type across bootstraps
    raw_boot_list <- vector("list", N_BOOTSTRAPS)
    
    for (b in seq_len(N_BOOTSTRAPS)) {
      
      # ---- 1. assign pseudo-replicates (separately per stage) ----------------
      pseudo_reps <- assign_pseudo_replicates_genotype(
        meta_subset = ct_meta,
        n_reps      = N_PSEUDOREPS,
        genotypes   = genotypes
        )
      
      ct_meta$pseudo_rep <- pseudo_reps
      
      # keep only cells that received a pseudo-replicate label (balanced set)
      meta_boot <- ct_meta %>%
        dplyr::filter(!is.na(pseudo_rep))
      
      if (nrow(meta_boot) < (N_PSEUDOREPS * length(genotypes))) {
        raw_boot_list[[b]] <- NULL
        next
      }
      
      # ---- 2. run propeller --------------------------------------------------
      prop_result <- tryCatch(
        {
          speckle::propeller(
            clusters  = meta_boot$cycling,        # cycling vs not_cycling
            sample    = meta_boot$pseudo_rep,     # pseudo-replicates
            group     = meta_boot$genotype,       # testing control vs ko
            transform = "asin"
          )
        },
        error = function(e) {
          message(glue::glue("    Bootstrap {b} failed: {conditionMessage(e)}"))
          NULL
        }
      )
      
      if (!is.null(prop_result)) {
        # tag which bootstrap this came from
        prop_result$bootstrap_id <- b
        prop_result$stage        <- stg
        prop_result$cell_type    <- ct
      }
      
      raw_boot_list[[b]] <- prop_result
      
    } # end bootstrap loop
    
    # ---- 3. compile raw results and append summary --------------------------
    
    # bind all non-null bootstrap results into one data frame
    raw_results_df <- dplyr::bind_rows(
      Filter(Negate(is.null), raw_boot_list)
    )
    
    # summary across bootstraps (per cycling category row returned by propeller)
    summary_df <- raw_results_df %>%
      dplyr::group_by(BaselineProp.clusters) %>%    # propeller row = cycling level
      dplyr::summarise(
        n_bootstraps_run       = dplyr::n(),
        n_bootstraps_sig_fdr05 = sum(FDR < 0.05,      na.rm = TRUE),
        prop_sig_fdr05         = mean(FDR < 0.05,     na.rm = TRUE),
        median_FDR             = median(FDR,           na.rm = TRUE),
        median_PropRatio       = median(PropRatio,     na.rm = TRUE),
        CI_FDR_lower           = quantile(FDR, 0.025, na.rm = TRUE),
        CI_FDR_upper           = quantile(FDR, 0.975, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        stage  = stg,
        cell_type = ct
      )
    
    # store in the named list
    bootstrap_results_genotype[[stg]][[ct]] <- list(
      raw_results = raw_results_df,
      summary     = summary_df
    )
    
    message(glue::glue(
      "    Done. {nrow(raw_results_df)} propeller results stored across bootstraps."
    ))
    
  } # end cell type loop
  
} # end genotype loop

message("\nAll bootstraps complete.")
```

```{r bootstrap-save-genotype}
# save the full named list to disk
saveRDS(
  bootstrap_results_genotype,
  file = file.path(out_dir, "bootstrap_propeller_cycling_genotype_results.rds")
)

message(glue::glue("Results saved to: {out_dir}"))
```

```{r bootstrap-quick-peek-genotype}
# quick sanity check — peek at one combination
example_stg <- stages[1]
example_ct  <- celltypes[1]

message(glue::glue("Preview: {example_stg} / {example_ct}"))

# summary table
print(bootstrap_results_genotype[[example_stg]][[example_ct]]$summary)

# first few rows of raw results
head(bootstrap_results_genotype[[example_stg]][[example_ct]]$raw_results)
```

# Visualise bootstrap results

```{r extract-raw-results-genotype}
# extract and combine all raw results into a single long data frame
all_raw_results_genotype <- purrr::map_dfr(
  stages, function(stg) {
    purrr::map_dfr(
      celltypes, function(ct) {
        res <- bootstrap_results_genotype[[stg]][[ct]]$raw_results
        if (is.null(res)) return(NULL)
        res
      }
    )
  }
)

# ensure cell_type and BaselineProp.clusters are factors for consistent plotting
all_raw_results_genotype <- all_raw_results_genotype %>%
  dplyr::mutate(
    cell_type             = factor(cell_type, levels = celltypes),
    BaselineProp.clusters = factor(
      BaselineProp.clusters,
      levels = c("not_cycling", "cycling")
    )
  )

# quick sanity check
glimpse(all_raw_results_genotype)

summary_table_genotype <- all_raw_results_genotype %>%
  dplyr::group_by(
    BaselineProp.clusters,
    stage,
    cell_type
  ) %>%
  dplyr::summarise(
    n_bootstraps     = dplyr::n(),
    mean_PropRatio   = mean(PropRatio, na.rm = TRUE),
    median_PropRatio = median(PropRatio, na.rm = TRUE),
    sd_PropRatio     = sd(PropRatio, na.rm = TRUE),
    CI_lower         = quantile(PropRatio, 0.025, na.rm = TRUE),
    CI_upper         = quantile(PropRatio, 0.975, na.rm = TRUE),
    mean_log2PR      = mean(log2(PropRatio), na.rm = TRUE),
    .groups = "drop"
  )
```

```{r plot-pvalue-boxplots-genotype, fig.width = 10, fig.height = 5}

# bootstrap_theme <- theme_bw() +
#   theme(
#     strip.background   = element_rect(fill = "grey92", colour = "grey70"),
#     strip.text         = element_text(size = 11, face = "bold"),
#     axis.title         = element_text(size = 11),
#     axis.text.x        = element_text(size = 10),
#     axis.text.y        = element_text(size = 10),
#     legend.title       = element_text(size = 11),
#     legend.text        = element_text(size = 10),
#     legend.position    = "bottom",
#     panel.grid.major.x = element_blank()
#   )
# 
# # shared plot builder function
# make_bootstrap_plot <- function(data, title) {
#   
#   ggplot(
#     data,
#     aes(
#       x    = log2(PropRatio),
#       y    = cell_type,
#       fill = BaselineProp.clusters
#     )
#   ) +
#     
#     ggdist::stat_halfeye(
#       adjust        = 0.7,
#       height        = 0.6,
#       justification = -0.2,
#       .width        = c(0.5, 0.95),
#       point_colour  = NA,
#       alpha         = 0.8,
#       position      = position_dodge(width = 0.7)
#     ) +
#     
#     scale_fill_manual(
#       values = c(
#         "cycling"     = "#E69F00",
#         "not_cycling" = "#56B4E9"
#       ),
#       labels = c(
#         "cycling"     = "Cycling",
#         "not_cycling" = "Not cycling"
#       ),
#       name = "Cell cycle state"
#     ) +
#     
#     geom_vline(
#       xintercept = 0,
#       linetype   = "dashed",
#       colour     = "red"
#     ) +
#     
#     labs(
#       title = title,
#       y     = "Cell type",
#       x     = expression(Log[2]~"Proportion Ratio (100 bootstraps)")
#     ) +
#     
#     bootstrap_theme
# }

# ---- control figure ---------------------------------------------------------
p_e11 <- all_raw_results_genotype %>%
  dplyr::filter(stage == "e11") %>%
  make_bootstrap_plot(
    title = "Bootstrap propeller proportions — e11 (Mib1-KO - Control)"
  )

# save the figure
ggsave(plot = p_e11, filename = file.path(out_dir, "bootstrap_genotype_e11.png"))

# ---- Mib1_KO figure ---------------------------------------------------------
p_e12 <- all_raw_results_genotype %>%
  dplyr::filter(stage == "e12") %>%
  make_bootstrap_plot(
    title = "Bootstrap propeller proportions — e12 (Mib1-KO - Control)"
  )
# save the figure
ggsave(plot = p_e12, filename = file.path(out_dir, "bootstrap_genotype_e12.png"))

# ---- display ----------------------------------------------------------------
p_e11
p_e12
```


