---
title: "08-dreamlet_analysis_cycling_orig.ident_pseudobulk"
author: "Melanie Smith"
date: "2026-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Based on the code found [here](https://github.com/GabrielHoffman/dreamlet_analysis/blob/devel/Mathys_2023/Mathys_2023_Major_Cell_Type.Rmd)

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Load required libraries

```{r}
# if (!require("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }

# BiocManager::install("dreamlet")

# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(SingleCellExperiment)
library(dreamlet)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)

# additional table functions
library(kableExtra)
# annotation help
library(homologene)

```

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
data_dir <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes")
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}
```

# Cell cycle gene sets (mouse)

Cell cycle gene lists are re-derived here from `cc.genes.updated.2019` via
human → mouse ortholog conversion. This is the same conversion used in the
upstream pre-processing script.

```{r cc-gene-sets}
human_genes_to_mouse <- function(human_genes) {
  hom_data <- homologene::homologene(human_genes, inTax = 9606, outTax = 10090)
  setNames(hom_data$`10090`, hom_data$`9606`)
}

s_genes   <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$s.genes)
g2m_genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$g2m.genes)
cc_genes  <- c(s_genes, g2m_genes)

cat("S-phase genes retained: ",   length(s_genes),   "\n")
cat("G2M-phase genes retained: ", length(g2m_genes), "\n")
cat("Total CC genes: ",           length(cc_genes),  "\n")
```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(data_dir, "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
# and is a factor
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

# add collapsed cycling phase to the metadata
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = ifelse(
    re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M"),
    "cycling",
    "not_cycling"
  ),
  col.name = "cycling"
)

# add collapsed seurat cluster + proliferative status to the metadata
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = paste0(
    re_embed_interesting_cells@meta.data$orig.ident,
    "_",
    re_embed_interesting_cells@meta.data$cycling
  ),
  col.name = "orig.ident.cycling"
  )

re_embed_interesting_cells@meta.data %>%
  dplyr::count(orig.ident.cycling) %>%
  dplyr::arrange(as.numeric(stringr::str_extract(orig.ident.cycling, "^[0-9]+")))
```

# Trim down the Seurat object metadata -> keep only what we need
- There a are a bunch of columns in the metadata that we don't need at all for this analysis and it makes the sce object and pseudibulked object more complicated so I want to reduce the noise here and just keep the columns we use downstream.

```{r}
# Check what you currently have
names(re_embed_interesting_cells@meta.data)

# Define columns to KEEP (easier than listing what to remove)
cols_to_keep <- c(
  "orig.ident",
  "seurat_clusters",
  "new_celltypes",
  "stage",
  "genotype",
  "Phase",
  "S.Score",
  "G2M.Score",
  "cycling",
  "orig.ident.cycling"
  )

# Remove everything not in that list
cols_to_remove <- setdiff(names(re_embed_interesting_cells@meta.data), cols_to_keep)
re_embed_interesting_cells@meta.data[, cols_to_remove] <- NULL

# Verify
names(re_embed_interesting_cells@meta.data)
```

# make some umaps
- Let's remind ourselves where the object is at

```{r}
# make some umaps
# DimPlot(re_embed_interesting_cells,
#         reduction = "umap",
#         group.by = "seurat_clusters",
#         label = TRUE,
#         label.size = 6) +
#   NoLegend()
# 
# DimPlot(re_embed_interesting_cells,
#         reduction = "umap",
#         group.by = "cycling",
#         label = TRUE,
#         label.size = 6) +
#   NoLegend()
# 
# DimPlot(re_embed_interesting_cells,
#         reduction = "umap",
#         group.by = "new_celltypes",
#         label = TRUE,
#         label.size = 6) +
#   NoLegend()


# make some summary tables
table(re_embed_interesting_cells@meta.data$seurat_clusters,
      re_embed_interesting_cells@meta.data$new_celltypes)

# Build the table
cc_table <- table(
  re_embed_interesting_cells@meta.data$seurat_clusters,
  re_embed_interesting_cells@meta.data$cycling
)

# Add percent cycling column
cc_table_df <- as.data.frame.matrix(cc_table) %>%
  tibble::rownames_to_column("seurat_clusters") %>%
  dplyr::mutate(
    total       = cycling + not_cycling,
    pct_cycling = round(cycling / total * 100, 1)
  ) %>%
  dplyr::arrange(as.numeric(seurat_clusters))

cc_table_df

```

# standardise metadata types

```{r}
# Convert categorical variables to factors
re_embed_interesting_cells$orig.ident <-
  factor(re_embed_interesting_cells$orig.ident)

re_embed_interesting_cells$orig.ident.cycling <-
  factor(re_embed_interesting_cells$orig.ident.cycling)

re_embed_interesting_cells$seurat_clusters <-
  factor(re_embed_interesting_cells$seurat_clusters)

re_embed_interesting_cells$stage <-
  factor(re_embed_interesting_cells$stage)

re_embed_interesting_cells$genotype <-
  factor(re_embed_interesting_cells$genotype)

re_embed_interesting_cells$cycling <-
  factor(
    re_embed_interesting_cells$cycling,
    levels = c("not_cycling", "cycling")
  )

```

# Convert Seurat to SCE object

```{r}
# Convert Seurat object
sce <- Seurat::as.SingleCellExperiment(
  re_embed_interesting_cells,
  assay = "RNA"
  )

# Check assays
assayNames(sce)

# Check metadata survived conversion
colnames(colData(sce))

# inspect
assayNames(sce)
colData(sce)

# make sure cell type is a sorted factor
# sce$new_celltypes = factor(sce$new_celltypes, sort(levels(sce$new_celltypes)))
sce$new_celltypes <- factor(sce$new_celltypes)
```

# create the pseudobulk object

```{r pseudobulk}
# create pseudobulk 
pb <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "new_celltypes",  
    sample_id = "orig.ident.cycling")
```

# UMAP embedding

```{r embedding}
# make sure the umap embedding survived the conversion to sce
reducedDimNames(sce)
# plot
plotProjection(sce, "UMAP", annotation='new_celltypes') 
plotProjection(sce, "UMAP", annotation='orig.ident.cycling')
```

## Properties of dataset

```{r properties}
df = data.frame(table(pb$stage))
colnames(df) = c("Embryonic Stage", "Count")

df %>%
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 


df = data.frame(table(pb$cycling))
colnames(df) = c("Cycling Status", "Count")

df %>%
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 

df = list()
df$`# Samples` = ncol(pb)
df$`# Subjects` = nlevels(sce$orig.ident.cycling)
df$`# Cells` = ncol(sce)
df = unlist(df)
df = data.frame(Propert=names(df), count = df)

df %>%  
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 
```

## Process pseudobulk data to estimate precision weights
# ~ cycling

```{r voom}
types = c("cycling", "not_cycling" )
pb$cycling = factor(pb$cycling, types)

types = c("e11", "e12")
pb$stage = factor(pb$stage, types)

# Normalize and apply voomWithDreamWeights
# form = ~ age + sex + Study + pmi
form <- ~ cycling
res.proc <- processAssays(
    pb,
    form,
    min_cells = 5,
    min.count = 5,
    min.samples = 3
    )
```

### Show details of processing

```{r details}
details(res.proc) 
```

### Show voom-style mean-variance trends

```{r voom.plot, fig.height=6, fig.width=12}
# show voom plot for each cell clusters  
plotVoom(res.proc, 
         ncol=2) 
```

# Variance partitioning analysis
## ~ cycling

```{r vp}
# run variance partitioning analysis
vp.list <- fitVarPart(res.proc, ~ cycling)
```

```{r vp.plot, fig.height=6, fig.width=12}
# Summarize variance fractions genome-wide for each cell type
plotVarPart(sortCols(vp.list),
            label.angle=60,
            ncol=4)   
plotPercentBars(vp.list)

# all genes ordered by variance explained
all_genes_cycling <- rownames(vp.list)[order(vp.list$cycling, decreasing = TRUE)]
plotPercentBars(vp.list[all_genes_cycling, ]) +
  ggtitle("Percent variance explained: All genes",
          subtitle = "") +
  theme(legend.position = "right")

# Plot per-gene percent bars for the top 5 genes
# Sort by genotype variance
top_genes_cycling <- rownames(vp.list)[order(vp.list$cycling, decreasing = TRUE)][1:5000]
plotPercentBars(vp.list[top_genes_cycling, ])

# plot the cell cycle genes
plotPercentBars(vp.list,
                genes = c(s_genes, g2m_genes)) +
  ggtitle("Percent variance explained: S phase and G2M genes",
          subtitle = "") +
  theme(legend.position = "right")

# Make one bar plot per cell type with each cell type ordered by the highest variance
celltypes <- unique(vp.list$assay)

percent_bar_plots <- list()

for (celltype_of_interest in celltypes) {
  
  vp_ct <- vp.list[vp.list$assay == celltype_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  cc_genes <- c(s_genes, g2m_genes)
  vp_ct_cc <- vp_ct[vp_ct$gene %in% cc_genes, ]
  
  cc_rank <- order(vp_ct_cc[, "cycling"], decreasing = TRUE)
  
  p <- plotPercentBars(vp_ct_cc[cc_rank, ]) +
    ggtitle(paste("Percent variance explained:", celltype_of_interest))
  
  percent_bar_plots[[celltype_of_interest]] <- p
}

# summary object of cell cycle genes per cell type
cc_genes <- c(s_genes, g2m_genes)

vp_df <- as.data.frame(vp.list)

cellcycle_variance_matrix <- vp_df %>%
  filter(gene %in% cc_genes) %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

cellcycle_variance_matrix

all_genes_variance_matrix <- vp_df %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

# let's heatmap that sucker to see if anything stands out
gene_order <- order(cellcycle_variance_matrix[, "Endocardium"], decreasing = TRUE)

pheatmap(cellcycle_variance_matrix[gene_order, ],
         main = "Cell-cycle variance explained by cell type: Endocardium order",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

# Make one bar plot per cell type with each cell type ordered by the highest variance
celltypes <- unique(vp.list$assay)

percent_bar_plots <- list()

for (celltype_of_interest in celltypes) {
  
  vp_ct <- vp.list[vp.list$assay == celltype_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  gene_order <- order(vp_ct[, "cycling"], decreasing = TRUE)
  
  p <- plotPercentBars(vp_ct[gene_order, ]) +
    ggtitle(paste("Percent variance explained:", celltype_of_interest))
  
  percent_bar_plots[[celltype_of_interest]] <- p
}

# top 10 genes per cell type
celltypes <- unique(vp.list$assay)
percent_bar_plots <- list()

for (celltype_of_interest in celltypes) {
  
  vp_ct <- vp.list[vp.list$assay == celltype_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  gene_order <- order(vp_ct[, "cycling"], decreasing = TRUE)
  
  # keep only the top 10
  top20_order <- gene_order[1:100]
  
  p <- plotPercentBars(vp_ct[top20_order, ]) +
    ggtitle(paste("Percent variance explained:", celltype_of_interest))
  
  percent_bar_plots[[celltype_of_interest]] <- p
}
  
```

### Consensus genes
- generate a list of genes who's variance is >=0.95 cell cycle associated.  
- then, because some genes will be NA (failed VP regression), manually inspect the variance matrix.  

```{r}


# library(tidyverse)
# library(patchwork)
library(ggVennDiagram)

# convert the matrix to a long data.frame for convenience where needed
all_genes_variance_df <- all_genes_variance_matrix %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "celltype", values_to = "variance")

ggplot(all_genes_variance_df,
       aes(x = variance,
           fill = celltype)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~celltype, ncol = 2) +
  labs(
    x = "Variance explained by cycling",
    y = "Density",
    title = "Distribution of cell-cycle variance by cell type"
  ) +
  theme(legend.position = "none")

# create a vector of genes where the variance is >=0.95 in every cell type
# note drops a gene if any cell type is NA (we will look at these manually later)
consensus_genes_95_celltype <- all_genes_variance_df %>%
  group_by(gene) %>%
  filter(all(!is.na(variance) & variance >= 0.95)) %>%
  ungroup() %>%
  distinct(gene) %>%
  pull(gene)

# as a bit of a positive control check we can take a look at the variance associated with only the cell cycle genes (S.Phase and G2M.Phase)
# convert the cellcycle_variance_matrix matrix (only contains cell cycle genes) to a long data.frame for convenience where needed
s_g2m_genes_variance_df <- cellcycle_variance_matrix %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "celltype", values_to = "variance")

# return a list object with one slot per cell type. Each cell type contains only genes that are >=0.95 and not NA
gene_sets_95 <- lapply(colnames(cellcycle_variance_matrix), function(ct) {
  all_genes_variance_df %>%
    filter(celltype == ct, variance >= 0.95) %>%
    pull(gene)
  })
names(gene_sets_95) <- colnames(cellcycle_variance_matrix)

ggVennDiagram(gene_sets_95) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Genes with \u2265 0.95 cell-cycle variance explained, by cell type")

# take the 106 genes in the consensus set and denote membership in cell cycle phase
consensus_cellcycle_check <- tibble(gene = consensus_genes_95_celltype) %>%
  mutate(
    gene_set = case_when(
      gene %in% s_genes   ~ "S phase",
      gene %in% g2m_genes ~ "G2M",
      TRUE ~ "Not cell cycle"
    )
  )

table(consensus_cellcycle_check$gene_set)
cat("There are", sum(consensus_genes_95_celltype %in% s_genes), "of the possible", length(s_genes), "s_genes and", sum(consensus_genes_95_celltype %in% g2m_genes), "of the possible", length(g2m_genes), "G2M_genes in the consensus")

sum(!consensus_genes_95_celltype %in% c(s_genes, g2m_genes))

other_celltypes <- setdiff(colnames(cellcycle_variance_matrix), "Endocardium")

# create a set of scatter plots to compare the variance % for the cell cycle (S.Phase and G2M.Phase genes)
scatter_plots <- lapply(other_celltypes, function(ct) {
  ggplot(as.data.frame(cellcycle_variance_matrix),
         aes(x = Endocardium, y = .data[[ct]])) +
    geom_point(alpha = 0.4, size = 0.8) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(
      x = "Endocardium (cycling variance)",
      y = paste0(ct, " (cycling variance)"),
      title = paste("Endocardium vs", ct, "(cell cycle genes only)")
    )
  })

wrap_plots(scatter_plots, ncol = 2)

full_variance_matrix <- vp_df %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

dim(full_variance_matrix)   # [1] 16348     5

# create a matrix of consensus genes where all variance values are >=0.95 or NA
full_variance_matrix_95 <- full_variance_matrix[
  apply(full_variance_matrix, 1, function(x) {
    !all(is.na(x)) && all(x >= 0.95, na.rm = TRUE)
  }),
  ]

dim(full_variance_matrix_95)

other_celltypes <- setdiff(colnames(full_variance_matrix), "Endocardium")

scatter_plots_full <- lapply(other_celltypes, function(ct) {
  ggplot(as.data.frame(full_variance_matrix_95),
         aes(x = Endocardium, y = .data[[ct]])) +
    geom_point(alpha = 0.2, size = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      x = "Endocardium (cycling variance)",
      y = paste0(ct, " (cycling variance)"),
      title = paste("Endocardium vs", ct, "(all genes >=0.95 or NA)")
    )
})

wrap_plots(scatter_plots_full, ncol = 2)

```

### missed NA genes
- here I take a manual look at the variance matrix and select genes that are >=0.95 but are missing from the consensus set due to NAs

```{r}
# let's remind ourselves where the cycling genes are
DimPlot(re_embed_interesting_cells,
        reduction = "umap",
        group.by = "cycling") +
  labs(title = "UMAP coloured by cycling")

# View the full matrix
View(full_variance_matrix)

# select a gene of interest after viewing
gene_of_interest <- "Echdc3"

# feature plot all combined
FeaturePlot(re_embed_interesting_cells,
            features = gene_of_interest)
# feature plot split by phase so we can see where each of the genes is expressed
FeaturePlot(re_embed_interesting_cells,
            features = gene_of_interest,
            split.by = "Phase")

# make a (labourious) manual list of lost genes
lost_genes <- c("Gtse1", "Rad51ap1", "Pask", "Prr11", "Mybl2", "Knl1", "Ddias", "Fancb", "Mcm10", "Ttk", "Polq", "Tedc1", "Fancd2", "Mastl", "Espl1", "E2f2", "Cdc45", "Nsl1", "Kifc1", "Pbk", "Cdca5", "Iqgap3", "Ska3", "Arhgap19", "Depdc1a", "Cenpi", "Cdc6", "Ube2t", "Rad54b", "Atf7ip2", "Aurkb", "Nuf2", "Cep55", "Rmi2", "Ska1", "Lrr1", "Sapcd2", "Pimreg", "BC030867", "E2f8", "Neil3", "Pif1", "Ticrr", "Recql4", "Xkr5", "Mtfr2", "Kif14", "Cenpl", "Cit", "Kif18b", "Cdc25c", "Ankle1", "Hist1h2ab", "Oip5", "Haspin", "Troap", "Esco2", "Spag5", "Kifc5b", "Arhgef39", "Hist1h2bh", "2810408I11Rik", "Hist1h2bb", "Rtkn2", "Hist1h2bj", "Gm3550", "Hist1h2bp", "Aunip", "Tnfaip8l1", "Hist1h3g", "Fanci", "Hist1h2bf", "Gas2l3", "Hist1h3f", "Hist1h2bl")

# how many of these genes are in the original cell cycle sets?
sum(lost_genes %in% c(s_genes, g2m_genes))
# how many genes in total?
length(lost_genes)

# we can check the expression of individual genes if we want to
expr_values <- FetchData(re_embed_interesting_cells, vars = gene_of_interest)
# make an average of the expression
avg_expr_table <- expr_values %>%
  mutate(celltype = re_embed_interesting_cells$new_celltypes) %>%
  group_by(celltype) %>%
  summarise(
    mean_expression = mean(.data[[gene_of_interest]]),
    median_expression = median(.data[[gene_of_interest]]),
    n_cells = n(),
    .groups = "drop"
  )
# an take a look
avg_expr_table
```

### Add consensus gene lists and plot

```{r}
# quick sanity check to make sure the lost genes don't overlap with the consensus set
sum(lost_genes %in% consensus_genes_95_celltype)
# [1] 0

full_consensus <- c(lost_genes, consensus_genes_95_celltype)

# how many of these genes are in the original cell cycle sets?
sum(full_consensus %in% c(s_genes, g2m_genes)) # we end up with about half of the cell cycle genes


```

## Heatmap of the top cell cycle associated genes
### Full consensus gene set

```{r}
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = paste(re_embed_interesting_cells@meta.data$new_celltypes,
                   re_embed_interesting_cells@meta.data$Phase,
                   sep = "_"),
  col.name = "celltype_phase"
  )

colnames(re_embed_interesting_cells@meta.data)
table(re_embed_interesting_cells$celltype_phase)

# pull average expression per cell type for these genes
avg_expr_celltypes_phase <- AverageExpression(
  re_embed_interesting_cells,
  features = full_consensus,
  group.by = "celltype_phase"
)

expr_mat_celltypes_phase <- as.matrix(avg_expr_celltypes_phase$RNA)

# row annotation data.frame - rownames must match expr_mat rownames exactly
row_annotation <- full_consensus_cellcycle_check %>%
  filter(gene %in% rownames(expr_mat_celltypes_phase)) %>%
  column_to_rownames("gene")

# reorder expr_mat to match annotation row order (defensive - AverageExpression can drop genes not found)
expr_mat_cell <- expr_mat_celltypes_phase[rownames(row_annotation), ]
# plot the heatmap
pheatmap(
  expr_mat_cell,
  scale = "row",                     # z-score each gene across cell types, easier to compare shape not magnitude
  annotation_row = row_annotation,
  annotation_colors = list(
    gene_set = c(`S phase` = "#d95f02", `G2M` = "#7570b3", `Not cell cycle` = "grey80")
    ),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  cutree_cols = 3,
  show_rownames = TRUE,
  fontsize_row = 6,
  main = "Expression of consensus cell-cycle-associated genes by cell type + phase"
  )

# save the heatmap as a PDF (pheatmap's built-in filename argument handles sizing automatically)
pheatmap(
  expr_mat_cell,
  scale = "row",
  annotation_row = row_annotation,
  annotation_colors = list(
    gene_set = c(`S phase` = "#d95f02", `G2M` = "#7570b3", `Not cell cycle` = "grey80")
  ),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  cutree_cols = 3,
  show_rownames = TRUE,
  fontsize_row = 6,
  main = "Expression of consensus cell-cycle-associated genes by cell type + phase",
  filename = file.path(out_dir, "heatmap_consensus_celltype_phase.pdf"),
  width = 8,
  height = 20
)


# save the underlying expression matrix as a csv
expr_mat_cell %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  write_csv("out_dir/expr_mat_celltype_phase.csv")


```

### Official cell cycle gene sets
- S.Phase
- G2M.Phase

```{r}

# pull average expression per cell type for these genes
avg_expr_celltypes_phase_cc <- AverageExpression(
  re_embed_interesting_cells,
  features = cc_genes,
  group.by = "celltype_phase"
)

expr_mat_celltypes_phase_cc <- as.matrix(avg_expr_celltypes_phase_cc$RNA)

# row annotation data.frame - rownames must match expr_mat rownames exactly
row_annotation <- full_consensus_cellcycle_check %>%
  filter(gene %in% rownames(expr_mat_celltypes_phase_cc)) %>%
  column_to_rownames("gene")

# reorder expr_mat to match annotation row order (defensive - AverageExpression can drop genes not found)
expr_mat_cell_cc <- expr_mat_celltypes_phase_cc[rownames(row_annotation), ]

pheatmap(
  expr_mat_cell_cc,
  # scale = "row",                     # z-score each gene across cell types, easier to compare shape not magnitude
  annotation_row = row_annotation,
    annotation_colors = list(
    gene_set = c(`S phase` = "#d95f02", `G2M` = "#7570b3")
  ),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  cutree_cols = 3,
  show_rownames = TRUE,
  fontsize_row = 6,
  main = "Expression of official cell-cycle-associated genes by cell type + phase"
)

```

```{r}
vp_df <- as.data.frame(vp.list)
top_n <- 5000

# top N genes *per cell type*, based on cycling variance, then union across cell types
top_genes <- vp_df %>%
  group_by(assay) %>%
  slice_max(order_by = cycling, n = top_n) %>%
  pull(gene) %>%
  unique()

# gene x cell type matrix of cycling variance, restricted to that gene set
heatmap_matrix <- vp_df %>%
  filter(gene %in% top_genes) %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

# row annotation: which cell-cycle gene set each gene belongs to
row_annotation <- data.frame(
  gene_set = case_when(
    rownames(heatmap_matrix) %in% s_genes   ~ "S phase",
    rownames(heatmap_matrix) %in% g2m_genes ~ "G2M",
    TRUE ~ "Other"
  ),
  row.names = rownames(heatmap_matrix)
)

heatmap_matrix[is.na(heatmap_matrix)] <- 2

gene_order <- order(heatmap_matrix[, "Endocardium"], decreasing = TRUE)

pheatmap(heatmap_matrix[gene_order, ],
         annotation_row = row_annotation,
         main = "Top 100 cell-cycle-associated genes per cell type",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = FALSE,
         show_colnames = TRUE)

sum(s_genes %in% top_genes)
cat("There are", sum(s_genes %in% top_genes), "of the", length(s_genes), "S.Score genes in the top 100")
sum(g2m_genes %in% top_genes)
cat("There are", sum(g2m_genes %in% top_genes), "of the", length(g2m_genes), "G2M.Score genes in the top 100")

# or as a table
table(row_annotation$gene_set)

df <- as.data.frame(heatmap_matrix) %>%
  tibble::rownames_to_column("genes") %>%
  dplyr::left_join(., tibble::rownames_to_column(row_annotation, "genes"))
```


## Process pseudobulk data to estimate precision weights
## ~ cycling + stage

```{r voom}
types = c("cycling", "not_cycling" )
pb$cycling = factor(pb$cycling, types)

types = c("e11", "e12")
pb$stage = factor(pb$stage, types)

# Normalize and apply voomWithDreamWeights
# form = ~ age + sex + Study + pmi
form.2 <- ~ cycling + stage
res.proc.2 <- processAssays(
    pb,
    form.2,
    min_cells = 5,
    min.count = 5,
    min.samples = 3
    )
```

### Show details of processing

```{r details}
details(res.proc.2) 
```

### Show voom-style mean-variance trends

```{r voom.plot, fig.height=6, fig.width=12}
# show voom plot for each cell clusters  
plotVoom(res.proc.2, 
         ncol=2) 
```

```{r vp}
# run variance partitioning analysis
vp.list.2 <- fitVarPart(res.proc.2, ~ cycling + stage)
```

```{r vp.plot, fig.height=6, fig.width=12}
# Summarize variance fractions genome-wide for each cell type
plotVarPart(sortCols(vp.list.2),
            label.angle=60,
            ncol=4)   

# Plot per-gene percent bars for the top 5 genes
# Sort by genotype variance
top_genes_cycling <- rownames(vp.list.2)[order(vp.list.2$cycling, decreasing = TRUE)][1:5000]
plotPercentBars(vp.list.2[top_genes_cycling, ])

# plot the cell cycle genes
plotPercentBars(vp.list.2,
                genes = c(s_genes, g2m_genes))
```

# Session Info

```{r session-info}
sessioninfo::session_info()
```