---
title: "08-dreamlet_analysis_orig.ident_cycling_stage_pseudobulk"
author: "Melanie Smith"
date: "2026-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Based on the code found [here](https://github.com/GabrielHoffman/dreamlet_analysis/blob/devel/Mathys_2023/Mathys_2023_Major_Cell_Type.Rmd)

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Load required libraries

```{r}
# if (!require("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }

# BiocManager::install("dreamlet")

# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(SingleCellExperiment)
library(dreamlet)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)

# additional table functions
library(kableExtra)
# annotation help
library(homologene)

```

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
data_dir <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes")
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}
```

# Cell cycle gene sets (mouse)

Cell cycle gene lists are re-derived here from `cc.genes.updated.2019` via
human → mouse ortholog conversion. This is the same conversion used in the
upstream pre-processing script.

```{r cc-gene-sets}
human_genes_to_mouse <- function(human_genes) {
  hom_data <- homologene::homologene(human_genes, inTax = 9606, outTax = 10090)
  setNames(hom_data$`10090`, hom_data$`9606`)
}

s_genes   <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$s.genes)
g2m_genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$g2m.genes)
cc_genes  <- c(s_genes, g2m_genes)

cat("S-phase genes retained: ",   length(s_genes),   "\n")
cat("G2M-phase genes retained: ", length(g2m_genes), "\n")
cat("Total CC genes: ",           length(cc_genes),  "\n")
```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(data_dir, "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
# and is a factor
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

# add collapsed cycling phase to the metadata
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = ifelse(
    re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M"),
    "cycling",
    "not_cycling"
  ),
  col.name = "cycling"
)

# add collapsed seurat cluster + proliferative status to the metadata
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = paste0(
    re_embed_interesting_cells@meta.data$orig.ident,
    "_",
    re_embed_interesting_cells@meta.data$cycling
  ),
  col.name = "orig.ident.cycling"
  )

re_embed_interesting_cells@meta.data %>%
  dplyr::count(orig.ident.cycling) %>%
  dplyr::arrange(as.numeric(stringr::str_extract(orig.ident.cycling, "^[0-9]+")))
```

# Trim down the Seurat object metadata -> keep only what we need
- There a are a bunch of columns in the metadata that we don't need at all for this analysis and it makes the sce object and pseudibulked object more complicated so I want to reduce the noise here and just keep the columns we use downstream.

```{r}
# Check what you currently have
names(re_embed_interesting_cells@meta.data)

# Define columns to KEEP (easier than listing what to remove)
cols_to_keep <- c(
  "orig.ident",
  "seurat_clusters",
  "new_celltypes",
  "stage",
  "genotype",
  "Phase",
  "S.Score",
  "G2M.Score",
  "cycling",
  "orig.ident.cycling"
  )

# Remove everything not in that list
cols_to_remove <- setdiff(names(re_embed_interesting_cells@meta.data), cols_to_keep)
re_embed_interesting_cells@meta.data[, cols_to_remove] <- NULL

# Verify
names(re_embed_interesting_cells@meta.data)
```

# make some umaps
- Let's remind ourselves where the object is at

```{r}

# make some summary tables
table(re_embed_interesting_cells@meta.data$seurat_clusters,
      re_embed_interesting_cells@meta.data$new_celltypes)

# Build the table
cc_table <- table(
  re_embed_interesting_cells@meta.data$seurat_clusters,
  re_embed_interesting_cells@meta.data$cycling
)

# Add percent cycling column
cc_table_df <- as.data.frame.matrix(cc_table) %>%
  tibble::rownames_to_column("seurat_clusters") %>%
  dplyr::mutate(
    total       = cycling + not_cycling,
    pct_cycling = round(cycling / total * 100, 1)
  ) %>%
  dplyr::arrange(as.numeric(seurat_clusters))

cc_table_df

```

# standardise metadata types

```{r}
# Convert categorical variables to factors
re_embed_interesting_cells$orig.ident <-
  factor(re_embed_interesting_cells$orig.ident)

re_embed_interesting_cells$orig.ident.cycling <-
  factor(re_embed_interesting_cells$orig.ident.cycling)

re_embed_interesting_cells$seurat_clusters <-
  factor(re_embed_interesting_cells$seurat_clusters)

re_embed_interesting_cells$stage <-
  factor(re_embed_interesting_cells$stage)

re_embed_interesting_cells$genotype <-
  factor(re_embed_interesting_cells$genotype)

re_embed_interesting_cells$cycling <-
  factor(
    re_embed_interesting_cells$cycling,
    levels = c("not_cycling", "cycling")
  )

```

# Convert Seurat to SCE object

```{r}
# Convert Seurat object
sce <- Seurat::as.SingleCellExperiment(
  re_embed_interesting_cells,
  assay = "RNA"
  )

# Check assays
assayNames(sce)

# Check metadata survived conversion
colnames(colData(sce))

# inspect
assayNames(sce)
colData(sce)

# make sure cell type is a sorted factor
# sce$new_celltypes = factor(sce$new_celltypes, sort(levels(sce$new_celltypes)))
sce$new_celltypes <- factor(sce$new_celltypes)
```

# create the pseudobulk object

```{r pseudobulk}
# create pseudobulk 
pb <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "stage",  
    sample_id = "orig.ident.cycling")
```

# UMAP embedding

```{r embedding}
# make sure the umap embedding survived the conversion to sce
reducedDimNames(sce)
# plot
plotProjection(sce, "UMAP", annotation='new_celltypes') 
plotProjection(sce, "UMAP", annotation='orig.ident.cycling')
```

## Properties of dataset

```{r properties}
df = data.frame(table(pb$orig.ident))
colnames(df) = c("orig.ident", "Count")

df %>%
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 


df = data.frame(table(pb$cycling))
colnames(df) = c("Cycling Status", "Count")

df %>%
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 

df = list()
df$`# Samples` = ncol(pb)
df$`# Subjects` = nlevels(sce$orig.ident.cycling)
df$`# Cells` = ncol(sce)
df = unlist(df)
df = data.frame(Propert=names(df), count = df)

df %>%  
  kbl(row.names=FALSE) %>% 
  kable_classic(full_width = FALSE) 
```

## Process pseudobulk data to estimate precision weights
# ~ cycling

```{r voom}
types = c("cycling", "not_cycling" )
pb$cycling = factor(pb$cycling, types)

types = c("control", "Mib1_KO")
pb$genotype = factor(pb$genotype, types)

# Normalize and apply voomWithDreamWeights
# form = ~ age + sex + Study + pmi
form <- ~ cycling
res.proc <- processAssays(
    pb,
    form,
    min_cells = 5,
    min.count = 5,
    min.samples = 3
    )
```

### Show details of processing

```{r details}
details(res.proc) 
```

### Show voom-style mean-variance trends

```{r voom.plot, fig.height=6, fig.width=12}
# show voom plot for each cell clusters  
plotVoom(res.proc, 
         ncol=2) 
```

# Variance partitioning analysis
## ~ cycling

```{r vp}
# run variance partitioning analysis
vp.list <- fitVarPart(res.proc, ~ cycling)
```

```{r vp.plot, fig.height=6, fig.width=12}
# Summarize variance fractions genome-wide for each cell type
plotVarPart(sortCols(vp.list),
            label.angle=60,
            ncol=4)   
plotPercentBars(vp.list)

# all genes ordered by variance explained
all_genes_cycling <- rownames(vp.list)[order(vp.list$cycling, decreasing = TRUE)]
plotPercentBars(vp.list[all_genes_cycling, ]) +
  ggtitle("Percent variance explained: All genes",
          subtitle = "") +
  theme(legend.position = "right")

# Plot per-gene percent bars for the top 5 genes
# Sort by genotype variance
top_genes_cycling <- rownames(vp.list)[order(vp.list$cycling, decreasing = TRUE)][1:5000]
plotPercentBars(vp.list[top_genes_cycling, ])

# plot the cell cycle genes
plotPercentBars(vp.list,
                genes = c(s_genes, g2m_genes)) +
  ggtitle("Percent variance explained: S phase and G2M genes",
          subtitle = "") +
  theme(legend.position = "right")

# Make one bar plot per cell type with each cell type ordered by the highest variance
stages <- unique(vp.list$assay)

percent_bar_plots <- list()

for (stage_of_interest in stages) {
  
  vp_ct <- vp.list[vp.list$assay == stage_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  cc_genes <- c(s_genes, g2m_genes)
  vp_ct_cc <- vp_ct[vp_ct$gene %in% cc_genes, ]
  
  cc_rank <- order(vp_ct_cc[, "cycling"], decreasing = TRUE)
  
  p <- plotPercentBars(vp_ct_cc[cc_rank, ]) +
    ggtitle(paste("Percent variance explained cc genes:", stage_of_interest))
  
  percent_bar_plots[[stage_of_interest]] <- p
}

# summary object of cell cycle genes per cell type
cc_genes <- c(s_genes, g2m_genes)

vp_df <- as.data.frame(vp.list)

cellcycle_variance_matrix <- vp_df %>%
  filter(gene %in% cc_genes) %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

cellcycle_variance_matrix
# super quick xy correlation plot
ggplot(cellcycle_variance_matrix,
       aes(x = e11, y = e12)) +
  geom_point()

# let's heatmap that sucker to see if anything stands out
gene_order <- order(cellcycle_variance_matrix[, "e11"], decreasing = TRUE)

pheatmap(cellcycle_variance_matrix[gene_order, ],
         main = "Cell-cycle variance explained by stage: e11 order",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

# Make one bar plot per time point with each time point ordered by the highest variance
stages <- unique(vp.list$assay)

percent_bar_plots <- list()

for (stages_of_interest in stages) {
  
  vp_ct <- vp.list[vp.list$assay == stages_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  gene_order <- order(vp_ct[, "cycling"], decreasing = TRUE)
  
  p <- plotPercentBars(vp_ct[gene_order, ]) +
    ggtitle(paste("Percent variance explained (all genes):", stages_of_interest))
  
  percent_bar_plots[[stages_of_interest]] <- p
}

all_genes_variance_matrix <- vp_df %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

# super quick xy correlation plot
ggplot(all_genes_variance_matrix,
       aes(x = e11, y = e12)) +
  geom_point()

# top 50 genes per cell type
stages <- unique(vp.list$assay)
percent_bar_plots <- list()

for (stages_of_interest in stages) {
  
  vp_ct <- vp.list[vp.list$assay == stages_of_interest, ]
  
  # order genes by variance explained *within this cell type only*
  gene_order <- order(vp_ct[, "cycling"], decreasing = TRUE)
  
  # keep only the top 10
  top20_order <- gene_order[1:50]
  
  p <- plotPercentBars(vp_ct[top20_order, ]) +
    ggtitle(paste("Percent variance explained:", stages_of_interest))
  
  percent_bar_plots[[stages_of_interest]] <- p
}
  
vp_df_e11 <- dplyr::filter(vp_df, assay == "e11")
ggplot(vp_df_e11, aes(x = cycling)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  geom_vline(xintercept = 0.95,
             colour = "red",
             linetype = "dashed") +
  labs(
    x = "Variance explained by cycling",
    y = "Density",
    title = "Density of cell-cycle variance explained (E11.5)"
  )
cat("in e11, there are", dim(dplyr::filter(vp_df_e11, cycling >= 0.95))[1], "genes at >= 0.95")

vp_df_e12 <- dplyr::filter(vp_df, assay == "e12")
ggplot(vp_df_e12, aes(x = cycling)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  geom_vline(xintercept = 0.95,
             colour = "red",
             linetype = "dashed") +
  labs(
    x = "Variance explained by cycling",
    y = "Density",
    title = "Density of cell-cycle variance explained (E12.5)"
  )
cat("in e12, there are", dim(dplyr::filter(vp_df_e12, cycling >= 0.95))[1], "genes at >= 0.95")
```

# Consensus gene set

```{r}
consensus_genes_95 <- vp_df_e11 %>%
  select(gene, cycling_e11 = cycling) %>%
  inner_join(
    vp_df_e12 %>% select(gene, cycling_e12 = cycling),
    by = "gene"
  ) %>%
  filter(cycling_e11 >= 0.95, cycling_e12 >= 0.95)

consensus_genes_95

consensus_genes_95 <- consensus_genes_95 %>%
  mutate(
    gene_set = case_when(
      gene %in% s_genes   ~ "S phase",
      gene %in% g2m_genes ~ "G2M",
      TRUE ~ "Not cell cycle"
    )
  )

consensus_genes_95

# quick scatter plot to see the correlation
ggplot(consensus_genes_95, aes(x = cycling_e11, y = cycling_e12, colour = gene_set)) + geom_point()
table(consensus_genes_95$gene_set)

# venn diagram of genes >=95%
genes_e11_95 <- vp_df_e11 %>%
  filter(cycling >= 0.95) %>%
  pull(gene)

genes_e12_95 <- vp_df_e12 %>%
  filter(cycling >= 0.95) %>%
  pull(gene)

# let's make a venn diagram...
ggVennDiagram(
  list(E11.5 = genes_e11_95, E12.5 = genes_e12_95)
) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Genes with \u2265 0.95 cell-cycle variance explained")

cat("of the", length(consensus_genes_95$gene), "consensus genes, there are", table(consensus_genes_95$gene_set)[[1]], "of the possible", length(g2m_genes), "G2M.Score genes and", table(consensus_genes_95$gene_set)[[3]], "of the possible", length(s_genes), "S.Score genes in the consensus gene set")

# lets save the consensus set so we can compare to the cell type version of this
write_csv(consensus_genes_95, file = file.path(out_dir, "consensus_genes_95_stage.csv"))
```

## Heatmap of the top cell cycle associated genes

```{r}
vp_df <- as.data.frame(vp.list)
top_n <- 500

# top N genes *per cell type*, based on cycling variance, then union across cell types
top_genes <- vp_df %>%
  group_by(assay) %>%
  slice_max(order_by = cycling, n = top_n) %>%
  pull(gene) %>%
  unique()

# gene x cell type matrix of cycling variance, restricted to that gene set
heatmap_matrix <- vp_df %>%
  filter(gene %in% top_genes) %>%
  select(gene, assay, cycling) %>%
  pivot_wider(names_from = assay, values_from = cycling) %>%
  column_to_rownames("gene") %>%
  as.matrix()

# row annotation: which cell-cycle gene set each gene belongs to
row_annotation <- data.frame(
  gene_set = case_when(
    rownames(heatmap_matrix) %in% s_genes   ~ "S phase",
    rownames(heatmap_matrix) %in% g2m_genes ~ "G2M",
    TRUE ~ "Other"
  ),
  row.names = rownames(heatmap_matrix)
)

heatmap_matrix[is.na(heatmap_matrix)] <- 2

gene_order <- order(heatmap_matrix[, "e11"], decreasing = TRUE)

pheatmap(heatmap_matrix[gene_order, ],
         annotation_row = row_annotation,
         main = paste0("Top (", top_n, ") cell-cycle-associated genes per stage"),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = FALSE,
         show_colnames = TRUE)

sum(s_genes %in% top_genes)
cat("There are", sum(s_genes %in% top_genes), "of the", length(s_genes), "S.Score genes in the top", top_n)
sum(g2m_genes %in% top_genes)
cat("There are", sum(g2m_genes %in% top_genes), "of the", length(g2m_genes), "G2M.Score genes in the top", top_n)

# or as a table
table(row_annotation$gene_set)

df <- as.data.frame(heatmap_matrix) %>%
  tibble::rownames_to_column("genes") %>%
  dplyr::left_join(., tibble::rownames_to_column(row_annotation, "genes"))
```

# Session Info

```{r session-info}
sessioninfo::session_info()
```