---
title: "Dreamlet analysis of scRNAseq"
author: "Melanie Smith"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Load required libraries

```{r}
# if (!require("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }

# BiocManager::install("dreamlet")

# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(SingleCellExperiment)
library(dreamlet)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)

# annotation help
library(homologene)

```

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
data_dir <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes")
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}
```

# Cell cycle gene sets (mouse)

Cell cycle gene lists are re-derived here from `cc.genes.updated.2019` via
human → mouse ortholog conversion. This is the same conversion used in the
upstream pre-processing script.

```{r cc-gene-sets}
human_genes_to_mouse <- function(human_genes) {
  hom_data <- homologene::homologene(human_genes, inTax = 9606, outTax = 10090)
  setNames(hom_data$`10090`, hom_data$`9606`)
}

s_genes   <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$s.genes)
g2m_genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$g2m.genes)
cc_genes  <- c(s_genes, g2m_genes)

cat("S-phase genes retained: ",   length(s_genes),   "\n")
cat("G2M-phase genes retained: ", length(g2m_genes), "\n")
cat("Total CC genes: ",           length(cc_genes),  "\n")
```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(data_dir, "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
# and is a factor
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

# add collapsed cycling phase to the metadata
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = ifelse(
    re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M"),
    "cycling",
    "not_cycling"
  ),
  col.name = "cycling"
)
```

# Trim down the Seurat object metadata -> keep only what we need
- There a are a bunch of columns in the metadata that we don't need at all for this analysis and it makes the sce object and pseudibulked object more complicated so I want to reduce the noise here and just keep the columns we use downstream.

```{r}
# Check what you currently have
names(re_embed_interesting_cells@meta.data)

# Define columns to KEEP (easier than listing what to remove)
cols_to_keep <- c(
  "orig.ident",
  "seurat_clusters",
  "new_celltypes",
  "stage",
  "genotype",
  "Phase",
  "S.Score",
  "G2M.Score",
  "cycling"
)

# Remove everything not in that list
cols_to_remove <- setdiff(names(re_embed_interesting_cells@meta.data), cols_to_keep)
re_embed_interesting_cells@meta.data[, cols_to_remove] <- NULL

# Verify
names(re_embed_interesting_cells@meta.data)
```

# make some umaps
- Let's remind ourselves where the object is at

```{r}
# make some umaps
DimPlot(re_embed_interesting_cells,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = TRUE,
        label.size = 6) +
  NoLegend()

DimPlot(re_embed_interesting_cells,
        reduction = "umap",
        group.by = "cycling",
        label = TRUE,
        label.size = 6) +
  NoLegend()

DimPlot(re_embed_interesting_cells,
        reduction = "umap",
        group.by = "new_celltypes",
        label = TRUE,
        label.size = 6) +
  NoLegend()


# make some summary tables
table(re_embed_interesting_cells@meta.data$seurat_clusters,
      re_embed_interesting_cells@meta.data$new_celltypes)

# Build the table
cc_table <- table(
  re_embed_interesting_cells@meta.data$seurat_clusters,
  re_embed_interesting_cells@meta.data$cycling
)

# Add percent cycling column
cc_table_df <- as.data.frame.matrix(cc_table) %>%
  tibble::rownames_to_column("seurat_clusters") %>%
  dplyr::mutate(
    total       = cycling + not_cycling,
    pct_cycling = round(cycling / total * 100, 1)
  ) %>%
  dplyr::arrange(as.numeric(seurat_clusters))

cc_table_df

```

# Add percent cycling to metadata
-  let's add the percentage of cycling cells to the metadata before we aggregate to pseudobulk. I'm hoping hat if the values are identical then the collapse will keep the metadata column and we can use it in the VariancePartition() function downstream.

```{r}
# Calculate proportion cycling per seurat_cluster
cluster_cycling_pct <- re_embed_interesting_cells@meta.data %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    pct_cycling = mean(cycling == "cycling") * 100,
    .groups = "drop"
  )

# Build a named vector of pct_cycling, named by cell barcode
# This is what AddMetaData expects for safe barcode-aligned assignment
pct_cycling_named <- re_embed_interesting_cells@meta.data %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::left_join(cluster_cycling_pct, by = "seurat_clusters") %>%
  dplyr::select(barcode, pct_cycling) %>%
  tibble::deframe()  # converts two-column data frame to named vector

# Add to Seurat object — AddMetaData aligns by barcode name
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = pct_cycling_named,
  col.name = "pct_cycling_in_cluster"
)

# Verify
re_embed_interesting_cells@meta.data %>%
  dplyr::distinct(seurat_clusters, pct_cycling_in_cluster) %>%
  dplyr::arrange(as.numeric(as.character(seurat_clusters)))

```

# standardise metadata types

```{r}
# Convert categorical variables to factors
re_embed_interesting_cells$orig.ident <-
  factor(re_embed_interesting_cells$orig.ident)

re_embed_interesting_cells$seurat_clusters <-
  factor(re_embed_interesting_cells$seurat_clusters)

re_embed_interesting_cells$stage <-
  factor(re_embed_interesting_cells$stage)

re_embed_interesting_cells$genotype <-
  factor(re_embed_interesting_cells$genotype)

re_embed_interesting_cells$cycling <-
  factor(
    re_embed_interesting_cells$cycling,
    levels = c("not_cycling", "cycling")
  )

# Check
# str(re_embed_interesting_cells[[]][, required_meta])
```

# Chunk 3: calculate proliferation percentage at sample × cluster level

```{r}
# # Extract metadata
# cell_metadata <- re_embed_interesting_cells[[]]
# 
# # Summarise cycling status per sample and Seurat cluster
# pb_metadata <- cell_metadata %>%
#   group_by(
#     orig.ident,
#     seurat_clusters
#   ) %>%
#   summarise(
#     n_cells = n(),
#     n_cycling = sum(cycling == "cycling"),
#     proliferation_percentage = n_cycling / n_cells,
#     stage = dplyr::first(stage),
#     genotype = dplyr::first(genotype),
#     .groups = "drop"
#   )
# 
# # Convert back to factors
# pb_metadata <- pb_metadata %>%
#   mutate(
#     orig.ident = factor(orig.ident),
#     seurat_clusters = factor(seurat_clusters),
#     stage = factor(stage),
#     genotype = factor(genotype)
#   )
# 
# # Inspect
# pb_metadata
# 
# head(pb_metadata)
# summary(pb_metadata$proliferation_percentage)
# table(pb_metadata$seurat_clusters, pb_metadata$orig.ident)
```

# Convert Seurat to SCE object

```{r}
# Convert Seurat object
sce <- Seurat::as.SingleCellExperiment(
  re_embed_interesting_cells,
  assay = "RNA"
)

# Check assays
assayNames(sce)

# Check metadata survived conversion
colnames(colData(sce))

# inspect
assayNames(sce)
colData(sce)
```

# create the pseudobulk object

```{r}
pb <- aggregateToPseudoBulk(
    sce,
    assay = "counts",
    cluster_id = "seurat_clusters",
    sample_id = "orig.ident",
    verbose = FALSE
    )

assayNames(pb)

pb
```

# Voom for pseudobulk

```{r}
# Normalize and apply voom
res.proc <- processAssays(pb, ~1, min_cells = 5, min.count = 5)

# Diagnostics
details(res.proc)
plotVoom(res.proc)
```

```{r}
# Build sample -> pct_cycling lookup from sce colData
cycling_summary <- as.data.frame(colData(sce)) %>%
  dplyr::group_by(orig.ident) %>%
  dplyr::summarise(
    pct_cycling = mean(cycling == "cycling") * 100,
    .groups = "drop"
  ) %>%
  tibble::column_to_rownames("orig.ident")

# rownames of colData(res.proc) are sample IDs directly (e.g. "e11_control")
# no extraction needed — they match cycling_summary rownames directly
stopifnot(all(rownames(colData(res.proc)) %in% rownames(cycling_summary)))

# Add to colData
colData(res.proc)$pct_cycling <- cycling_summary[rownames(colData(res.proc)), "pct_cycling"]

# Verify
colData(res.proc)

vp_cycling <- fitVarPart(res.proc, ~ pct_cycling)
class(vp_cycling)
head(vp_cycling)

# Plot
plotVarPart(sortCols(vp_cycling), label.angle = 60)
# Plot per-gene percent bars for the top 5 genes
# Sort by genotype variance
top_genes_geno <- rownames(vp_cycling)[order(vp_cycling$pct_cycling, decreasing = TRUE)][1:20]
plotPercentBars(vp_cycling[top_genes_geno, ])

# Find s_genes present in the gene column
s_in_vp <- s_genes[s_genes %in% vp_cycling$gene]
cat("S-phase genes in vp_cycling: ", length(s_in_vp), "\n")

# Subset and order by pct_cycling
s_vp <- vp_cycling[vp_cycling$gene %in% s_in_vp, ]
s_vp <- s_vp[order(s_vp$pct_cycling, decreasing = TRUE), ]

plotPercentBars(s_vp)

# e.g. just cluster 0
s_vp <- vp_cycling[vp_cycling$gene %in% s_in_vp & vp_cycling$assay == c("0", "2", "3", "5", "8", "10"), ]
s_vp <- s_vp[order(s_vp$pct_cycling, decreasing = TRUE), ]

plotPercentBars(s_vp)
```

