---
title: "Dreamlet analysis of scRNAseq"
author: "Melanie Smith"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Load required libraries

```{r}
# if (!require("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }

# BiocManager::install("dreamlet")

# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(SingleCellExperiment)
library(dreamlet)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)
```

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
data_dir <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes")
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}
```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(data_dir, "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
# and is a factor
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"
```

# Convert Seurat object to single cell experiment object

```{r}
# The Dreamlet package prefers to work with a single cell experiment object so we can use the Seurat package to convert
sce <- Seurat::as.SingleCellExperiment(re_embed_interesting_cells,
                                       assay = "RNA")

# Calculate mean S.Score and G2M.Score per sample (orig.ident) BEFORE removing the Seurat object
# This is done here to retain access to cell-level metadata
cc_means <- aggregate(
  cbind(S.Score, G2M.Score) ~ orig.ident,
  data = as.data.frame(colData(sce)),
  FUN = mean
)
rownames(cc_means) <- cc_means$orig.ident

# remove the large Seurat object, we don't need it anymore
rm(re_embed_interesting_cells)
# clean up the memory
gc()
```

# Aggregate pseudobulk

```{r}
# The object contains a unique sample identifier in `orig.ident`
# unique(colData(sce)$orig.ident)
# [1] "e11_control" "e11_ko" "e12_control" "e12_ko"

# Create pseudobulk data by specifying cluster_id and sample_id
# Count data for each cell type is then stored in the `assay` field
# assay: entry in assayNames(sce) storing raw counts
# cluster_id: variable in colData(sce) indicating cell clusters
# sample_id: variable in colData(sce) indicating sample id for aggregating cells
pb <- aggregateToPseudoBulk(sce,
  assay = "counts",
  cluster_id = "new_celltypes",
  sample_id = "orig.ident",
  verbose = FALSE
)

# one 'assay' per cell type
assayNames(pb)

# colData(pb) rows are cell-type x sample combinations (e.g. "Mesenchyme.e11_control")
# Extract the sample component from these rownames to key into cc_means
pb_sample_ids <- sub("^[^.]+\\.", "", rownames(colData(pb)))

# Confirm all extracted IDs are present in cc_means before assignment
stopifnot(all(pb_sample_ids %in% rownames(cc_means)))

# Add mean cell cycle scores to pb colData in matching order
colData(pb)$S.Score   <- cc_means[pb_sample_ids, "S.Score"]
colData(pb)$G2M.Score <- cc_means[pb_sample_ids, "G2M.Score"]

# Verify all 4 variables are now present and non-NA
colData(pb)[, c("S.Score", "G2M.Score", "genotype", "stage")]
```

# Voom for pseudobulk

```{r}
# Normalize and apply voom
res.proc <- processAssays(pb, ~1, min_cells = 5, min.count = 5)

# colData(res.proc) rows are also cell-type x sample combinations
# Extract the sample component from these rownames to key into cc_means
rp_sample_ids <- sub("^[^.]+\\.", "", rownames(colData(res.proc)))

# Confirm all extracted IDs are present in cc_means before assignment
stopifnot(all(rp_sample_ids %in% rownames(cc_means)))

# Add mean cell cycle scores to res.proc colData in matching order
colData(res.proc)$S.Score   <- cc_means[rp_sample_ids, "S.Score"]
colData(res.proc)$G2M.Score <- cc_means[rp_sample_ids, "G2M.Score"]

# Verify scores are present and non-NA
colData(res.proc)[, c("S.Score", "G2M.Score", "genotype", "stage")]

# Diagnostics
details(res.proc)
plotVoom(res.proc)
```

# Variance partitioning by cell cycle score
- What fraction of gene-expression variation is due to cell cycle?

```{r}
# Rename cell cycle score columns to avoid issues with "." delimiter in variable names
# Work directly with colData(res.proc) — no need to extract a separate data frame
# as fitVarPart will use colData(res.proc) by default
colData(res.proc)$SScore   <- colData(res.proc)$S.Score
colData(res.proc)$G2MScore <- colData(res.proc)$G2M.Score

# Fit variance partition model for cell cycle scores
vp.form <- ~ SScore + G2MScore
vp.res <- fitVarPart(res.proc, vp.form)

# Plot genome-wide variance fractions
plotVarPart(sortCols(vp.res))

# Sort by SScore variance (or G2MScore — whichever is of primary interest)
top_genes_sscore <- rownames(vp.res)[order(vp.res$SScore, decreasing = TRUE)][1:20]
plotPercentBars(vp.res[top_genes_sscore, ])

plotPercentBars(vp.res[
  rownames(vp.res)[order(vp.res$SScore, decreasing = TRUE)][21:40], ])


# Sort by SScore variance (or G2MScore — whichever is of primary interest)
top_genes_g2m <- rownames(vp.res)[order(vp.res$G2MScore, decreasing = TRUE)][1:20]
plotPercentBars(vp.res[top_genes_g2m, ])

```

# Variance partitioning by genotype
- What fraction of gene-expression variation is due to genotype?

```{r}
# Fit variance partition model for genotype
vp.lst <- fitVarPart(res.proc, ~genotype)

# Plot genome-wide variance fractions
plotVarPart(vp.lst, label.angle = 60)

# Plot per-gene percent bars for the top 5 genes
# Sort by genotype variance
top_genes_geno <- rownames(vp.lst)[order(vp.lst$genotype, decreasing = TRUE)][1:20]
plotPercentBars(vp.lst[top_genes_geno, ])
```
