---
title: "09-remove_cellcycle_genes_reembed"
author: "Melanie Smith"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Purpose: re-embed the dataset after removing the consensus cell-cycle gene
set identified in `08-dreamlet_analysis_cycling_orig.ident_pseudobulk.Rmd`,
and compare UMAPs before vs after removal to assess whether cell-cycle-driven
structure was contributing to the embedding independent of the biological
cell type / developmental structure of interest.

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE)
set.seed(42) # set seed for reproducibility
```

# Load required libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(SingleCellExperiment)

# additional plotting functions
library(patchwork)
library(pheatmap)

# additional table functions
library(kableExtra)
```

# Set project directories

```{r}
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
data_dir     <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes")
cc_dir       <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")
out_dir      <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

# Load required Seurat object

- Load the same post-QC, post-filtering Seurat object used in `08-...Rmd`.

```{r}
re_embed_interesting_cells <- readRDS(file.path(data_dir, "final_interesting_cells_0.6.rds"))
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

Idents(re_embed_interesting_cells) <- "seurat_clusters"
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

DefaultAssay(re_embed_interesting_cells) <- "RNA"

# recreate the cycling / orig.ident.cycling columns exactly as in 08-...Rmd
re_embed_interesting_cells <- AddMetaData(
  re_embed_interesting_cells,
  metadata = ifelse(
    re_embed_interesting_cells@meta.data$Phase %in% c("S", "G2M"),
    "cycling",
    "not_cycling"
  ),
  col.name = "cycling"
)

re_embed_interesting_cells@meta.data$orig.ident    <- factor(re_embed_interesting_cells@meta.data$orig.ident)
re_embed_interesting_cells@meta.data$seurat_clusters <- factor(re_embed_interesting_cells@meta.data$seurat_clusters)
re_embed_interesting_cells@meta.data$new_celltypes <- factor(re_embed_interesting_cells@meta.data$new_celltypes)
re_embed_interesting_cells@meta.data$cycling <- factor(
  re_embed_interesting_cells@meta.data$cycling,
  levels = c("not_cycling", "cycling")
)
```

```{r}
DimPlot(re_embed_interesting_cells,
        reduction = "pca",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = PCA - Phase")
      )
```


# Preserve the original PCA and UMAP structures
- I'd like to preserve these here so when I split the object and re-embed after removing the cell cycle genes I can still plot the old structure and compare

```{r}
# ----------------------------------------------------------
# Preserve original PCA and UMAP before splitting samples
# ----------------------------------------------------------

if ("pca" %in% Reductions(re_embed_interesting_cells)) {
  re_embed_interesting_cells[["pca_original"]] <-
    re_embed_interesting_cells[["pca"]]
  
  Key(re_embed_interesting_cells[["pca_original"]]) <- "PCAOrig_"
}

if ("umap" %in% Reductions(re_embed_interesting_cells)) {
  re_embed_interesting_cells[["umap_original"]] <-
    re_embed_interesting_cells[["umap"]]
  
  Key(re_embed_interesting_cells[["umap_original"]]) <- "UMAPOrig_"
}
```

```{r}
# let's double check the PCAs, they should be identical
p1 <- DimPlot(re_embed_interesting_cells,
        reduction = "pca",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca - Phase")
      )

p2 <- DimPlot(re_embed_interesting_cells,
        reduction = "pca_original",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca_original - Phase")
      )
p1 + p2
```

# Load the consensus cell-cycle gene set

- This is the `full_consensus` gene vector built in `08-...Rmd`
  (`lost_genes` + `consensus_genes_95_celltype`).

```{r load-consensus-genes}
full_consensus <- read_csv(file.path(cc_dir, "expr_mat_celltype_phase.csv"))

length(full_consensus)

# how many of these genes are actually present in the object
sum(full_consensus$gene %in% rownames(re_embed_interesting_cells)) # [1] 181
```

# 1. Baseline UMAPs (before gene removal)

```{r baseline-umaps, fig.width=14, fig.height=10}
p1 <- DimPlot(re_embed_interesting_cells,
              reduction = "umap",
              group.by = "seurat_clusters",
              label = TRUE,
              label.size = 5) +
  NoLegend() +
  labs(title = "Seurat clusters")

p2 <- DimPlot(re_embed_interesting_cells,
              reduction = "umap",
              group.by = "Phase") +
  labs(title = "Cell cycle phase")

p3 <- DimPlot(re_embed_interesting_cells,
              reduction = "umap",
              group.by = "cycling") +
  labs(title = "Cycling status")

p4 <- DimPlot(re_embed_interesting_cells,
              reduction = "umap",
              group.by = "new_celltypes") +
  labs(title = "Cell type")

baseline_umaps <- (p1 | p2) / (p3 | p4) +
  plot_annotation(title = "Baseline UMAPs (all genes, before cell-cycle gene removal)")

baseline_umaps
```

```{r save-baseline-umaps}
ggsave(file.path(out_dir, "baseline_umaps.pdf"), baseline_umaps, width = 14, height = 10)
```

# 2. Remove consensus cell-cycle genes and re-embed

## 2.1 Subset the object to exclude the consensus gene set

```{r subset-genes}
genes_to_keep <- setdiff(rownames(re_embed_interesting_cells), full_consensus$gene)

length(rownames(re_embed_interesting_cells))
length(genes_to_keep)

re_embed_no_cc <- subset(re_embed_interesting_cells, features = genes_to_keep)

# sanity check - none of the excluded genes should remain
stopifnot(!any(full_consensus %in% rownames(re_embed_no_cc)))
```

```{r}
p3 <- DimPlot(re_embed_interesting_cells,
        reduction = "pca",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca - cc genes removed - Phase")
      )

p4 <- DimPlot(re_embed_interesting_cells,
        reduction = "pca_original",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca_original - cc genes removed - Phase")
      )
p3 + p4
```

## 2.2 Standard re-processing on the reduced gene set

Note: `NormalizeData()` is not re-run here, since the log-normalization
factors were computed from total counts across all genes in the original
object and removing a small gene subset shouldn't require renormalizing
from raw counts. If you'd prefer to renormalize from scratch (e.g. if the
excluded gene set is large enough to meaningfully shift per-cell size
factors), swap in `NormalizeData()` as the first step below.

```{r reprocess}
re_embed_no_cc <- FindVariableFeatures(re_embed_no_cc,
                                       selection.method = "vst",
                                       nfeatures = 2000)

re_embed_no_cc <- ScaleData(re_embed_no_cc,
                            features = rownames(re_embed_no_cc))

re_embed_no_cc <- RunPCA(re_embed_no_cc,
                         reduction.name = "pca_no_cc",
                         reduction.key = "PCNoCC_",
                         features = VariableFeatures(re_embed_no_cc),
                         npcs = 30)

ElbowPlot(re_embed_no_cc, ndims = 30)
```

```{r neighbours-clusters-umap}
# adjust dims if the elbow plot above suggests a different cutoff
n_pcs <- 20

re_embed_no_cc <- FindNeighbors(re_embed_no_cc,
                                reduction = "pca_no_cc",
                                dims = 1:n_pcs)
re_embed_no_cc <- FindClusters(re_embed_no_cc, resolution = 0.6)
re_embed_no_cc <- RunUMAP(re_embed_no_cc,
                          reduction = "pca_no_cc",
                          reduction.name = "umap_no_cc",
                          reduction.key = "UMAPNoCC_",
                          dims = 1:n_pcs)

# preserve the new clustering separately from the original seurat_clusters
re_embed_no_cc$seurat_clusters_no_cc <- Idents(re_embed_no_cc)
```

```{r}
p5 <- DimPlot(re_embed_no_cc,
        reduction = "pca",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca - cc genes removed - re-embed etc - Phase")
      )
p6 <- DimPlot(re_embed_no_cc,
        reduction = "pca_original",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca_original - cc genes removed - re-embed etc - Phase")
      )
p7 <- DimPlot(re_embed_no_cc,
        reduction = "pca_no_cc",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = pca_no_cc - cc genes removed - re-embed etc - Phase")
      )
p5 + p6 + p7
```

```{r save-reduced-object}
saveRDS(re_embed_no_cc, file.path(out_dir, "re_embed_no_cellcycle_genes.rds"))
```

---

# 3. Post-removal UMAPs

```{r post-removal-umaps, fig.width=14, fig.height=10}
p1_new <- DimPlot(re_embed_no_cc,
                   reduction = "umap",
                   group.by = "seurat_clusters_no_cc",
                   label = TRUE,
                   label.size = 5) +
  NoLegend() +
  labs(title = "Seurat clusters (re-clustered, no CC genes)")

p2_new <- DimPlot(re_embed_no_cc,
                   reduction = "umap",
                   group.by = "Phase") +
  labs(title = "Cell cycle phase")

p3_new <- DimPlot(re_embed_no_cc,
                   reduction = "umap",
                   group.by = "cycling") +
  labs(title = "Cycling status")

p4_new <- DimPlot(re_embed_no_cc,
                   reduction = "umap",
                   group.by = "new_celltypes") +
  labs(title = "Cell type (original annotation)")

post_removal_umaps <- (p1_new | p2_new) / (p3_new | p4_new) +
  plot_annotation(title = "Post-removal UMAPs (consensus cell-cycle genes excluded)")

post_removal_umaps

(p1 | p2) / (p1_new | p2_new) +
  plot_annotation(title = "Pre-(top) and Post-removal (bottom) UMAPs (consensus cell-cycle genes excluded)")

(p3 | p4) / (p3_new | p4_new) +
  plot_annotation(title = "Pre-(top) and Post-removal (bottom) UMAPs (consensus cell-cycle genes excluded)")
```

```{r save-post-removal-umaps}
ggsave(file.path(out_dir, "post_removal_umaps.pdf"), post_removal_umaps, width = 14, height = 10)
```

---

# 4. Side-by-side comparison

```{r comparison, fig.width=14, fig.height=14}
comparison_phase <- (p2 + labs(title = "Phase: before")) |
  (p2_new + labs(title = "Phase: after"))

comparison_cycling <- (p3 + labs(title = "Cycling: before")) |
  (p3_new + labs(title = "Cycling: after"))

comparison_celltype <- (p4 + labs(title = "Cell type: before")) |
  (p4_new + labs(title = "Cell type: after"))

full_comparison <- comparison_phase / comparison_cycling / comparison_celltype +
  plot_annotation(title = "Before vs after consensus cell-cycle gene removal")

full_comparison
```

```{r save-comparison}
ggsave(file.path(out_dir, "before_after_comparison.pdf"), full_comparison, width = 14, height = 16)
```

## What to look for

- **Phase / cycling panels**: if cell-cycle-driven structure was
  contributing meaningfully to the original embedding, the "before" panels
  should show cycling cells forming a distinguishable arm/cluster
  independent of cell type, while the "after" panels should show cycling
  status distributed more evenly *within* each cell-type region rather than
  forming its own separate structure.
- **Cell type panel**: this is the sanity check — cell-type separation
  should be preserved (ideally as clean or cleaner) after removal. If cell
  types that were previously well-separated become mixed together after
  removing the consensus genes, that's a signal some of the excluded genes
  may have been carrying real developmental/cell-type signal alongside
  cell-cycle signal, and the exclusion threshold (Section 10 of the 08
  script) may be worth revisiting.
- **Cluster count/structure**: a reduced number of clusters, or clusters
  that no longer split by cycling status, is the expected and desired
  outcome if cell-cycle genes were previously driving spurious
  subclustering within a single biological cell type.

# ============================================================
# Does residual "cell-cycle" signal track cell type / cell state,
# or does it look like leftover proliferation noise?
# ============================================================

```{r}
# original UMAP - split by cell type
DimPlot(re_embed_interesting_cells,
        reduction = "umap",
        group.by = "Phase",
        split.by = "new_celltypes",
        label = TRUE,
        label.size = 5) +
  NoLegend() +
  labs(title = "Cell cycle phase - pre cell cycle gene removal (split by cell type)")

DimPlot(re_embed_interesting_cells,
        reduction = "pca",
        dims = c(3, 4),
        group.by = "Phase",
        split.by = "new_celltypes",
        label = TRUE,
        label.size = 5) +
  NoLegend() +
  labs(title = "Cell cycle phase - pre cell cycle gene removal (split by cell type)")

# subset UMAP - split by cell type
DimPlot(re_embed_no_cc,
        reduction = "umap",
        group.by = "Phase",
        split.by = "new_celltypes",
        label = TRUE,
        label.size = 5) +
  NoLegend() +
  labs(title = "Cell cycle phase - re-clustered, no CC genes (split by cell type)")

DimPlot(re_embed_no_cc,
        reduction = "pca",
        dims = c(3, 4),
        group.by = "Phase",
        split.by = "new_celltypes",
        label = TRUE,
        label.size = 5) +
  NoLegend() +
  labs(title = "Cell cycle phase - re-clustered, no CC genes (split by cell type)")


celltypes <- levels(re_embed_no_cc$new_celltypes)

phase_umaps_by_celltype_pre <- list()

for (ct in celltypes) {

  cells_ct <- WhichCells(re_embed_interesting_cells, expression = new_celltypes == ct)

  p <- DimPlot(re_embed_interesting_cells,
               reduction = "umap",
               group.by = "Phase",
               cells = cells_ct,
               label = FALSE) +
    labs(title = ct)

  phase_umaps_by_celltype_pre[[ct]] <- p
}

wrap_plots(phase_umaps_by_celltype_pre, ncol = 3)  +
  plot_annotation(title = "Pre subset cell cycle genes")



phase_umaps_by_celltype <- list()

for (ct in celltypes) {

  cells_ct <- WhichCells(re_embed_no_cc, expression = new_celltypes == ct)

  p <- DimPlot(re_embed_no_cc,
               reduction = "umap",
               group.by = "Phase",
               cells = cells_ct,
               label = FALSE) +
    labs(title = ct)

  phase_umaps_by_celltype[[ct]] <- p
}

wrap_plots(phase_umaps_by_celltype, ncol = 3)  +
  plot_annotation(title = "Post subset cell cycle genes")

```

```{r}

celltypes <- levels(re_embed_interesting_cells$new_celltypes)

phase_umaps_by_celltype <- list()

for (ct in celltypes) {

  cells_ct <- WhichCells(re_embed_interesting_cells, expression = new_celltypes == ct)

  p <- DimPlot(re_embed_interesting_cells,
               reduction = "umap",
               group.by = "Phase",
               cells = cells_ct,
               label = FALSE) +
    labs(title = ct)

  phase_umaps_by_celltype[[ct]] <- p
}

wrap_plots(phase_umaps_by_celltype, ncol = 3)
```

# Save the subset and re-embedded Seurat object

```{r}
write_rds(re_embed_no_cc,
          file = file.path(out_dir, "re_embed_no_cc.rds"))
```


# Session info

```{r session-info}
sessionInfo()
```
