---
title: "regress_cell_cycle"
author: "Melanie Smith"
date: "2026-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(loupeR)

# additional plotting functions
library(patchwork)
library(pheatmap)

```

The aim of this script is to mitigate the effects of cell cycle (CC) heterogeneity in our scRNA-seq data. To accomplish this we first write in the complete Seurat object containing all 4 samples, followed by using the Seurat function `ScaleData()` with `vars.to.regress = c("CC.Difference")` to regress together the "S.Score" and the "G2M.Score". This milder alternative (compared to fully regressing out all CC information) is regressing only the difference (S.Score - G2M.Score). This keeps the broad “cycling vs. quiescent” signal while removing phase-specific noise among proliferating cells.  

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

```

## Set up the cell-type colours

```{r}
celltype_cols <- c("Myocardium" = "#E41A1C",
                   "VSMC" = "#377EB8", 
                   "Epicardium" = "#4DAF4A",
                   "Mesenchyme" = "#984EA3", 
                   "Endocardium" = "#FF7F00",
                   "Other/Unknown" = "grey70")
```

# Quick sanity check

```{r}
print(re_embed_interesting_cells)
table(Idents(re_embed_interesting_cells))
stopifnot(!anyDuplicated(colnames(re_embed_interesting_cells)))
```

# Sanity plots
## cell cycle genes -> ridge plots

```{r}
# set active identity to cell cycle
Idents(re_embed_interesting_cells) <- "Phase"

# Visualize the distribution of cell cycle markers across
RidgePlot(re_embed_interesting_cells,
          features = c("Pcna", "Top2a", "Mcm6", "Mki67"),
          ncol = 2)
```

# Heatmap of highly variable genes (nFeaures=200, downsample to 5000 cells)

```{r}
# 1. Get the top highly variable genes (usually 50–100 is readable)
top_hvg <- VariableFeatures(re_embed_interesting_cells)[1:200]   # or 50, 200 — adjust as needed

# Downsample cells if you have > ~5,000–10,000 cells total
#    This makes the heatmap much faster and less crowded
set.seed(123)  # for reproducibility
cells_to_plot <- subset(re_embed_interesting_cells, downsample = 5000)

# 3. Plot the heatmap — group by your current clusters (or "Phase", "orig.ident", etc.)
DoHeatmap(cells_to_plot,
          features = top_hvg,
          group.by = "seurat_clusters",     # change to "Phase" or "orig.ident" if you prefer
          size = 3,                         # font size for gene labels
          angle = 90) +                     # rotate column labels if needed
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +   # nice diverging scale
  NoLegend() +
  plot_annotation(
    title = "Heatmap of Top 200 Highly Variable Genes",
    subtitle = "Scaled expression (scale.data)",
    caption = "Pre cell-cycle regression | Grouped by current clusters"
  )

# 3. Plot the heatmap — group by your current clusters (or "Phase", "orig.ident", etc.)
DoHeatmap(cells_to_plot,
          features = top_hvg,
          group.by = "orig.ident",     # change to "Phase" or "orig.ident" if you prefer
          size = 3,                         # font size for gene labels
          angle = 90) +                     # rotate column labels if needed
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +   # nice diverging scale
  NoLegend() +
  plot_annotation(
    title = "Heatmap of Top 200 Highly Variable Genes",
    subtitle = "Scaled expression (scale.data)",
    caption = "Pre cell-cycle regression | Grouped by sample"
  )

```


## cell cycle classification on PCA

```{r}
# Get loadings for PC1:4
pc_loadings <- Loadings(re_embed_interesting_cells)

# Create a nice sorted data frame
top_loadings_pc1 <- as.data.frame(pc_loadings) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::select(gene, loading = PC_1) %>%
  mutate(abs_loading = abs(loading)) %>%
  mutate(rank_positive = rank(-loading),      # lower number = more positive
         rank_negative = rank(loading)) %>%   # lower number = more negative
  filter(rank_positive <= 20 | rank_negative <= 20) %>%
  arrange(desc(loading))

top_loadings_pc2 <- as.data.frame(pc_loadings) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::select(gene, loading = PC_2) %>%
  mutate(abs_loading = abs(loading)) %>%
  mutate(rank_positive = rank(-loading),      # lower number = more positive
         rank_negative = rank(loading)) %>%   # lower number = more negative
  filter(rank_positive <= 20 | rank_negative <= 20) %>%
  arrange(desc(loading))

top_loadings_pc3 <- as.data.frame(pc_loadings) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::select(gene, loading = PC_3) %>%
  mutate(abs_loading = abs(loading)) %>%
  mutate(rank_positive = rank(-loading),      # lower number = more positive
         rank_negative = rank(loading)) %>%   # lower number = more negative
  filter(rank_positive <= 20 | rank_negative <= 20) %>%
  arrange(desc(loading))

top_loadings_pc4 <- as.data.frame(pc_loadings) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::select(gene, loading = PC_4) %>%
  mutate(abs_loading = abs(loading)) %>%
  mutate(rank_positive = rank(-loading),      # lower number = more positive
         rank_negative = rank(loading)) %>%   # lower number = more negative
  filter(rank_positive <= 20 | rank_negative <= 20) %>%
  arrange(desc(loading))
```

## Visualisation

```{r}
# Function to create and save heatmap for one PC
create_pc_heatmap <- function(pc_number, object = re_embed_interesting_cells) {
  
  # Dynamically get the loadings table (assumes you have top_loadings_pc1, pc2, etc.)
  loadings_table <- get(paste0("top_loadings_pc", pc_number))
  
  # Extract genes
  genes_of_interest <- loadings_table$gene
  genes_of_interest <- genes_of_interest[genes_of_interest %in% rownames(object)]
  
  if (length(genes_of_interest) == 0) {
    message("No genes found for PC", pc_number)
    return(NULL)
  }
  
  # Get scaled data
  scaled_data <- GetAssayData(object, 
                              assay = "RNA",
                              slot = "scale.data")[genes_of_interest, ]
  
  # Cell annotation
  cell_annotation <- data.frame(
    CellType = object$new_celltypes,
    Phase    = object$Phase,
    row.names = colnames(object)
  )
  
  # Plot title
  plot_title <- paste0("Expression of Top Loading Genes - PC", pc_number)
  
  # Generate pheatmap
  p <- pheatmap(scaled_data,
                cluster_rows = TRUE,
                cluster_cols = TRUE,
                treeheight_row = 0,
                treeheight_col = 0,
                show_colnames = FALSE,
                annotation_col = cell_annotation,
                annotation_colors = list(CellType = celltype_cols),
                scale = "none",
                color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
                main = plot_title,
                fontsize_row = 10,
                fontsize = 12,
                silent = TRUE)   # prevents printing in loop
  
  # Save as PDF (recommended for high quality)
  pdf_filename <- file.path(out_dir, paste0("08-PC", pc_number, "_top_loadings_heatmap.pdf"))
  ggsave(filename = pdf_filename, 
         plot = p$gtable, 
         width = 12, 
         height = 10, 
         dpi = 300)
  
  # Optional: also save as PNG
  png_filename <- file.path(out_dir, paste0("08-PC", pc_number, "_top_loadings_heatmap.png"))
  ggsave(filename = png_filename, 
         plot = p$gtable, 
         width = 12, 
         height = 10, 
         dpi = 300)
  
  message("✓ Saved PC", pc_number, " heatmap: ", pdf_filename)
  
  return(p)
}

# Run the loop for PC1 to PC4
for (pc in 1:4) {
  create_pc_heatmap(pc)
}
```


```{r}

# Variance explained by top PCs
ElbowPlot(re_embed_interesting_cells, ndims = 50)

# Heatmap of top PCA genes
pca_heatmap_pre_cc_regress <- DimHeatmap(re_embed_interesting_cells,
                                          dims = c(1:4),
                                          ncol = 2,
                                         fast = FALSE) +
  plot_annotation(
    title = "Heatmap top genes - PC 1:4 - all samples - pre cc regress",
    caption = "Final dataset"
    )
ggsave(pca_heatmap_pre_cc_regress,
        filename = file.path(out_dir, "08-pca_heatmap_pre_cc_regress.png"),
        width = 12,
        height = 12)

# Running a PCA on cell cycle genes reveals, unsurprisingly, that cells separate entirely by
# phase (dims 1&2)
pca_1_2_pre_cc_regress <- DimPlot(re_embed_interesting_cells,
                              dims = c(1,2),
                              group.by = "Phase",
                              split.by = "orig.ident",
                              combine = TRUE,
                              reduction = "pca",    
                              label = TRUE,
                              repel = TRUE,
                              label.size = 5) +
  plot_annotation(
    title = "PCA 1&2 - all samples - pre cc regress",
    caption = "Final dataset"
    )
ggsave(pca_1_2_pre_cc_regress,
        filename = file.path(out_dir, "08-pca_1_2_pre_cc_regress.png"),
        width = 12,
        height = 8)

# phase (dims 3&4)
pca_3_4_pre_cc_regress <- DimPlot(re_embed_interesting_cells,
                                  dims = c(3,4),
                                  group.by = "Phase",
                                  split.by = "orig.ident",
                                  reduction = "pca",
                                  label = TRUE,
                                  repel = TRUE,
                                  label.size = 5) +
  plot_annotation(
    title = "PCA 3&4 - all samples - pre cc regress",
    caption = "Final dataset"
    )
ggsave(pca_3_4_pre_cc_regress,
        filename = file.path(out_dir, "08-pca_3_4_pre_cc_regress.png"),
        width = 12,
        height = 8)

# Cell types
pca_1_2_pre_cc_regress_celltype <- DimPlot(re_embed_interesting_cells,
                                           dims = c(1,2),
                                           group.by = "new_celltypes",
                                           split.by = "orig.ident",
                                           reduction = "pca",
                                           label = TRUE,
                                           repel = TRUE,
                                           label.size = 5) +
  plot_annotation(
    title = "PCA 1&2 - all samples - pre cc regress",
    caption = "Final dataset"
    ) +
  scale_color_manual(values = celltype_cols)
ggsave(pca_1_2_pre_cc_regress_celltype,
        filename = file.path(out_dir, "08-pca_1_2_pre_cc_regress_celltype.png"),
        width = 12,
        height = 8)
```

# UMAP - colour thing of interest

```{r}
umap_1_2_pre_cc_regress <- DimPlot(re_embed_interesting_cells,
        group.by = "Phase",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "UMAP 1&2 - all samples - pre cc regress",
    caption = "Final dataset"
    )
ggsave(umap_1_2_pre_cc_regress,
        filename = file.path(out_dir, "08-umap_1_2_pre_cc_regress.png"),
        width = 12,
        height = 8)

# feature plot of the cell cycle scores
umap_s_g2m_pre_cc_regress <- FeaturePlot(re_embed_interesting_cells,
            features = c("S.Score", "G2M.Score"), 
            reduction = "umap") &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Score") &
  plot_annotation(
    title = "UMAP 1&2 - all samples - pre cc regress",
    caption = "Final dataset"
    )
ggsave(umap_s_g2m_pre_cc_regress,
        filename = file.path(out_dir, "08-umap_s_g2m_pre_cc_regress.png"),
        width = 12,
        height = 8)

DimPlot(re_embed_interesting_cells,
        group.by = "seurat_clusters",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )
DimPlot(re_embed_interesting_cells,
        group.by = "seurat_clusters",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )

DimPlot(re_embed_interesting_cells,
        group.by = "new_celltypes",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Types",
    caption = "Final dataset"
  )
```

# Cell cycle gene enrichment mudule gene sets
- The cell cycle genes aren't stored in the Seurat object according to the original workflow so I will reproduce them here for use in visualisation. Going forward it might be a good idea to store them in `object@misc$s.features <- s.genes` and `object@misc$g2m.features <- g2m.genes` so we have a reporoducible record should the list from Seurat change.  

```{r}
# this is the function Nick wrote in the initial pre-processing
# Seurat comes with a set of cell cycle genes but these are human genes derived from the Tirosh et al. 2015 paper (the melanoma single-cell study). We need to convert them to mouse gene symbols before we can use them.
human_genes_to_mouse <- function(human_genes){
  hom_data <- homologene::homologene(human_genes, inTax = 9606, outTax = 10090)
  result <- setNames(hom_data$`10090`, hom_data$`9606`)
  return(result)
}

# convert S.Score genes from human to mouse, starting with the 2019 updated gene list
s.genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$s.genes)
# convert G2M scores from human to mouse, starting with the 2019 updated gene list
g2m.genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$g2m.genes)
```

# Distribution of cell cycle scores

```{r}
# Violin plots for S.Score and G2M.Score split by orig.ident
VlnPlot(re_embed_interesting_cells,
        features = c("S.Score", "G2M.Score"),
        group.by = "orig.ident",        # one violin per sample
        split.by = "Phase",             # color by cell cycle phase (very useful!)
        pt.size = 1,                    # remove individual points for cleaner look
        ncol = 1,                       # stack them vertically
        cols = c("G1" = "#1f77b4", "S" = "#ff7f0e", "G2M" = "#2ca02c")) +  # nice colors
  plot_annotation(
    title = "Distribution of Cell Cycle Scores by Sample",
    subtitle = "S.Score and G2M.Score across orig.ident",
    caption = "Pre cell-cycle regression"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# stacked barchart of cycling cells per cell type

```{r}

```


# Store existing data before over-writing
```{r}
# 1. Store a copy of the current reductions
re_embed_interesting_cells[["umap_original"]] <- re_embed_interesting_cells[["umap"]]
re_embed_interesting_cells[["pca_original"]]   <- re_embed_interesting_cells[["pca"]]

# 2. Store current clusters and Phase in case I want to access later
re_embed_interesting_cells$seurat_clusters_original <- re_embed_interesting_cells$seurat_clusters
re_embed_interesting_cells$Phase_original <- re_embed_interesting_cells$Phase
```

# Regress cell cycle difference

```{r}
# Create the difference score (milder regression)
re_embed_interesting_cells$CC.Difference <- re_embed_interesting_cells$S.Score - re_embed_interesting_cells$G2M.Score

# Scale with regression
DefaultAssay(re_embed_interesting_cells) <- "RNA"  # make sure RNA is the default assay

re_embed_interesting_cells <- ScaleData(re_embed_interesting_cells,
                                        vars.to.regress = c("CC.Difference"),  # onlu regressing for cell cycle difference
                                        features = VariableFeatures(re_embed_interesting_cells))

# Re-run downstream (use same dims/resolution as before, or tune slightly)
re_embed_interesting_cells <- RunPCA(re_embed_interesting_cells, npcs = 50)
re_embed_interesting_cells <- FindNeighbors(re_embed_interesting_cells, dims = 1:30)
re_embed_interesting_cells <- FindClusters(re_embed_interesting_cells, resolution = 0.6)  # your previous choice
re_embed_interesting_cells <- RunUMAP(re_embed_interesting_cells, dims = 1:30)
```

# UMAP - colour thing of interest

```{r}
DimPlot(re_embed_interesting_cells,
        group.by = "Phase",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )

# feature plot of the cell cycle scores
FeaturePlot(re_embed_interesting_cells,
            features = c("S.Score", "G2M.Score"), 
            reduction = "umap") &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Score")

DimPlot(re_embed_interesting_cells,
        group.by = "seurat_clusters",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )
DimPlot(re_embed_interesting_cells,
        group.by = "seurat_clusters",
        split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )

DimPlot(re_embed_interesting_cells,
        group.by = "new_celltypes",
        # split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Types",
    caption = "Final dataset"
  )
```



