---
title: "Test cell type proportions"
author: "Melanie Smith"
date: "2026-04-17"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(loupeR)

# gene set enrichment
library(fgsea)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(UpSetR)

```

The aim of this script is to mitigate the effects of cell cycle (CC) heterogeneity in our scRNA-seq data. To accomplish this we first write in the complete Seurat object containing all 4 samples, followed by using the Seurat function `ScaleData()` with `vars.to.regress = c("CC.Difference")` to regress together the "S.Score" and the "G2M.Score". This milder alternative (compared to fully regressing out all CC information) is regressing only the difference (S.Score - G2M.Score). This keeps the broad “cycling vs. quiescent” signal while removing phase-specific noise among proliferating cells.  

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

```

# create the slim object
- I've found this analyses difficult to get through without accumulating enough RAM to crash RStudio.  
- Here I will create a slim version of my Seurat object so that it and all subsequent objects are taking up less space.  

```{r}
# Slim down while keeping exactly what FindMarkers needs
slim_re_embed_interesting_cells <- DietSeurat(
  object = re_embed_interesting_cells,
  
  # Keep only the normalized data layer (this is what FindMarkers uses)
  layers = "data",                    # ← most important line
  
  # Keep only the main assay
  assays = "RNA",                     # or "SCT" if that's your main one
  
  # Keep only the reductions you actually use
  dimreducs = c("pca", "umap"),
  
  # Drop everything else
  graphs = FALSE
)

# remove the large object
rm(re_embed_interesting_cells)
# clean up
gc()
```


## Set up the cell-type and phase colours for consistenc
- setting the colours here to make sure they are consistent across all scripts.  

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",  
                   "Epicardium"    =  "#FF7F00",
                   "Mesenchyme"    =  "#4DAF4A", 
                   "Endocardium"   = "#377EB8",
                   "Other/Unknown" = "grey70")

phase_cols <- c(
  "G1" = "#1f77b4",
  "G2M" = "#2ca02c",
  "S" = "#ff7f0e"
  )
```

# Cell cycle gene enrichment mudule gene sets
- The cell cycle genes aren't stored in the Seurat object according to the original workflow so I will reproduce them here for use in visualisation. Going forward it might be a good idea to store them in `object@misc$s.features <- s.genes` and `object@misc$g2m.features <- g2m.genes` so we have a reporoducible record should the list from Seurat change.  

```{r}
# this is the function Nick wrote in the initial pre-processing
# Seurat comes with a set of cell cycle genes but these are human genes derived from the Tirosh et al. 2015 paper (the melanoma single-cell study). We need to convert them to mouse gene symbols before we can use them.
human_genes_to_mouse <- function(human_genes){
  hom_data <- homologene::homologene(human_genes, inTax = 9606, outTax = 10090)
  result <- setNames(hom_data$`10090`, hom_data$`9606`)
  return(result)
}

# convert S.Score genes from human to mouse, starting with the 2019 updated gene list
s.genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$s.genes)
# convert G2M scores from human to mouse, starting with the 2019 updated gene list
g2m.genes <- human_genes_to_mouse(Seurat::cc.genes.updated.2019$g2m.genes)
```

# Quick sanity check

```{r}
print(slim_re_embed_interesting_cells)
# which ident is currently active
table(Idents(slim_re_embed_interesting_cells))
stopifnot(!anyDuplicated(colnames(slim_re_embed_interesting_cells)))
```

# Collaps cell cycle phase into cycling vs not-cycling

```{r}
# set the cell cycle phase as a factor to ensure plots are always in the same order
slim_re_embed_interesting_cells$Phase <- factor(
  slim_re_embed_interesting_cells$Phase,
  levels = c("G1", "G2M", "S")  # match the order in phase_cols exactly
)

## collapse the S and G2M phase allocations
slim_re_embed_interesting_cells[[]] <- slim_re_embed_interesting_cells[[]] %>%
  dplyr::mutate(
    cycling = dplyr::if_else(
      Phase %in% c("S", "G2M"),
      "cycling",
      "not_cycling"
    )
  )

# lets make these a factor so they always plot in the same order
slim_re_embed_interesting_cells$cycling <- factor(
  slim_re_embed_interesting_cells$cycling,
  levels = c("not_cycling", "cycling")
)

# quick sanity check
table(slim_re_embed_interesting_cells$Phase,
      slim_re_embed_interesting_cells$cycling)
  #     cycling not_cycling
  # G1        0       12436
  # G2M    5796           0
  # S      6095           0
```

# plot collapse cell cycle phase

```{r}

DimPlot(slim_re_embed_interesting_cells,
        group.by = "cycling",
        # split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )

```

# Test cell cycle proportions
## Combined cell types
- pool time points or genotypes to create pseudo biological replicates and test differences in cell type proportions.  
- important caveat here -> we don't "really" have biological replicates so we need to always keep in mind the complexity introduces by timepoint and genotype.  

```{r}
# Q1. Are there differences in the proportion of cells in each cycle phase (G1, S, and G2M)) between the different genotypes?
propeller_phase_genotype_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$Phase,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$genotype,       # condition of interest
  transform = "asin")   

# Q2. Are there differences in the proportion of cycling cells (cycling or not-cycling) between the different genotypes?
propeller_cycling_genotype_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$cycling,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$genotype,       # condition of interest
  transform = "asin")  

# Q3. Does the developmental stage affect the proportion cell cycle phase (G1, S, and G2M) between timepoints?
propeller_phase_stage_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$Phase,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$stage,       # condition of interest
  transform = "asin")   

# Q4. Does the developmental stage affect the proportion of cycling cells (cycling or not-cycling) between timepoints?
propeller_cycling_stage_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$cycling,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$stage,       # condition of intrest
  transform = "asin")

# speckle style bar plots (per cycling/not-cycling)
props_cycling <- speckle::getTransformedProps(
  slim_re_embed_interesting_cells$cycling, slim_re_embed_interesting_cells$orig.ident,
  transform="asin")
barplot(props_cycling$Proportions,
        col = c("orange","purple"),
        legend=TRUE, 
        ylab="Proportions")

# speckle style bar plots (per cell cycle phase)
props_phase <- speckle::getTransformedProps(
  slim_re_embed_interesting_cells$Phase, slim_re_embed_interesting_cells$orig.ident,
  transform="asin")
barplot(props_phase$Proportions,
        col = phase_cols,
        legend=TRUE, 
        ylab="Proportions")

```

## Individual cell types
- Subset each of the cell types and ask the proportion question again for each.  
- The hope is that this retains cell type specific behaviour that might be masked at the higher level.  
- Keep in mind that we still can't interpret the p-values because we don't have true biological replicates. 

### Quick visualisation
- Stat half-eye plots showing the distribution of S.Score and G2M.Score for each of the cell types.  
- Keep in mind each cell type contains different time points and different genotypes.  

```{r}
# stat halfeye plots of S.Score for the full data set, facet wrap by cell type
slim_re_embed_interesting_cells@meta.data %>%
  tibble::as_tibble(rownames = "barcode") %>%
  dplyr::select(barcode, orig.ident, S.Score, G2M.Score, Phase, cycling, new_celltypes) %>%
  ggplot2::ggplot(aes(x = Phase, y = S.Score)) +
  
  # Right half: density + median line
  ggdist::stat_halfeye(
    side = "right",
    aes(fill = Phase),
    # alpha = 0.5,
    .width = 0.5,           # median line
    justification = -0.2,
    scale = 0.9,
    linewidth = 0.4,
    color = "black"
  ) +
  scale_fill_manual(values = phase_cols) +
  
  # Left half: dots
  ggdist::stat_dots(
    side = "left",
    alpha = 0.3,
    justification = 1.25,
    quantiles = NULL,       # prevents binwidth error
    dotsize = 0.7
  ) +
  
  facet_wrap(~ new_celltypes,
             nrow = 1) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(title = "S.Score",
       caption = "Pre cell-cycle regression")

# stat halfeye plots of G2M.Score for the full data set, facet wrap by cell type
slim_re_embed_interesting_cells@meta.data %>%
  tibble::as_tibble(rownames = "barcode") %>%
  dplyr::select(barcode, orig.ident, S.Score, G2M.Score, Phase, cycling, new_celltypes) %>%
  ggplot2::ggplot(aes(x = Phase, y = G2M.Score)) +
  
  # Right half: density + median line
  ggdist::stat_halfeye(
    side = "right",
    aes(fill = Phase),
    # alpha = 0.5,
    .width = 0.5,           # median line
    justification = -0.2,
    scale = 0.9,
    linewidth = 0.4,
    color = "black"
  ) +
  scale_fill_manual(values = phase_cols) +
  
  # Left half: dots
  ggdist::stat_dots(
    side = "left",
    alpha = 0.3,
    justification = 1.25,
    quantiles = NULL,       # prevents binwidth error
    dotsize = 0.7
  ) +
  
  facet_wrap(~ new_celltypes,
             nrow = 1) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(title = "G2M.Score",
       caption = "Pre cell-cycle regression")

```

### perform the tests  

```{r}
# make a subset object for each of the cell types
# collect the cell type names
cell_types <- unique(slim_re_embed_interesting_cells@meta.data$new_celltypes)

# make the subset objects
# for (ct in cell_types) {
#   obj_name <- paste0("subset_", ct)
#   
#   assign(
#     obj_name,
#     subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
#   )
# }

# testing cell cycle phase % differences between genotypes on individual cell types
propeller_phase_genotype_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$Phase,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$genotype,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_phase_genotype_celltype_results) <- cell_types

# testing cycling vs not-cycling % differences between time genotypes
propeller_cycling_genotype_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$cycling,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$genotype,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_cycling_genotype_celltype_results) <- cell_types

# testing cell cycle phase % differences between time points
propeller_phase_stage_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$Phase,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$stage,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_phase_stage_celltype_results) <- cell_types

# testing cycling vs not-cycling % differences between time points
propeller_cycling_stage_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$cycling,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$stage,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
  })

names(propeller_cycling_stage_celltype_results) <- cell_types

combined_cycling_stage_celltype_results <- bind_rows(propeller_cycling_stage_celltype_results, .id = "cell_type") %>%
  as_tibble()

```

# Find markers
- For each cell type, find markers between cycling and not-cycling cells.  
- do these markers make sense biologically?  

```{r}

# using lapply, run FindMarkers() for all cell types
# this test doesn not take genotype or time point into consideration
marker_results <- lapply(cell_types, function(ct) {
  
  # I know we already made these subset objects, but this makes the function stand alone
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  # make sure "cycling" is the active ident
  Idents(sub_obj) <- "cycling"
  
  res <- FindMarkers(
    sub_obj,
    ident.1 = "cycling",
    ident.2 = "not_cycling",
    logfc.threshold = 0.25,
    min.pct = 0.1
  )
  
  res <- res %>%
    rownames_to_column("gene") %>%   # move rownames to column becaude we need to keep them
    as_tibble() %>%                 # drop rownames completely, messes with bind_rows 
    mutate(cell_type = ct)
  
  return(res)
})

combined_markers <- bind_rows(marker_results)

# top 10 markers by effect size (positive logFC) for each cell type
# positive log2FC means higher in ident.1 -> cycling
top10_markers_up_cycling <- combined_markers %>%
  group_by(cell_type) %>%
  slice_max(avg_log2FC, n = 10)

# top 10 markers by effect size (negative logFC) for each cell type
# positive log2FC means higher in ident.2 -> not-cycling
top10_markers_up_not_cycling <- combined_markers %>%
  group_by(cell_type) %>%
  slice_min(avg_log2FC, n = 10)
```

## Find markers between genotypes for cycling and not-cycling cells for each cell type subset for each time point

```{r}
# Configuration
MIN_CELLS      <- 30
LOGFC_THRESH   <- 0.0
PADJ_THRESH    <- 0.05
IDENT_1        <- "control"
IDENT_2        <- "Mib1_KO"

# Get all combinations
cell_types <- unique(slim_re_embed_interesting_cells@meta.data$new_celltypes)
stages     <- unique(slim_re_embed_interesting_cells@meta.data$stage)
cyclings   <- unique(slim_re_embed_interesting_cells@meta.data$cycling)

combos <- expand.grid(
  cell_type = cell_types,
  stage     = stages,
  cycling   = cyclings,
  stringsAsFactors = FALSE
)

# Run FindMarkers() across all combinations
results_list      <- list()   # significant only  → heatmap, UpSet
results_list_full <- list()   # all genes         → fgsea

for (i in seq_len(nrow(combos))) {
  
  ct  <- combos$cell_type[i]
  st  <- combos$stage[i]
  cy  <- combos$cycling[i]
  key <- paste(ct, st, cy, sep = "__")
  
  meta <- slim_re_embed_interesting_cells@meta.data
  
  cells_group <- rownames(meta)[
    meta$new_celltypes == ct &
      meta$stage         == st &
      meta$cycling       == cy
  ]
  
  if (length(cells_group) == 0) next
  
  sub_obj <- subset(slim_re_embed_interesting_cells, cells = cells_group)
  
  genotype_counts <- table(sub_obj@meta.data$genotype)
  n_ident1 <- as.integer(genotype_counts[IDENT_1])
  n_ident2 <- as.integer(genotype_counts[IDENT_2])
  
  if (is.na(n_ident1) || is.na(n_ident2)) next
  if (n_ident1 < MIN_CELLS || n_ident2 < MIN_CELLS) next
  
  Idents(sub_obj) <- "genotype"
  
  markers <- tryCatch(
    FindMarkers(
      sub_obj,
      ident.1         = IDENT_1,
      ident.2         = IDENT_2,
      logfc.threshold = LOGFC_THRESH,
      min.pct         = 0.1,
      test.use        = "wilcox"
    ),
    error = function(e) NULL
  )
  
  if (is.null(markers) || nrow(markers) == 0) next
  
  # Annotate before splitting
  markers$gene       <- rownames(markers)
  markers$cell_type  <- ct
  markers$stage      <- st
  markers$cycling    <- cy
  markers$comparison <- paste0(IDENT_1, "_vs_", IDENT_2)
  
  # Save full results for fgsea
  results_list_full[[key]] <- markers
  
  # Save significant only for heatmap / UpSet
  markers_sig <- markers[markers$p_val_adj < PADJ_THRESH, ]
  if (nrow(markers_sig) == 0) next
  results_list[[key]] <- markers_sig
}

cat("Groups with any DEGs tested:      ", length(results_list_full), "\n")
cat("Groups with significant DEGs:     ", length(results_list), "\n")

```

# Volcano plots

```{r}
# ── Volcano plot function ───────────────────────────────────────────────────

make_volcano <- function(df,
                         title,
                         padj_thresh  = 0.05,
                         lfc_thresh   = 0.25,
                         n_label      = 10,
                         ident1_label = "control",
                         ident2_label = "Mib1_KO") {
  
  # guard: must have required columns
  stopifnot(all(c("gene", "avg_log2FC", "p_val_adj") %in% colnames(df)))
  
  # replace zero adjusted p-values to avoid -log10(0) = Inf
  # use the smallest non-zero value in the data as a floor
  min_nonzero <- min(df$p_val_adj[df$p_val_adj > 0], na.rm = TRUE)
  df <- df %>%
    dplyr::mutate(
      p_val_adj_plot = dplyr::if_else(p_val_adj == 0, min_nonzero, p_val_adj),
      neg_log10_padj = -log10(p_val_adj_plot),
      
      # classify each gene
      direction = dplyr::case_when(
        p_val_adj < padj_thresh & avg_log2FC >  lfc_thresh ~ paste0("Up in ", ident1_label),
        p_val_adj < padj_thresh & avg_log2FC < -lfc_thresh ~ paste0("Up in ", ident2_label),
        TRUE                                                ~ "NS"
      ),
      direction = factor(direction,
                         levels = c(paste0("Up in ", ident1_label),
                                    paste0("Up in ", ident2_label),
                                    "NS"))
    )
  
  # top n genes to label (by adjusted p-value, among significant only)
  label_genes <- df %>%
    dplyr::filter(direction != "NS") %>%
    dplyr::slice_min(p_val_adj_plot, n = n_label) %>%
    dplyr::pull(gene)
  
  df <- df %>%
    dplyr::mutate(label = dplyr::if_else(gene %in% label_genes, gene, NA_character_))
  
  # colour palette
  cols <- setNames(
    c("#E41A1C", "#377EB8", "grey70"),
    c(paste0("Up in ", ident1_label),
      paste0("Up in ", ident2_label),
      "NS")
  )
  
  # count sig genes for subtitle
  n_up1 <- sum(df$direction == paste0("Up in ", ident1_label), na.rm = TRUE)
  n_up2 <- sum(df$direction == paste0("Up in ", ident2_label), na.rm = TRUE)
  
  ggplot(df, aes(x = avg_log2FC,
                 y = neg_log10_padj,
                 colour = direction)) +
    geom_point(size = 1.2, alpha = 0.6) +
    scale_colour_manual(values = cols, name = NULL) +
    
    # significance threshold lines
    geom_hline(yintercept = -log10(padj_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    
    # gene labels
    ggrepel::geom_text_repel(
      aes(label = label),
      size          = 2.8,
      max.overlaps  = 20,
      segment.size  = 0.3,
      segment.color = "grey50",
      na.rm         = TRUE,
      show.legend   = FALSE
    ) +
    
    labs(
      title    = title,
      subtitle = glue::glue(
        "Up in {ident1_label}: {n_up1}  |  Up in {ident2_label}: {n_up2}",
        "  |  padj < {padj_thresh}, |log2FC| > {lfc_thresh}"
      ),
      x = expression(log[2]~"fold change"),
      y = expression(-log[10]~"(adjusted p-value)")
    ) +
    theme_bw(base_size = 11) +
    theme(
      plot.title    = element_text(size = 10, face = "bold"),
      plot.subtitle = element_text(size = 8,  colour = "grey40"),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
}

# ── Generate and save all volcano plots ────────────────────────────────────
# results_list_full contains all genes (needed for proper volcano plots —
# using results_list (sig only) would hide the NS cloud entirely)

volcano_out_dir <- file.path(out_dir, "volcano_plots")
if (!dir.exists(volcano_out_dir)) dir.create(volcano_out_dir, recursive = TRUE)

# iterate over all tested groups
for (key in names(results_list_full)) {
  
  df  <- results_list_full[[key]]
  
  # parse the key back into readable components
  # key format: "CellType__Stage__cycling"
  parts <- stringr::str_split_fixed(key, "__", n = 3)
  ct    <- parts[1]
  st    <- parts[2]
  cy    <- parts[3]
  
  plot_title <- glue::glue(
    "{ct}  |  {st}  |  {cy} cells\ncontrol vs Mib1_KO"
  )
  
  p <- make_volcano(
    df           = df,
    title        = plot_title,
    ident1_label = "control",
    ident2_label = "Mib1_KO"
  )
  
  # sanitise key for use as a filename
  safe_key  <- stringr::str_replace_all(key, "[^A-Za-z0-9_-]", "_")
  file_path <- file.path(volcano_out_dir, paste0("volcano__", safe_key, ".pdf"))
  
  ggsave(
    filename = file_path,
    plot     = p,
    width    = 7,
    height   = 5.5,
    device   = "pdf"
  )
}

cat("Saved", length(results_list_full), "volcano plots to:\n", volcano_out_dir, "\n")


```


# Heatmap visualisation

```{r}
summary_df <- do.call(rbind, lapply(names(results_list), function(key) {
  df    <- results_list[[key]]
  parts <- strsplit(key, "__")[[1]]
  data.frame(
    cell_type = parts[1],
    stage     = parts[2],
    cycling   = parts[3],
    n_up      =  sum(df$avg_log2FC > 0), # higher in control
    n_down    = -sum(df$avg_log2FC < 0), # higher in Mib1_KO
    n_total   =  nrow(df),
    key       = key,
    stringsAsFactors = FALSE
  )
  }))

summary_df <- summary_df %>%
  mutate(row_label = paste0(cell_type, " | ", stage, " | ", cycling)) %>%
  arrange(cell_type, stage, cycling)

mat <- cbind(Up_Control = summary_df$n_up, Up_Mib1KO = summary_df$n_down)
rownames(mat) <- summary_df$row_label

row_ann <- rowAnnotation(
  CellType = summary_df$cell_type,
  col      = list(CellType = celltype_cols),
  annotation_legend_param = list(CellType = list(title = "Cell Type")),
  show_annotation_name = TRUE
)

stage_cols <- setNames(
  colorRampPalette(c("#FEE08B", "#3288BD"))(length(unique(summary_df$stage))),
  sort(unique(summary_df$stage))
)

row_ann2 <- rowAnnotation(
  Stage   = summary_df$stage,
  Cycling = summary_df$cycling,
  col = list(
    Stage   = stage_cols,
    Cycling = c("cycling" = "#D95F02", "not_cycling" = "#7570B3")
  )
)

col_fun <- colorRamp2(
  c(min(mat), 0, max(mat)),
  c("#2166AC", "white", "#D6604D")
)

# generate the heatmap object
ht <- Heatmap(
  mat,
  name                = "DEG count\n(red = up in Control\nblue = up in Mib1_KO)",
  col                 = col_fun,
  cluster_rows        = TRUE,
  cluster_columns     = FALSE,
  show_row_names      = TRUE,
  row_names_gp        = gpar(fontsize = 8),
  column_names_gp     = gpar(fontsize = 10, fontface = "bold"),
  left_annotation     = row_ann,
  right_annotation    = row_ann2,
  rect_gp             = gpar(col = "white", lwd = 0.5),
  row_title           = NULL,
  column_title        = "Control vs Mib1_KO — significant DEGs per group",
  column_title_gp     = gpar(fontsize = 12, fontface = "bold"),
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- mat[i, j]
    if (abs(val) > 0)
      grid.text(abs(val), x, y, gp = gpar(fontsize = 7, col = "black"))
  }
)
# plot and save the heatmap
png(file.path(out_dir, "heatmap_control_vs_mib1ko.png"),
    width = 10,
    height = 12,
    units = "in",
    res = 300)
draw(ht,
     heatmap_legend_side = "right",
     annotation_legend_side = "right")
dev.off()
```

# fgsea and visualisation 

```{r}
# Supply your cycling gene set
cycling_genes <- unique(c(g2m.genes, s.genes))
# convert to a list as required by
gene_sets <- list(CyclingSignature = cycling_genes)

# Run fgsea across all groups
fgsea_results <- lapply(names(results_list_full), function(key) {
  df <- results_list_full[[key]]   # full unfiltered gene list
  
  df <- df %>%
    mutate(rank_metric = sign(avg_log2FC) * -log10(p_val_adj + 1e-300)) %>%
    arrange(desc(rank_metric))
  
  ranked_genes <- setNames(df$rank_metric, df$gene)
  ranked_genes <- ranked_genes[!duplicated(names(ranked_genes))]
  
  # Check overlap before running
  n_overlap <- length(intersect(names(ranked_genes), gene_sets$CyclingSignature))
  if (n_overlap < 3) return(NULL)
  
  res <- tryCatch(
    fgsea(
      pathways    = gene_sets,
      stats       = ranked_genes,
      minSize     = 3,
      maxSize     = 500,
      nPermSimple = 10000
    ),
    error = function(e) NULL
  )
  
  if (is.null(res) || nrow(res) == 0) return(NULL)
  
  parts <- strsplit(key, "__")[[1]]
  res$cell_type <- parts[1]
  res$stage     <- parts[2]
  res$cycling   <- parts[3]
  res$key       <- key
  res
})

fgsea_df <- do.call(rbind, Filter(Negate(is.null), fgsea_results))

cat("Groups with fgsea results:", nrow(fgsea_df), "\n")
cat("Significant (padj < 0.05):", sum(fgsea_df$padj < 0.05, na.rm = TRUE), "\n")

# Check overlap per group
overlap_check <- sapply(names(results_list), function(key) {
  genes <- results_list[[key]]$gene
  length(intersect(genes, gene_sets$CyclingSignature))
})

print(sort(overlap_check))
cat("\nGroups with overlap >= 5:", sum(overlap_check >= 5), "\n")
cat("Groups with overlap < 5:", sum(overlap_check < 5), "\n")
cat("Groups with overlap == 0:", sum(overlap_check == 0), "\n")


# Plot
fgsea_plot_df <- fgsea_df %>%
  filter(pathway == "CyclingSignature") %>%
  mutate(
    sig       = ifelse(padj < 0.05, "padj < 0.05", "NS"),
    direction = ifelse(NES > 0, "Enriched in Control", "Enriched in Mib1_KO"),
    label     = paste0(cell_type, " | ", stage, " | ", cycling)
  ) %>%
  arrange(NES)

png(file.path(out_dir, "GSEA_enrichment_control_vs_mib1ko.png"),
    width = 12,
    height = 10,
    units = "in",
    res = 300)
ggplot(fgsea_plot_df, aes(x = reorder(label, NES), y = NES, fill = direction)) +
  geom_col(aes(alpha = sig), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(
    "Enriched in Control" = "#D6604D",
    "Enriched in Mib1_KO" = "#2166AC"
  )) +
  scale_alpha_manual(
    values = c("padj < 0.05" = 1, "NS" = 0.35),
    name   = "Significance"
  ) +
  coord_flip() +
  labs(
    title    = "fgsea — Cycling gene signature enrichment",
    subtitle = "Control vs Mib1_KO, per cell type × stage × cycling group",
    caption  = "Gene set = unique(S.Score + G2M.Score) genes",
    x        = NULL,
    y        = "Normalised Enrichment Score (NES)",
    fill     = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )
dev.off()
```

# Upset plot of common genes

```{r}

# Build gene × group membership matrix
# Collect all unique DEGs across all groups
all_genes <- unique(unlist(lapply(results_list, function(df) df$gene)))

# Binary membership matrix
upset_mat <- sapply(results_list, function(df) {
  as.integer(all_genes %in% df$gene)
})
rownames(upset_mat) <- all_genes

upset_df <- as.data.frame(upset_mat)

# Optional: separate up- and down-regulated
# Up = higher in Mib1_KO
upset_mat_up <- sapply(results_list, function(df) {
  as.integer(all_genes %in% df$gene[df$avg_log2FC > 0])
})
rownames(upset_mat_up) <- all_genes
upset_df_up <- as.data.frame(upset_mat_up)

# Down = higher in Control
upset_mat_down <- sapply(results_list, function(df) {
  as.integer(all_genes %in% df$gene[df$avg_log2FC < 0])
})
rownames(upset_mat_down) <- all_genes
upset_df_down <- as.data.frame(upset_mat_down)

# Plot — all DEGs
png(file.path(out_dir, "08-upset_all_deg.png"),
    width = 10,
    height = 10,
    units = "in",
    res = 300)
upset(
  upset_df,
  nsets         = min(20, ncol(upset_df)),   # top 20 groups by size
  nintersects   = 40,
  order.by      = "freq",
  decreasing    = TRUE,
  point.size    = 2.5,
  line.size     = 0.8,
  text.scale    = c(1.2, 1, 1, 1, 1, 0.9),  # title, axis, etc.
  mainbar.y.label = "Shared DEG count",
  sets.x.label    = "DEGs per group",
  mb.ratio        = c(0.6, 0.4)
)
dev.off()

# Plot — upregulated in Mib1_KO only
png(file.path(out_dir, "08-upset_up_ko_deg.png"),
    width = 10,
    height = 10,
    units = "in",
    res = 300)
upset(
  upset_df_up,
  nsets         = min(20, ncol(upset_df_up)),
  nintersects   = 40,
  order.by      = "freq",
  main.bar.color = "#D6604D",
  point.size    = 2.5,
  line.size     = 0.8,
  mainbar.y.label = "Shared DEGs (up in Mib1_KO)",
  sets.x.label    = "DEGs per group"
)
dev.off()
# Plot — upregulated in Control only
png(file.path(out_dir, "08-upset_up_wt_deg.png"),
    width = 10,
    height = 10,
    units = "in",
    res = 300)
upset(
  upset_df_down,
  nsets         = min(20, ncol(upset_df_down)),
  nintersects   = 40,
  order.by      = "freq",
  main.bar.color = "#2166AC",
  point.size    = 2.5,
  line.size     = 0.8,
  mainbar.y.label = "Shared DEGs (up in Control)",
  sets.x.label    = "DEGs per group"
)
dev.off()
```




