---
title: "Test cell type proportions"
author: "Melanie Smith"
date: "2026-04-17"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Core single cell
library(Seurat)
library(loupeR)

# additional plotting functions
library(patchwork)
library(pheatmap)
library(ggExtra)

```

The aim of this script is to mitigate the effects of cell cycle (CC) heterogeneity in our scRNA-seq data. To accomplish this we first write in the complete Seurat object containing all 4 samples, followed by using the Seurat function `ScaleData()` with `vars.to.regress = c("CC.Difference")` to regress together the "S.Score" and the "G2M.Score". This milder alternative (compared to fully regressing out all CC information) is regressing only the difference (S.Score - G2M.Score). This keeps the broad “cycling vs. quiescent” signal while removing phase-specific noise among proliferating cells.  

# Set project directories

```{r}
# set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "08-regress-cell-cycle")

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

```

# Load required Seurat object
- Load the final Seurat object as created in `06-cellType_filter.Rmd`. This object is post-QC, post-filtering and has layers for PCA, UMAP etc.  
- This object contains all 4 samples and has all layers merged.  

```{r}
# read in the complete Seurat object
re_embed_interesting_cells <- readRDS(file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds"))
# sanity check to ensure this is a Seurat object
stopifnot(inherits(re_embed_interesting_cells, "Seurat"))

# ensure "seurat_cluster" is the current identity
Idents(re_embed_interesting_cells) <- "seurat_clusters"
Idents(re_embed_interesting_cells) <- factor(Idents(re_embed_interesting_cells))

# ensure "RNA" is the default assay
DefaultAssay(re_embed_interesting_cells) <- "RNA"

```

# create the slim object
- I've found this analyses difficult to get through without accumulating enough RAM to crash RStudio.  
- Here I will create a slim version of my Seurat object so that it and all subsequent objects are taking up less space.  

```{r}
# Slim down while keeping exactly what FindMarkers needs
slim_re_embed_interesting_cells <- DietSeurat(
  object = re_embed_interesting_cells,
  
  # Keep only the normalized data layer (this is what FindMarkers uses)
  layers = "data",                    # ← most important line
  
  # Keep only the main assay
  assays = "RNA",                     # or "SCT" if that's your main one
  
  # Keep only the reductions you actually use
  dimreducs = c("pca", "umap"),
  
  # Drop everything else
  graphs = FALSE
)

# remove the large object
rm(re_embed_interesting_cells)
# clean up
gc()
```


## Set up the cell-type and phase colours for consistenc
- setting the colours here to make sure they are consistent across all scripts.  

```{r}
celltype_cols <- c(
  "Myocardium" = "#E41A1C",
  "VSMC" = "#377EB8", 
  "Epicardium" = "#4DAF4A",
  "Mesenchyme" = "#984EA3", 
  "Endocardium" = "#FF7F00",
  "Other/Unknown" = "grey70"
  )

phase_cols <- c(
  "G1" = "#1f77b4",
  "G2M" = "#2ca02c",
  "S" = "#ff7f0e"
  )
```

# Quick sanity check

```{r}
print(slim_re_embed_interesting_cells)
# which ident is currently active
table(Idents(slim_re_embed_interesting_cells))
stopifnot(!anyDuplicated(colnames(slim_re_embed_interesting_cells)))
```

# Collaps cell cycle phase into cycling vs not-cycling

```{r}
# set the cell cycle phase as a factor to ensure plots are always in the same order
slim_re_embed_interesting_cells$Phase <- factor(
  slim_re_embed_interesting_cells$Phase,
  levels = c("G1", "G2M", "S")  # match the order in phase_cols exactly
)

## collapse the S and G2M phase allocations
slim_re_embed_interesting_cells[[]] <- slim_re_embed_interesting_cells[[]] %>%
  dplyr::mutate(
    cycling = dplyr::if_else(
      Phase %in% c("S", "G2M"),
      "cycling",
      "not_cycling"
    )
  )

# lets make these a factor so they always plot in the same order
slim_re_embed_interesting_cells$cycling <- factor(
  slim_re_embed_interesting_cells$cycling,
  levels = c("not_cycling", "cycling")
)

# quick sanity check
table(slim_re_embed_interesting_cells$Phase,
      slim_re_embed_interesting_cells$cycling)
  #     cycling not_cycling
  # G1        0       12436
  # G2M    5796           0
  # S      6095           0
```

# plot collapse cell cycle phase

```{r}

DimPlot(slim_re_embed_interesting_cells,
        group.by = "cycling",
        # split.by = "orig.ident",
        reduction = "umap",
        label = FALSE,
        repel = TRUE,
        label.size = 3) +
  plot_annotation(
    title = "Cell Cycle Phase",
    caption = "Final dataset"
  )

```

# Test cell cycle proportions
## Combined cell types
- pool time points or genotypes to create pseudo biological replicates and test differences in cell type proportions.  
- important caveat here -> we don't "really" have biological replicates so we need to always keep in mind the complexity introduces by timepoint and genotype.  

```{r}
# Q1. Are there differences in the proportion of cells in each cycle phase (G1, S, and G2M)) between the different genotypes?
propeller_phase_genotype_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$Phase,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$genotype,       # condition of interest
  transform = "asin")   

# Q2. Are there differences in the proportion of cycling cells (cycling or not-cycling) between the different genotypes?
propeller_cycling_genotype_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$cycling,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$genotype,       # condition of interest
  transform = "asin")  

# Q3. Does the developmental stage affect the proportion cell cycle phase (G1, S, and G2M) between timepoints?
propeller_phase_stage_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$Phase,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$stage,       # condition of interest
  transform = "asin")   

# Q4. Does the developmental stage affect the proportion of cycling cells (cycling or not-cycling) between timepoints?
propeller_cycling_stage_results <- speckle::propeller(
  clusters = slim_re_embed_interesting_cells$cycling,           # what you want to test proportions of
  sample   = slim_re_embed_interesting_cells$orig.ident,      # biological replicates
  group    = slim_re_embed_interesting_cells$stage,       # condition of intrest
  transform = "asin")

# speckle style bar plots (per cycling/not-cycling)
props_cycling <- speckle::getTransformedProps(
  slim_re_embed_interesting_cells$cycling, slim_re_embed_interesting_cells$orig.ident,
  transform="asin")
barplot(props_cycling$Proportions,
        col = c("orange","purple"),
        legend=TRUE, 
        ylab="Proportions")

# speckle style bar plots (per cell cycle phase)
props_phase <- speckle::getTransformedProps(
  slim_re_embed_interesting_cells$Phase, slim_re_embed_interesting_cells$orig.ident,
  transform="asin")
barplot(props_phase$Proportions,
        col = phase_cols,
        legend=TRUE, 
        ylab="Proportions")

```

## Individual cell types
- Subset each of the cell types and ask the proportion question again for each.  
- The hope is that this retains cell type specific behaviour that might be masked at the higher level.  
- Keep in mind that we still can't interpret the p-values because we don't have true biological replicates. 

### Quick visualisation
- Stat half-eye plots showing the distribution of S.Score and G2M.Score for each of the cell types.  
- Keep in mind each cell type contains different time points and different genotypes.  

```{r}
# stat halfeye plots of S.Score for the full data set, facet wrap by cell type
slim_re_embed_interesting_cells@meta.data %>%
  tibble::as_tibble(rownames = "barcode") %>%
  dplyr::select(barcode, orig.ident, S.Score, G2M.Score, Phase, cycling, new_celltypes) %>%
  ggplot2::ggplot(aes(x = Phase, y = S.Score)) +
  
  # Right half: density + median line
  ggdist::stat_halfeye(
    side = "right",
    aes(fill = Phase),
    # alpha = 0.5,
    .width = 0.5,           # median line
    justification = -0.2,
    scale = 0.9,
    linewidth = 0.4,
    color = "black"
  ) +
  scale_fill_manual(values = phase_cols) +
  
  # Left half: dots
  ggdist::stat_dots(
    side = "left",
    alpha = 0.3,
    justification = 1.25,
    quantiles = NULL,       # prevents binwidth error
    dotsize = 0.7
  ) +
  
  facet_wrap(~ new_celltypes,
             nrow = 1) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(title = "S.Score",
       caption = "Pre cell-cycle regression")

# stat halfeye plots of G2M.Score for the full data set, facet wrap by cell type
slim_re_embed_interesting_cells@meta.data %>%
  tibble::as_tibble(rownames = "barcode") %>%
  dplyr::select(barcode, orig.ident, S.Score, G2M.Score, Phase, cycling, new_celltypes) %>%
  ggplot2::ggplot(aes(x = Phase, y = G2M.Score)) +
  
  # Right half: density + median line
  ggdist::stat_halfeye(
    side = "right",
    aes(fill = Phase),
    # alpha = 0.5,
    .width = 0.5,           # median line
    justification = -0.2,
    scale = 0.9,
    linewidth = 0.4,
    color = "black"
  ) +
  scale_fill_manual(values = phase_cols) +
  
  # Left half: dots
  ggdist::stat_dots(
    side = "left",
    alpha = 0.3,
    justification = 1.25,
    quantiles = NULL,       # prevents binwidth error
    dotsize = 0.7
  ) +
  
  facet_wrap(~ new_celltypes,
             nrow = 1) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(title = "G2M.Score",
       caption = "Pre cell-cycle regression")

```

### perform the tests  

```{r}
# make a subset object for each of the cell types
# collect the cell type names
cell_types <- unique(slim_re_embed_interesting_cells@meta.data$new_celltypes)

# make the subset objects
# for (ct in cell_types) {
#   obj_name <- paste0("subset_", ct)
#   
#   assign(
#     obj_name,
#     subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
#   )
# }

# testing cell cycle phase % differences between genotypes on individual cell types
propeller_phase_genotype_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$Phase,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$genotype,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_phase_genotype_celltype_results) <- cell_types

# testing cycling vs not-cycling % differences between time genotypes
propeller_cycling_genotype_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$cycling,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$genotype,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_cycling_genotype_celltype_results) <- cell_types

# testing cell cycle phase % differences between time points
propeller_phase_stage_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$Phase,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$stage,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
})

names(propeller_phase_stage_celltype_results) <- cell_types

# testing cycling vs not-cycling % differences between time points
propeller_cycling_stage_celltype_results <- lapply(cell_types, function(ct) {
  
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  
  res <- speckle::propeller(
    clusters = sub_obj$cycling,
    sample   = sub_obj$orig.ident,
    group    = sub_obj$stage,
    transform = "asin"
  )
  
  res$cell_type <- ct  # keep track of origin
  
  return(res)
  })

names(propeller_cycling_stage_celltype_results) <- cell_types

combined_cycling_stage_celltype_results <- bind_rows(propeller_cycling_stage_celltype_results, .id = "cell_type") %>%
  as.tibble()

```

# Find markers
- For each cell type, find markers between cycling and not-cycling cells.  
- do these markers make sense biologically?  

```{r}

# using lapply, run FindMarkers() for all cell types
# this test doesn not take genotype or time point into consideration
marker_results <- lapply(cell_types, function(ct) {
  
  # I know we already made these subset objects, but this makes the function stand alone
  sub_obj <- subset(slim_re_embed_interesting_cells, subset = new_celltypes == ct)
  # make sure "cycling" is the active ident
  Idents(sub_obj) <- "cycling"
  
  res <- FindMarkers(
    sub_obj,
    ident.1 = "cycling",
    ident.2 = "not_cycling",
    logfc.threshold = 0.25,
    min.pct = 0.1
  )
  
  res <- res %>%
    rownames_to_column("gene") %>%   # move rownames to column becaude we need to keep them
    as_tibble() %>%                 # drop rownames completely, messes with bind_rows 
    mutate(cell_type = ct)
  
  return(res)
})

combined_markers <- bind_rows(marker_results)

# top 10 markers by effect size (positive logFC) for each cell type
# positive log2FC means higher in ident.1 -> cycling
top10_markers_up_cycling <- combined_markers %>%
  group_by(cell_type) %>%
  slice_max(avg_log2FC, n = 10)

# top 10 markers by effect size (negative logFC) for each cell type
# positive log2FC means higher in ident.2 -> not-cycling
top10_markers_up_not_cycling <- combined_markers %>%
  group_by(cell_type) %>%
  slice_min(avg_log2FC, n = 10)
```

## Find markers between genotypes for cycling and not-cycling cells for each cell type subset for each time point

```{r}
# ── Configuration ─────────────────────────────────────────────────────────────
MIN_CELLS      <- 3          # minimum cells per genotype to attempt comparison
LOGFC_THRESH   <- 0.25
PADJ_THRESH    <- 0.05
IDENT_1        <- "control"
IDENT_2        <- "Mib1_KO"


# ── Get all combinations ───────────────────────────────────────────────────────
cell_types <- unique(slim_re_embed_interesting_cells@meta.data$new_celltypes)
stages     <- unique(slim_re_embed_interesting_cells@meta.data$stage)
cyclings   <- unique(slim_re_embed_interesting_cells@meta.data$cycling)

combos <- expand.grid(
  cell_type = cell_types,
  stage     = stages,
  cycling   = cyclings,
  stringsAsFactors = FALSE
)

# ── Run FindMarkers across all combinations ────────────────────────────────────
results_list <- list()

for (i in seq_len(nrow(combos))) {
  
  ct  <- combos$cell_type[i]
  st  <- combos$stage[i]
  cy  <- combos$cycling[i]
  key <- paste(ct, st, cy, sep = "__")
  
  # --- Subset cells matching this combination ---------------------------------
  meta <- slim_re_embed_interesting_cells@meta.data
  
  cells_group <- rownames(meta)[
    meta$new_celltypes == ct &
      meta$stage         == st &
      meta$cycling       == cy
  ]
  
  if (length(cells_group) == 0) next
  
  sub_obj <- subset(slim_re_embed_interesting_cells, cells = cells_group)
  
  # --- Check each genotype has enough cells ----------------------------------
  genotype_counts <- table(sub_obj@meta.data$genotype)
  
  n_ident1 <- as.integer(genotype_counts[IDENT_1])
  n_ident2 <- as.integer(genotype_counts[IDENT_2])
  
  if (is.na(n_ident1) || is.na(n_ident2)) next
  if (n_ident1 < MIN_CELLS || n_ident2 < MIN_CELLS) next
  
  # --- Set genotype as active identity and run FindMarkers -------------------
  Idents(sub_obj) <- "genotype"
  
  markers <- tryCatch(
    FindMarkers(
      sub_obj,
      ident.1        = IDENT_1,
      ident.2        = IDENT_2,
      logfc.threshold = LOGFC_THRESH,
      min.pct        = 0.1,
      test.use       = "wilcox"   # default; swap for "MAST", "DESeq2", etc.
    ),
    error = function(e) NULL
  )
  
  if (is.null(markers) || nrow(markers) == 0) next
  
  # --- Annotate and filter ---------------------------------------------------
  markers$gene      <- rownames(markers)
  markers$cell_type <- ct
  markers$stage     <- st
  markers$cycling   <- cy
  markers$comparison <- paste0(IDENT_1, "_vs_", IDENT_2)
  
  # Keep only significant hits
  markers <- markers[markers$p_val_adj < PADJ_THRESH, ]
  
  if (nrow(markers) == 0) next
  
  results_list[[key]] <- markers
}

# ── Summary ───────────────────────────────────────────────────────────────────
cat("Comparisons with significant DEGs:", length(results_list), "\n")
cat("Groups tested:", nrow(combos), "\n")

# Quick peek at which groups yielded results
if (length(results_list) > 0) {
  cat("\nGroups with hits:\n")
  print(names(results_list))
}

# Optional: collapse everything into one data frame for easy export
all_results_df <- do.call(rbind, results_list)
rownames(all_results_df) <- NULL

```

