---
title: "gene_expression_gsea_genotype_cellcycle"
author: "Melanie Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE))
gc()
options(
  max.print              = .Machine$integer.max,
  scipen                 = 999,
  stringsAsFactors       = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42)
```

# Load libraries

```{r}
# Core tidyverse
library(tidyverse)

# Annotation
library(org.Mm.eg.db)
library(AnnotationDbi)

# GSEA
library(clusterProfiler)

# Plotting
library(enrichplot)
```

# Set project directories

```{r}
project_name     <- "20251104_sophieWiszniak_ncc_pa"
base_repo_dir    <- "/home/melanie-smith/workDir/sophieWiszniak"
project_repo_dir <- file.path(base_repo_dir, project_name)

# Input: DE results from script 08
de_dir <- file.path(project_repo_dir, "outDir", "07-gene_expression_de_genotype_cellcycle")

# Output: GSEA results go to their own directory
out_dir <- file.path(project_repo_dir, "outDir", "09-gsea_genotype_cellcycle")

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

# Define parameters

```{r}
cell_types     <- c("Mesenchyme", "Myocardium", "Endocardium", "Epicardium", "VSMC")
timepoints     <- c("e11", "e12")
cycle_states   <- c("cycling", "not_cycling")

# gseGO parameters
gsea_pval_cutoff  <- 0.05   # p-value cutoff for significant terms
gsea_min_gs_size  <- 10     # minimum gene set size
gsea_max_gs_size  <- 500    # maximum gene set size
gsea_n_perm       <- 1000   # number of permutations
simplify_cutoff   <- 0.7    # semantic similarity threshold for GO term redundancy removal

# Number of top terms to show on dotplot
n_top_terms <- 20
```

# Load DE results from CSV

Rebuilds `all_de_results` from saved CSVs from script 08

```{r}
all_de_results <- list()
csv_files <- list.files(de_dir, pattern = "_de_results\\.csv$", full.names = TRUE)

if (length(csv_files) == 0) {
  stop("No DE result CSVs found in: ", de_dir, "\nRun 07-gene_expression_de_genotype_cellcycle.Rmd first.")
}

for (f in csv_files) {
  key <- basename(f) %>% stringr::str_remove("_de_results\\.csv$")
  all_de_results[[key]] <- read_csv(f, show_col_types = FALSE)
}

message("Loaded ", length(all_de_results), " DE result sets.")
```

# Helper: convert gene symbols to Entrez IDs

clusterProfiler's `gseGO()` requires Entrez IDs. This function maps a
character vector of gene symbols to Entrez IDs using the mouse annotation
database, returning a named vector suitable for use as the GSEA universe or
ranked gene list.

```{r}
symbols_to_entrez <- function(gene_symbols) {
  mapped <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys      = gene_symbols,
    column    = "ENTREZID",
    keytype   = "SYMBOL",
    multiVals = "first"
  )
  # Drop any genes that failed to map
  mapped <- mapped[!is.na(mapped)]
  return(mapped)
}
```

# Run GSEA: GO:BP for all 20 comparisons

For each comparison:
1. Rank genes using: `sign(avg_log2FC) * -log10(p_val_adj)`
2. Map gene symbols to Entrez IDs
3. Set the universe to only genes expressed in that comparison (i.e. all genes
   tested, not the whole genome)
4. Run `gseGO()` on GO:BP
5. Save results CSV and dotplot PNG

```{r, message=FALSE, warning=FALSE}
all_gsea_results <- list()

for (ct in cell_types) {
  for (tp in timepoints) {
    for (cs in cycle_states) {
      
      comparison_label <- paste0(ct, "_", tp, "_KO_vs_control_", cs)
      message("Running GSEA: ", comparison_label)
      
      # --- Load DE results ---
      if (!comparison_label %in% names(all_de_results)) {
        warning("Skipping ", comparison_label, " — DE results not found.")
        next
      }
      
      de <- all_de_results[[comparison_label]]
      
      # --- Sanity check: required columns present ---
      required_cols <- c("gene", "avg_log2FC", "p_val_adj")
      if (!all(required_cols %in% colnames(de))) {
        warning("Skipping ", comparison_label, " — missing required columns.")
        next
      }
      
      # --- Compute rank metric ---
      de <- de %>%
        dplyr::mutate(
          p_val_adj_floored = dplyr::if_else(p_val_adj == 0, .Machine$double.eps, p_val_adj),
          rank = sign(avg_log2FC) * (-log10(p_val_adj_floored))
        )
      
      # --- Map gene symbols to Entrez IDs ---
      universe_entrez <- symbols_to_entrez(de$gene)
      
      if (length(universe_entrez) < gsea_min_gs_size) {
        warning("Skipping ", comparison_label, " — too few genes mapped to Entrez IDs.")
        next
      }
      
      de_mapped <- de %>%
        dplyr::filter(gene %in% names(universe_entrez)) %>%
        dplyr::mutate(entrez = universe_entrez[gene]) %>%
        dplyr::arrange(desc(rank))
      
      ranked_list <- de_mapped$rank
      names(ranked_list) <- de_mapped$entrez
      ranked_list <- ranked_list[!duplicated(names(ranked_list))]
      
      # --- Run gseGO ---
      gsea_result <- tryCatch(
        clusterProfiler::gseGO(
          geneList      = ranked_list,
          OrgDb         = org.Mm.eg.db,
          ont           = "BP",
          keyType       = "ENTREZID",
          minGSSize     = gsea_min_gs_size,
          maxGSSize     = gsea_max_gs_size,
          pvalueCutoff  = gsea_pval_cutoff,
          pAdjustMethod = "BH",
          nPermSimple   = gsea_n_perm,
          verbose       = FALSE
        ),
        error = function(e) {
          warning("gseGO failed for ", comparison_label, ": ", conditionMessage(e))
          return(NULL)
        }
      )
      
      if (is.null(gsea_result) || nrow(as.data.frame(gsea_result)) == 0) {
        message("No significant GO:BP terms found for: ", comparison_label)
        next
      }
      
      # --- Simplify: remove redundant GO terms by semantic similarity ---
      gsea_result_simplified <- tryCatch(
        clusterProfiler::simplify(
          gsea_result,
          cutoff     = simplify_cutoff,
          by         = "p.adjust",
          select_fun = min
        ),
        error = function(e) {
          warning("simplify() failed for ", comparison_label, ": ", conditionMessage(e),
                  "\nFalling back to unsimplified result.")
          return(gsea_result)
        }
      )
      
      n_before <- nrow(as.data.frame(gsea_result))
      n_after  <- nrow(as.data.frame(gsea_result_simplified))
      message("  Terms before simplify: ", n_before, " | after: ", n_after)
      
      all_gsea_results[[comparison_label]] <- gsea_result_simplified
      
      # --- Save results CSV (simplified) ---
      write_csv(
        as.data.frame(gsea_result_simplified),
        file.path(out_dir, paste0(comparison_label, "_gsea_GOBP_simplified.csv"))
      )
      
      # --- Dotplot ---
      n_terms <- min(n_top_terms, n_after)
      
      dotplot <- enrichplot::dotplot(
        gsea_result_simplified,
        showCategory = n_terms,
        title        = paste0(ct, " ", tp, " (", cs, "): GO:BP GSEA\nKO vs control (simplified)"),
        font.size    = 9
      )
      
      ggsave(
        dotplot,
        filename = file.path(out_dir, paste0(comparison_label, "_gsea_GOBP_simplified_dotplot.png")),
        width    = 10,
        height   = max(6, n_terms * 0.35),
        units    = "in",
        dpi      = 300
      )
      
      print(dotplot)
      message("Saved: ", comparison_label)
    }
  }
}

message("All GSEA complete. Outputs saved to: ", out_dir)
```

# Enrichment plot if I want one

```{r}
# Generate enrichment plot for a specific term of interest
# comparison_of_interest <- "Mesenchyme_e11_KO_vs_control_cycling"\
comparison_of_interest <- "Endocardium_e11_KO_vs_control_not_cycling"
temp_inspect_results <- all_gsea_results[[comparison_of_interest]] %>% data.frame 

enrichplot::gseaplot2(
  all_gsea_results[[comparison_of_interest]],
  geneSetID = "GO:0007267",   # paste the ID of the term you want
  title     = paste0(comparison_of_interest, ": ", "cell cell signalling")
)
```


# Summary: significant GO:BP terms per comparison

```{r}
gsea_summary <- map_dfr(names(all_gsea_results), function(name) {
  df <- as.data.frame(all_gsea_results[[name]]) %>%
    dplyr::rename(gsea_rank = rank)
  
  # Pre-compute top terms outside tibble() construction
  top_activated <- df %>%
    dplyr::filter(NES > 0) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::slice(1) %>%
    dplyr::pull(Description)
  top_activated <- if (length(top_activated) == 0) NA_character_ else top_activated
  
  top_suppressed <- df %>%
    dplyr::filter(NES < 0) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::slice(1) %>%
    dplyr::pull(Description)
  top_suppressed <- if (length(top_suppressed) == 0) NA_character_ else top_suppressed
  
  tibble(
    comparison     = name,
    n_sig_terms    = nrow(df),
    n_activated    = sum(df$NES > 0),
    n_suppressed   = sum(df$NES < 0),
    top_activated  = top_activated,
    top_suppressed = top_suppressed
  )
})
print(gsea_summary)
write_csv(gsea_summary, file.path(out_dir, "00-GSEA_summary_all_comparisons.csv"))
```

# Build long-format NES table

Combines all comparisons into a single tidy data frame, then filters to GO
terms significant in at least `min_comparisons` comparisons.

```{r}
nes_long <- map_dfr(names(all_gsea_results), function(name) {
  
  df <- as.data.frame(all_gsea_results[[name]]) %>%
    dplyr::rename(gsea_rank = rank)
  
  parts <- stringr::str_match(
    name,
    "^(.+)_(e\\d+)_KO_vs_control_(cycling|not_cycling)$"
  )
  
  if (anyNA(parts)) {
    warning("Could not parse comparison label: ", name, " — skipping.")
    return(NULL)
  }
  
  df %>%
    dplyr::select(Description, NES, p.adjust) %>%
    dplyr::mutate(
      comparison  = name,
      cell_type   = parts[, 2],
      timepoint   = parts[, 3],
      cycle_state = parts[, 4],
      NES_plot    = dplyr::if_else(p.adjust < 0.05, NES, NA_real_)
    )
})

message("Total term-comparison rows before filtering: ", nrow(nes_long))

# --- Filter: keep only terms significant in >= min_comparisons ---
# Minimum number of comparisons a GO term must be significant in to be shown
min_comparisons <- 4
# NES colour scale limits — symmetric around 0 so colour is meaningful
nes_limit <- 2.5  # values beyond +/- this are clipped to the colour scale ends

terms_to_keep <- nes_long %>%
  dplyr::filter(!is.na(NES_plot)) %>%
  dplyr::count(Description) %>%
  dplyr::filter(n >= min_comparisons) %>%
  dplyr::pull(Description)

message("GO terms significant in >= ", min_comparisons, " comparisons: ", length(terms_to_keep))

nes_filtered <- nes_long %>%
  dplyr::filter(Description %in% terms_to_keep)
```

# Hierarchical clustering of GO terms (rows)

Clusters GO terms by their NES profile across all comparisons so that
biologically related pathways appear adjacent on the heatmap.

```{r}
# Pivot to wide matrix for clustering — fill missing/non-sig with 0
nes_matrix <- nes_filtered %>%
  dplyr::select(Description, comparison, NES_plot) %>%
  tidyr::pivot_wider(
    names_from  = comparison,
    values_from = NES_plot,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("Description")

# Hierarchical clustering
row_clust   <- hclust(dist(nes_matrix), method = "complete")
term_order  <- rownames(nes_matrix)[row_clust$order]

# Apply clustering order as factor levels
nes_filtered <- nes_filtered %>%
  dplyr::mutate(Description = factor(Description, levels = term_order))
```

# Build column labels and ordering

Within each timepoint facet, columns are ordered: cycling then not_cycling,
and within each cycle state by cell type.

```{r}
col_order <- expand.grid(
  cell_type   = cell_types,
  cycle_state = cycle_states,
  stringsAsFactors = FALSE
) %>%
  dplyr::mutate(col_label = paste0(cell_type, "\n(", cycle_state, ")"))

nes_filtered <- nes_filtered %>%
  dplyr::mutate(
    col_label = paste0(cell_type, "\n(", cycle_state, ")"),
    col_label = factor(col_label, levels = col_order$col_label)
  )
```

# Plot: NES heatmap faceted by timepoint

```{r, fig.width=16, fig.height=max(8, length(term_order) * 0.28)}
# Clip NES values to colour scale limits for display
nes_filtered <- nes_filtered %>%
  dplyr::mutate(
    NES_clipped = dplyr::case_when(
      is.na(NES_plot)          ~ NA_real_,
      NES_plot >  nes_limit    ~  nes_limit,
      NES_plot < -nes_limit    ~ -nes_limit,
      TRUE                     ~ NES_plot
    )
  )

heatmap_plot <- ggplot(
  nes_filtered,
  aes(x = col_label, y = Description, fill = NES_clipped)
) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick",
    midpoint = 0,
    limits   = c(-nes_limit, nes_limit),
    na.value = "grey90",
    name     = "NES\n(KO / control)"
  ) +
  facet_grid(
    cols   = vars(timepoint),
    scales = "free_x",
    space  = "free_x"
  ) +
  labs(
    title    = "GO:BP GSEA — Normalised Enrichment Score (NES) across KO vs control comparisons",
    subtitle = paste0(
      "Mouse cardiac development | Cell cycle-stratified | ",
      "Terms significant (padj < 0.05) in \u2265 ", min_comparisons, " comparisons | ",
      "Rows ordered by hierarchical clustering | Grey = not significant"
    ),
    x = NULL,
    y = NULL
  ) +
  theme_classic() +
  theme(
    axis.text.x      = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 7),
    strip.text       = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "grey92", colour = NA),
    legend.position  = "right",
    plot.title       = element_text(size = 12, face = "bold"),
    plot.subtitle    = element_text(size = 8, colour = "grey40"),
    panel.spacing    = unit(1, "lines")
  )

print(heatmap_plot)

# Save — height scales with number of GO terms
ggsave(
  heatmap_plot,
  filename = file.path(
    out_dir,
    "KO_vs_control_GOBP_NES_heatmap_by_timepoint_cellcycle.png"
  ),
  width  = 16,
  height = max(8, length(term_order) * 0.28),
  units  = "in",
  dpi    = 300
)

message("Heatmap saved to: ", out_dir)
```

# Export: filtered NES table used for plotting

Saves the underlying data so the heatmap can be recreated or customised
without re-running the full GSEA pipeline.

```{r}
nes_filtered %>%
  dplyr::select(Description, cell_type, timepoint, cycle_state, NES, p.adjust, NES_plot) %>%
  dplyr::arrange(Description, timepoint, cell_type, cycle_state) %>%
  write_csv(file.path(out_dir, "KO_vs_control_GOBP_NES_heatmap_data.csv"))

message("Heatmap data exported.")
```


# Session info

```{r}
sessionInfo()
```
