---
title: "Import Velocyto Loom spliced and unspliced counts and Create Seurat Object"
author: "Melanie Smith"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects includuing hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)

# Single cell
library(Seurat)
# library(SeuratDisk)
library(Matrix)
library(data.table)

# plotting
library(ggokabeito)

library(sceasy)
library(reticulate)
library(anndata)
# library(scCustomize)

```

# Set project directories and parameters

```{r}
# Set sample ID -> Which sample do you want to run the script for?
sample_ID <- "e11_control"
# sample_ID <- "e11_ko"
# sample_ID <- "e12_control"
# sample_ID <- "e12_ko"

# Set directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID)

# Create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

# Set input files
spliced_obj_file <- file.path(out_dir, "velocyto_output", "spliced_counts.tsv.gz")
unspliced_obj_file <- file.path(out_dir, "velocyto_output", "unspliced_counts.tsv.gz")


```

# Import files

# Load the final Seurat object (1 sample only)
- Contains the final set of cells and genes in a per sample object  
- All counts are filtered and normalised  
- The only reason to read this object in is to use the cell and gene (feature) IDs to subset the .loom data

```{r}
seurat_obj <- readRDS(file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID, paste0("re_embed.", sample_ID, ".rds")))

stopifnot(inherits(seurat_obj, "Seurat"))

Idents(seurat_obj) <- "seurat_clusters"
Idents(seurat_obj) <- factor(Idents(seurat_obj))
DefaultAssay(seurat_obj) <- "RNA"

dim(seurat_obj)
# [1] 24767 (features)  5709 (cells)

# lets check out the cell barcode IDs as reported in the seurat object
head(rownames(seurat_obj@meta.data))
# [1] "AAACCCACAGACTCTA-1_1_1" "AAACCCAGTCAGGTAG-1_1_1"
# [3] "AAACCCAGTCTAATCG-1_1_1" "AAACCCAGTGAGTTTC-1_1_1"
# [5] "AAACGAAAGTTACGGG-1_1_1" "AAACGAAGTATCAGGG-1_1_1"
```

# Load the loom data (intron/exon counts)
- we will need to subset this object such that it only contains the high quality cells found in the final seurat object  
- first we import the spliced coutns as .tsv.gz files, then unpack and filter

```{r}
# import the spliced counts
dt_spliced <- fread(spliced_obj_file)
# save the cell barcodes
barcodes <- dt_spliced[[1]]
# clean up the cell barcodes
clean_barcodes <- gsub("^possorted_genome_bam_5Q1UF:", "", barcodes)
clean_barcodes <- gsub("x$", "", clean_barcodes)
# drop the barcodes from column 1
spliced_mat <- as.matrix(dt_spliced[, -1])
# ensure the counts are double -> not character
storage.mode(spliced_mat) <- "double"
# add the cell barcodes into the rownames
rownames(spliced_mat) <- clean_barcodes
# transform the matrix (seurat expects rows=genes and columns=cells)
spliced_mat <- t(spliced_mat)
# ensure our object is a matrix
spliced_mat <- Matrix(spliced_mat, sparse = TRUE)
# create the assay object
spliced_assay <- CreateAssayObject(counts = spliced_mat)
# quick sanity check of the row and column names
rownames(spliced_assay) %>% head()
colnames(spliced_assay) %>% head()

```

- then we import the unspliced counts
```{r}
# import the spliced counts
dt_unspliced <- fread(unspliced_obj_file)
# save the cell barcodes
barcodes <- dt_unspliced[[1]]
# clean up the cell barcodes
clean_barcodes <- gsub("^possorted_genome_bam_5Q1UF:", "", barcodes)
clean_barcodes <- gsub("x", "", clean_barcodes)
# drop the barcodes from column 1
unspliced_mat <- as.matrix(dt_unspliced[, -1])
# ensure the counts are double -> not character
storage.mode(unspliced_mat) <- "double"
# add the cell barcodes into the rownames
rownames(unspliced_mat) <- clean_barcodes
# transform the matrix (seurat expects rows=genes and columns=cells)
unspliced_mat <- t(unspliced_mat)
# ensure our object is a matrix
unspliced_mat <- Matrix(unspliced_mat, sparse = TRUE)
# create the assay object
unspliced_assay <- CreateAssayObject(counts = unspliced_mat)
# quick sanity check of the row and column names
rownames(unspliced_assay) %>% head()
colnames(unspliced_assay) %>% head()

```

# Subset the new objects to retain only high quality cells and non-filtered genes

```{r}

# add the expected suffix back onto the cell barcodes
# add suffix to spliced
colnames(spliced_assay) <- paste0(colnames(spliced_assay), "-1_1_1")
# add suffix to unspliced
colnames(unspliced_assay) <- paste0(colnames(unspliced_assay), "-1_1_1")

target_genes <- rownames(seurat_obj)
target_cells <- colnames(seurat_obj)

# pull counts
spliced_counts <- spliced_assay@counts
unspliced_counts <- unspliced_assay@counts

# clean up objects no longer needed
rm(dt_spliced, spliced_mat, spliced_assay,
   dt_unspliced, unspliced_mat, unspliced_assay)
gc()

# NOW subset strictly and reorder
spliced_counts <- spliced_counts[target_genes, target_cells]
unspliced_counts <- unspliced_counts[target_genes, target_cells]

# add suffix to the gene names so I can rbind the matrices
rownames(spliced_counts) <- paste0(rownames(spliced_counts), "-spliced")
rownames(unspliced_counts) <- paste0(rownames(unspliced_counts), "-unspliced")

combined_counts <- rbind(spliced_counts, unspliced_counts)

# double check the cell barcodes are in the same order before adding metadata
identical(
  colnames(combined_counts),
  colnames(seurat_obj)
)
# [1] TRUE

# Create a new Seurat object from the count matrix
new_seurat <- CreateSeuratObject(counts = combined_counts)

# Add the genotype metadata from the existing Seurat object
new_seurat <- AddMetaData(new_seurat, metadata = seurat_obj@meta.data[, c("orig.ident", "genotype", "stage", "predicted_sex", "new_celltypes", "full_dataset_clusters", "G2M.Score", "S.Score")])

# Verify the new object
print(new_seurat)
head(new_seurat@meta.data)

```

# Normalise, find variable features, scale, PCA, UMAP, cluster

```{r}
n_dims = 30

re_embed_new_seurat <- new_seurat %>%
  NormalizeData(
    assay = "RNA",
    normalization.method = "LogNormalize",
    scale.factor = 10000,
    verbose = FALSE
  ) %>%
  FindVariableFeatures(
    assay = "RNA",
    selection.method = "vst",
    nfeatures = 3000,
    verbose = FALSE
  ) %>%
  ScaleData(
    assay = "RNA",
    verbose = FALSE
  ) %>% 
  RunPCA(
    assay = "RNA",
    verbose = FALSE
  ) %>% 
  FindNeighbors(
    reduction = "pca",
    dims = 1:n_dims,
    verbose = FALSE
  ) %>%
  RunUMAP(
    reduction = "pca",
    dims = 1:n_dims,
    verbose = FALSE
  )

clustering_resolutions <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6)

for (res in clustering_resolutions) {
  clustered <- FindClusters(re_embed_new_seurat,
                            resolution = res,
                            verbose = FALSE)
  
  p1 <- DimPlot(clustered,
               group.by = paste0("RNA_snn_res.", res),
               label = TRUE,
               label.size = 6) +
    # # ggtitle(paste("Resolution:", res)) +
    # scale_color_okabe_ito() +
    labs(
      title = paste("Resolution:", res),
      caption = "Combined spliced/unspliced"
    )
  
  print(p1)
  
p2 <- FeaturePlot(clustered,
              features = c("G2M.Score", "S.Score"),
              reduction = "umap",
              blend = FALSE,
              label = TRUE,
              label.size = 6) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    )

print(p2)
}

DimPlot(re_embed_new_seurat,
        group.by = "full_dataset_clusters",
        label = TRUE,
        label.size = 6)
```


# Save the Seurat object for scVelo

```{r}
Idents(re_embed_new_seurat) <- "full_dataset_clusters"
Idents(re_embed_new_seurat) <- factor(Idents(re_embed_new_seurat))
DefaultAssay(re_embed_new_seurat) <- "RNA"


# Full path for output
# h5ad_file <- file.path(out_dir, paste0(sample_ID, ".h5ad"))

# Convert Assay5 → Assay3 (legacy) so sceasy can handle it
re_embed_new_seurat[["RNA"]] <- as(re_embed_new_seurat[["RNA"]], "Assay")

# Confirm class is now legacy
cat("Assay class:", class(re_embed_new_seurat[["RNA"]]), "\n")  # should print "Assay", not "Assay5"

obj_adata <- as.anndata(re_embed_new_seurat,
                        file_path = out_dir,
                        file_name = paste0(sample_ID, "_spliced_unspliced.h5ad"))
```


# Add the new assays to the existing Seurat object

```{r}
# add the new assays
# seurat_obj[["spliced"]] <- spliced_assay_subset
# seurat_obj[["unspliced"]] <- unspliced_assay_subset
# 
# # sanity check
# class(seurat_obj[["spliced"]]@counts)
# class(seurat_obj[["unspliced"]]@counts)
```



