---
title: "Sample level embeddings and save"
author: "Melanie Smith"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects includuing hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes in my full Seurat object and subsets it into 4 objects, 1 for each sample in the current analysis.  
- This markdown only splits the Seurat object into smaller objects and saves them, it does not refactor or re-embed anything.  

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)

# Plotting
library(patchwork)
```

# Set project directories

```{r}
# Set sample ID -> Which sample do you want to run the script for?
# sample_ID <- "e11_control"
# sample_ID <- "e11_ko"
# sample_ID <- "e12_control"
sample_ID <- "e12_ko"

project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

# Create out directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir,
             recursive = TRUE)
}

if (!dir.exists(file.path(out_dir, sample_ID))) {
  dir.create(file.path(out_dir, sample_ID),
             recursive = TRUE)
}

```

# User defined functions

```{r}
# Helper function (put this at the top of your script)
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object,
           filename = filename,
           width = width,
           height = height,
           dpi = 300)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

## Set up the cell-type colours

```{r}
celltype_cols <- c("Myocardium" = "#E41A1C",
                   "VSMC" = "#984EA3",
                   "Endocardium" = "#377EB8",
                   "Epicardium" = "#FF7F00",
                   "Mesenchyme" = "#4DAF4A",
                   "Other/Unknown" = "grey70")

epicardial_genes <- c("Wt1", "Upk3b", "Apela", "Upk1b", "Aldh1a2")

endocardial_genes <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
```

# Load required Seurat object

```{r}
seurat_obj <- readRDS(file.path(out_dir, paste0(sample_ID, ".rds")))

stopifnot(inherits(seurat_obj, "Seurat"))

Idents(seurat_obj) <- "seurat_clusters"
Idents(seurat_obj) <- factor(Idents(seurat_obj))
DefaultAssay(seurat_obj) <- "RNA"

# Save the existing Seurat clusters
seurat_obj$full_dataset_clusters_no_cc <- seurat_obj$seurat_clusters


```

# Re-embed Normalise, find variable features, scale, PCA, UMAP, cluster and visualise

```{r}
# quick dimplot using the existing clusters
DimPlot(seurat_obj,
        group.by = "seurat_clusters")

n_dims = 30
# Re-computation of graph-based clustering and UMAP on the cleaned dataset
re_embed_seurat_obj <- FindVariableFeatures(seurat_obj,
                                        selection.method = "vst", 
                                        nfeatures = 3000,
                                        verbose = FALSE) %>%
  ScaleData(features = rownames(seurat_obj),
            verbose = FALSE) %>%
  RunPCA(verbose = FALSE) %>%
  FindNeighbors(dims = 1:n_dims,
                verbose = FALSE) %>%
  RunUMAP(dims = 1:n_dims,
          verbose = FALSE)


# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(re_embed_seurat_obj), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(re_embed_seurat_obj)
plot2 <- LabelPoints(plot = plot1,
                     points = top10,
                     repel = TRUE) +
  labs(
    title = paste0(sample_ID, " - new embedding"),
    caption = sample_ID
    ) +
  NoLegend()
plot1
plot2

VizDimLoadings(re_embed_seurat_obj,
               dims = 1:2,
               reduction = "pca")

DimPlot(re_embed_seurat_obj,
        reduction = "pca") + 
  NoLegend()

DimHeatmap(re_embed_seurat_obj,
           dims = 1:6,
           cells = 500,
           balanced = TRUE)

ElbowPlot(re_embed_seurat_obj)

```

# UMAPs - marker genes

```{r}
# plot the umap with old cluster numbers and colours
umap_oldCluster <- DimPlot(re_embed_seurat_obj,
        reduction = "umap",
        group.by = "full_dataset_clusters_no_cc",
        label = TRUE,
        label.size = 6
        ) +
  labs(
    title = paste0(sample_ID, " - no cc - new embedding (old clusters)"),
    caption = sample_ID
  ) +
  NoLegend()
save_if(umap_oldCluster,
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".old_clusters.png")),
        width = 8,
        height = 8)

umap_newCluster <- DimPlot(re_embed_seurat_obj,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = TRUE,
        label.size = 6
        ) +
  labs(
    title = paste0(sample_ID, " - no cc - new embedding (new clusters)"),
    caption = sample_ID
  ) +
  NoLegend()
save_if(umap_newCluster,
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".new_clusters.png")),
        width = 8,
        height = 8)

# plot the umap with new celltypes
umap_newCelltype <- DimPlot(re_embed_seurat_obj,
        reduction = "umap",
        group.by = "new_celltypes",
        label = TRUE,
        label.size = 6
        ) +
  labs(
    title = paste0(sample_ID, " - new embedding - no cc (new celltypes)"),
    caption = sample_ID
  ) +
    scale_color_manual(values = celltype_cols) +
  NoLegend()
umap_newCelltype
save_if(umap_newCelltype,
           filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID,".new_celltypes.png")),
           width = 8,
           height = 8)

endocardial_markers <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("Ecscr", "Cdh5"),
  ncol = 1,                      # side-by-side panels
  order = TRUE                   # high-expression cells on top
  ) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Endocardial lineage markers"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
endocardial_markers
save_if(endocardial_markers, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".endocardial_genes.png")),
       width = 8,
       height = 8)

epicardial_markers <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("Wt1", "Upk3b"),
  ncol = 1,                      # side-by-side panels
  order = TRUE                   # high-expression cells on top
  ) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Epicardial lineage markers"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
epicardial_markers
save_if(epicardial_markers, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".epicardial_genes.png")),
       width = 8,
       height = 8)

gfp_feature <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("pEGFP-1"),
  ncol = 1,                      # side-by-side panels
  order = TRUE                   # high-expression cells on top
) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " pEGFP-1 Reporter"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
save_if(gfp_feature, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".gfp.png")),
       width = 8,
       height = 8)


vsmc_markers <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("Rgs5", "Cxcl12"),
  ncol = 1,                      # side-by-side panels
  order = TRUE
  ) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " VSMC lineage markers"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
save_if(vsmc_markers, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".vsmc_markers.png")),
       width = 8,
       height = 8)

mesenchymal_markers <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("Postn", "Cthrc1"),
  ncol = 1,                      # side-by-side panels
  order = TRUE
) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Mesenchymal lineage markers"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
save_if(mesenchymal_markers, 
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".mesenchymal_markers.png")),
        width = 8,
        height = 8)

myocardial_markers <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("Myh7", "Myl4"),
  ncol = 1,                      # side-by-side panels
  order = TRUE
) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Mesenchymal lineage markers"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
save_if(myocardial_markers, 
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".myocardial_markers.png")),
        width = 8,
        height = 8)

cellCycle_feature <- FeaturePlot(
  object = re_embed_seurat_obj,
  features = c("G2M.Score", "S.Score"),
  ncol = 1,                      # vertical panels
  order = TRUE                   # high-expression cells on top
) &
  scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Cell Cycle"),
    caption = sample_ID
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
save_if(cellCycle_feature, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.cellCycle.", sample_ID, ".png")),
       width = 8,
       height = 8)

# let's save here and reload in a new session
write_rds(x = re_embed_seurat_obj,
          file = file.path(out_dir, paste0("re_embed.", sample_ID, ".rds")))
```
