---
title: "Subset Seurat object and save"
author: "Melanie Smith"
date: "2026-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes in my full Seurat object and subsets it into 4 independent objects, 1 for each sample in the current analysis.  
- This markdown only splits the Seurat object into smaller objects and saves them, it does not refactor or re-embed anything.  

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)

# Plotting
library(patchwork)
```

# Set project directories

```{r}
# set the prject directory
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
# set the output directory -> this is where the new Seurat objects will be stored.
out_dir <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

# Set sample-level output directories
out_dir_e11_control <- file.path(out_dir, "e11_control")
out_dir_e11_ko      <- file.path(out_dir, "e11_ko")
out_dir_e12_control <- file.path(out_dir, "e12_control")
out_dir_e12_ko      <- file.path(out_dir, "e12_ko")

# Create directories if they don't exist
for (dir in c(
  out_dir_e11_control,
  out_dir_e11_ko,
  out_dir_e12_control,
  out_dir_e12_ko
)) {
  dir.create(dir, recursive = TRUE, showWarnings = FALSE)
}

```

# Load required Seurat object

```{r}
# read in the Seurat object
re_embed_no_cc <- readRDS(file.path(project_repo_dir, "outDir/08-regress-cell-cycle/re_embed_no_cc.rds"))

stopifnot(inherits(re_embed_no_cc, "Seurat"))

Idents(re_embed_no_cc) <- "seurat_clusters"
Idents(re_embed_no_cc) <- factor(Idents(re_embed_no_cc))
DefaultAssay(re_embed_no_cc) <- "RNA"

# Save the existing Seurat clusters
# these clusters were identified when analysing the merged data
re_embed_no_cc$full_dataset_clusters_no_cc <- re_embed_no_cc$seurat_clusters

```

```{r}
DimPlot(re_embed_no_cc,
        reduction = "pca_original",
        dims = c(3:4),
        split.by = "new_celltypes",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
      ggtitle(
        paste0("Full data object - reduction = PCA_original - cc genes removed - re-embed etc - Phase")
      )
```

# Split Seurat object

```{r}
# split the existing Seurat object into a named list based on the sample identifier "orig.ident"
seurat_list <- SplitObject(re_embed_no_cc,
                           split.by = "orig.ident")

e11_control <- seurat_list[["e11_control"]]
e12_control <- seurat_list[["e12_control"]]
e11_ko <- seurat_list[["e11_ko"]]
e12_ko <- seurat_list[["e12_ko"]]

```

# Save new Seurat objects before re-embedding

```{r}
sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

samples <- mget(sample_names)

for (sample in sample_names) {
  write_rds(
    x = samples[[sample]],
    file = file.path(out_dir, sample, paste0(sample, "_no_cc.rds"))
  )
}
```