---
title: "Sample level embeddings and save"
author: "Melanie Smith"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes in my sample level Seurat object, 1 for each sample in the current analysis, then performs refactor and re-embed.  

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)

# Plotting
library(patchwork)
```

# Set project directories

```{r}
# set project directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

# Set sample-level output directories
out_dir_e11_control <- file.path(out_dir, "e11_control")
out_dir_e11_ko      <- file.path(out_dir, "e11_ko")
out_dir_e12_control <- file.path(out_dir, "e12_control")
out_dir_e12_ko      <- file.path(out_dir, "e12_ko")


```

# User defined functions

```{r}
# Helper function (put this at the top of your script)
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object,
           filename = filename,
           width = width,
           height = height,
           dpi = 300)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

## Set up the cell-type colours

```{r}
celltype_cols <- c("Myocardium" = "#E41A1C",
                   "VSMC" = "#984EA3",
                   "Endocardium" = "#377EB8",
                   "Epicardium" = "#FF7F00",
                   "Mesenchyme" = "#4DAF4A",
                   "Other/Unknown" = "grey70")

epicardial_genes <- c("Wt1", "Upk3b", "Apela", "Upk1b", "Aldh1a2")

endocardial_genes <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
```

# Load required Seurat object
- This sample level Seurat object has had the variance partition identified, cell-cycle-associated genes removed but has not been re-embeded since the sample level objects were created.  

```{r}
sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

for (sample in sample_names) {
  assign(
    sample,
    read_rds(
      file = file.path(out_dir, sample, paste0(sample, "_no_cc.rds"))
    )
    )
  }
```

# Re-embed Normalise, find variable features, scale, PCA, UMAP, cluster and visualise
- perform all refactoring for the individual sample level Seurat objects  

```{r}
# ============================================================
# Re-embed cleaned Seurat objects
# ============================================================

sample_names <- c(
  "e11_control",
  "e11_ko",
  "e12_control",
  "e12_ko"
)

# Put the cleaned Seurat objects into a named list
samples <- mget(sample_names)


# ============================================================
# Function to re-embed one sample
# ============================================================

re_embed_sample <- function(seurat_obj,
                            sample_ID,
                            n_dims = 30,
                            nfeatures = 3000,
                            resolution = 0.5) {
  
  message("Processing: ", sample_ID)
  
  # ----------------------------------------------------------
  # Recalculate variable features
  # ----------------------------------------------------------
  
  re_embed_seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures = nfeatures,
    verbose = FALSE
  )
  
  
  # ----------------------------------------------------------
  # Scale data and calculate new PCA
  # ----------------------------------------------------------
  
  re_embed_seurat_obj <- ScaleData(
    re_embed_seurat_obj,
    features = rownames(re_embed_seurat_obj),
    verbose = FALSE
  )
  
  re_embed_seurat_obj <- RunPCA(
    re_embed_seurat_obj,
    reduction.name = "pca_no_cc_sample",
    reduction.key = "PCNoCCSample_",
    verbose = FALSE
  )
  
  
  # ----------------------------------------------------------
  # Calculate new neighbour graph
  # ----------------------------------------------------------
  
  re_embed_seurat_obj <- FindNeighbors(
    re_embed_seurat_obj,
    reduction = "pca_no_cc_sample",
    dims = 1:n_dims,
    verbose = FALSE
  )
  
  
  # ----------------------------------------------------------
  # Calculate new clusters
  # ----------------------------------------------------------
  
  re_embed_seurat_obj <- FindClusters(
    re_embed_seurat_obj,
    resolution = resolution,
    cluster.name = "seurat_clusters_new_sample",
    verbose = FALSE
  )
  
  
  # ----------------------------------------------------------
  # Calculate new UMAP
  # ----------------------------------------------------------
  
  re_embed_seurat_obj <- RunUMAP(
    re_embed_seurat_obj,
    reduction = "pca_no_cc_sample",
    dims = 1:n_dims,
    reduction.name = "umap_no_cc_sample",
    reduction.key = "UMAPNoCCSample_",
    verbose = FALSE
  )
  
  
  message("Finished: ", sample_ID)
  
  return(re_embed_seurat_obj)
  }


# ============================================================
# Run re-embedding on all samples
# ============================================================

re_embed_samples <- lapply(
  names(samples),
  function(sample_ID) {
    
    re_embed_sample(
      seurat_obj = samples[[sample_ID]],
      sample_ID = sample_ID
    )
    
  }
)

# Keep sample names
names(re_embed_samples) <- sample_names

# Quick check to make check reductions
lapply(
  re_embed_samples,
  Reductions
  )

# Check cluster numbers
lapply(
  re_embed_samples,
  function(x) {
    c(
      original_clusters = length(unique(x$seurat_clusters)),
      new_clusters = length(unique(x$seurat_clusters_new_sample))
      )
    }
  )

```

# Original UMAP
- The UMAP embedding generated when the object was still merged and before removing the cell cycle genes.

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "umap_original",
      group.by = "seurat_clusters"
    ) +
      ggtitle(
        paste0(sample_ID, " - original UMAP")
      )
  )
  }
```

# New UMAP coloured by original clusters

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "umap_no_cc_sample",
      group.by = "seurat_clusters"
    ) +
      ggtitle(
        paste0(sample_ID, " - no CC UMAP / original clusters")
      )
  )
  }
```

# New UMAP coloured by new clusters

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "umap_no_cc_sample",
      group.by = "seurat_clusters_new_sample"
    ) +
      ggtitle(
        paste0(sample_ID, " - no CC UMAP / new clusters")
      )
  )
  }
```

# New UMAP coloured by cell types

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "umap_no_cc_sample",
      group.by = "new_celltypes",
      cols = celltype_cols
    ) +
      ggtitle(
        paste0(sample_ID, " - no CC UMAP / cell types")
      )
  )
  }
```

# Original PCA
## Dims 1:2
```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "pca_original",
      group.by = "seurat_clusters",
      dim = c(1:2),
    ) +
      ggtitle(
        paste0(sample_ID, " - original PCA")
      )
  )
  }
```

## Dims 3:4

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  print(
    DimPlot(
      obj,
      reduction = "pca_no_cc_sample",
      group.by = "Phase",
      split.by = "new_celltypes",
      dim = c(3:4),
    ) +
      ggtitle(
        paste0(sample_ID, " - no cc sample level PCA")
      )
  )
  }
```

# UMAPs - save
## New Seurat Cluster

```{r}

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  umap_newCluster <- DimPlot(
      obj,
      reduction = "umap_no_cc_sample",
      group.by = "seurat_clusters_new_sample",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - no cc - new embedding (new clusters)"),
        caption = sample_ID
      ) +
      NoLegend()
    
  save_if(umap_newCluster,
          filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".new_clusters.png")),
          width = 8,
          height = 8)
}

```

## Official celltypes

```{r}
for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  umap_newCelltype <- DimPlot(
      obj,
      reduction = "umap_no_cc_sample",
      group.by = "new_celltypes",
      cols = celltype_cols,
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - no cc - new embedding (cell types)"),
        caption = sample_ID
      ) +
      NoLegend()
    
  save_if(umap_newCelltype,
          filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".new_celltype.png")),
          width = 8,
          height = 8)
  }
```

## Cell type markers

```{r}

######################################################
#   endocardial_markers
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]

  endocardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("Ecscr", "Cdh5"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Endocardial lineage markers"),
    caption = paste0("no cc - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

    save_if(endocardial_markers, 
            filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".endocardial_genes.png")),
            width = 8,
            height = 8)
    }

######################################################
#   epicardial_markers
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  epicardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("Wt1", "Upk3b"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
  ) &
  plot_annotation(
    title = paste0(sample_ID, " Epicardial lineage markers"),
    caption = paste0("no cc - re_embed - sample level")
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

save_if(epicardial_markers, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".epicardial_genes.png")),
       width = 8,
       height = 8)
}

######################################################
#   GFP reporter
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  gfp_feature <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("pEGFP-1"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " pEGFP-1 Reporter"),
    caption = paste0("no cc - re_embed - sample level")
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

save_if(gfp_feature, 
       filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".gfp.png")),
       width = 8,
       height = 8)
}

######################################################
#   VSMC_markers
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]

  vsmc_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("Rgs5", "Cxcl12"),
    ncol = 1,                      # side-by-side panels
    order = TRUE
    ) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) &
  plot_annotation(
    title = paste0(sample_ID, " VSMC lineage markers"),
    caption = paste0("no cc - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  save_if(vsmc_markers, 
          filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".vsmc_markers.png")),
        width = 8,
        height = 8)
}

######################################################
#   mesenchymal_markers
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  mesenchymal_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("Postn", "Cthrc1"),
    ncol = 1,                      # side-by-side panels
    order = TRUE
    ) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) &
    plot_annotation(
      title = paste0(sample_ID, " Mesenchymal lineage markers"),
      caption = paste0("no cc - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
save_if(mesenchymal_markers, 
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".mesenchymal_markers.png")),
        width = 8,
        height = 8)
    }

######################################################
#   myocardial_markers
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  myocardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("Myh7", "Myl4"),
    ncol = 1,                      # side-by-side panels
    order = TRUE
    ) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) &
    plot_annotation(
      title = paste0(sample_ID, " Mesenchymal lineage markers"),
      caption = paste0("no cc - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
save_if(myocardial_markers, 
        filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.", sample_ID, ".myocardial_markers.png")),
        width = 8,
        height = 8)
  }

######################################################
#   cell cycle
######################################################

for (sample_ID in sample_names) {
  
  obj <- re_embed_samples[[sample_ID]]
  
  cellCycle_feature <- FeaturePlot(
    object = obj,
    reduction = "umap_no_cc_sample",
    features = c("G2M.Score", "S.Score"),
    ncol = 1,                      # vertical panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) &
    plot_annotation(
      title = paste0(sample_ID, " Cell Cycle"),
      caption = paste0("no cc - re_embed - sample level")
      ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  save_if(cellCycle_feature, 
          filename = file.path(out_dir, sample_ID, paste0("10-umap.re_embed.cellCycle.", sample_ID, ".png")),
          width = 8,
          height = 8)
    }
```

# Write out the re-embedded object

```{r}

sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

for (sample in sample_names) {
  write_rds(
    x = re_embed_samples[[sample]],
    file = file.path(out_dir, sample, paste0(sample, "_no_cc_re_embed", ".rds"))
  )
}

```
