---
title: "Import Velocyto Loom spliced and unspliced counts and Create Seurat Object"
author: "Melanie Smith"
date: "2026-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

This workflow imports a per-sample Seurat object (containing the standard gene level counts, metadata, and embeddings from typical processing) along with the spliced and unspliced count matrices generated by Velocyto (here from a .loom file).
We then construct a new Seurat object using the spliced/unspliced counts, subset it to retain only high-quality cells and genes (based on the barcodes and gene IDs from the Seurat object), re-compute dimensionality reduction (PCA), and generate a UMAP embedding specifically from the concatenated spliced/unspliced counts.
Finally, we transfer the spliced-based UMAP embeddings (and any related reductions) back into the original sample-level Seurat object, then export the combined object as an .h5ad file. This format is compatible with downstream tools like scvelo/PAGA for RNA velocity-based trajectory inference (e.g., using the velocity-informed graph in PAGA-FDG layout).

# Load Libraries

```{r}
# Python environment constraint — must run before anything touches reticulate/Python
reticulate::py_require(c("anndata==0.10.9", "pandas<3.0"))

# Core tidyverse
library(tidyverse)
# Single cell
library(Seurat)
library(Matrix)
library(data.table)
# Plotting
library(ggokabeito)
library(cowplot)
# Python interop
library(reticulate)
library(sceasy)
library(anndata)
library(scCustomize)
```

# Set project directories and parameters

```{r}
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

# Named vector mapping each sample to its Velocyto loom column prefix
loom_colnames <- c(
  "e11_control" = "possorted_genome_bam_5Q1UF:",
  "e11_ko"      = "possorted_genome_bam_LZRTJ:",
  "e12_control" = "possorted_genome_bam_YIL7N:",
  "e12_ko"      = "possorted_genome_bam_0YX73:"
)

sample_names <- names(loom_colnames)
```

## Set up the cell-type colours

```{r}
celltype_cols <- c("Myocardium"   = "#E41A1C",
                   "VSMC"         = "#984EA3",
                   "Endocardium"  = "#377EB8",
                   "Epicardium"   = "#FF7F00",
                   "Mesenchyme"   = "#4DAF4A",
                   "Other/Unknown" = "grey70")

epicardial_genes  <- c("Wt1", "Upk3b", "Apela", "Upk1b", "Aldh1a2")
endocardial_genes <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
```

# Load the final Seurat objects (one per sample)
- Contains the final set of cells and genes (after removal of cell cycle-associated genes) in a per-sample object
- All counts are filtered and normalised
- These objects are used to supply the cell and gene (feature) IDs to subset the .loom data

```{r}
for (sample in sample_names) {
  assign(
    sample,
    readRDS(file.path(out_dir, sample, paste0(sample, "_no_cc_re_embed.rds")))
  )
}
```

# User defined functions

```{r}
# ============================================================
# Function to import loom counts and re-embed one sample
# ============================================================

process_loom_sample <- function(seurat_obj,
                                sample_ID,
                                loom_colname,
                                n_dims = 30,
                                nfeatures = 3000) {

  message("Processing: ", sample_ID)

  in_dir <- file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID)

  spliced_obj_file   <- file.path(in_dir, "velocyto_output", "spliced_counts.tsv.gz")
  unspliced_obj_file <- file.path(in_dir, "velocyto_output", "unspliced_counts.tsv.gz")

  # Validate input and set defaults
  stopifnot(inherits(seurat_obj, "Seurat"))
  Idents(seurat_obj) <- "seurat_clusters"
  Idents(seurat_obj) <- factor(Idents(seurat_obj))
  DefaultAssay(seurat_obj) <- "RNA"

  # ----------------------------------------------------------
  # Import spliced counts
  # ----------------------------------------------------------

  dt_spliced    <- fread(spliced_obj_file)
  barcodes      <- dt_spliced[[1]]
  clean_barcodes <- gsub(loom_colname, "", barcodes)
  clean_barcodes <- gsub("x$", "", clean_barcodes)
  spliced_mat   <- as.matrix(dt_spliced[, -1])
  storage.mode(spliced_mat) <- "double"
  rownames(spliced_mat) <- clean_barcodes
  spliced_mat   <- Matrix(t(spliced_mat), sparse = TRUE)
  spliced_assay <- CreateAssayObject(counts = spliced_mat)

  # ----------------------------------------------------------
  # Import unspliced counts
  # ----------------------------------------------------------

  dt_unspliced    <- fread(unspliced_obj_file)
  barcodes        <- dt_unspliced[[1]]
  clean_barcodes  <- gsub(loom_colname, "", barcodes)
  clean_barcodes  <- gsub("x$", "", clean_barcodes)
  unspliced_mat   <- as.matrix(dt_unspliced[, -1])
  storage.mode(unspliced_mat) <- "double"
  rownames(unspliced_mat) <- clean_barcodes
  unspliced_mat   <- Matrix(t(unspliced_mat), sparse = TRUE)
  unspliced_assay <- CreateAssayObject(counts = unspliced_mat)

  # ----------------------------------------------------------
  # Detect barcode suffix from the Seurat object
  # ----------------------------------------------------------

  example_bc     <- colnames(spliced_assay)[1]
  matched_bc     <- grep(paste0("^", example_bc), colnames(seurat_obj), value = TRUE)
  stopifnot(length(matched_bc) >= 1)
  barcode_suffix <- sub(paste0("^", example_bc), "", matched_bc[1])
  message("  Detected barcode suffix: '", barcode_suffix, "'")

  # ----------------------------------------------------------
  # Subset to high-quality cells and non-filtered genes
  # ----------------------------------------------------------

  colnames(spliced_assay)   <- paste0(colnames(spliced_assay),   barcode_suffix)
  colnames(unspliced_assay) <- paste0(colnames(unspliced_assay), barcode_suffix)

  target_genes <- rownames(seurat_obj)
  target_cells <- colnames(seurat_obj)

  spliced_counts   <- spliced_assay@counts[target_genes, target_cells]
  unspliced_counts <- unspliced_assay@counts[target_genes, target_cells]

  rm(dt_spliced, spliced_mat, spliced_assay,
     dt_unspliced, unspliced_mat, unspliced_assay)
  gc()

  stopifnot(identical(dim(spliced_counts), dim(unspliced_counts)))

  rownames(spliced_counts)   <- paste0(rownames(spliced_counts),   "-spliced")
  rownames(unspliced_counts) <- paste0(rownames(unspliced_counts), "-unspliced")

  combined_counts <- rbind(spliced_counts, unspliced_counts)

  stopifnot(identical(colnames(combined_counts), colnames(seurat_obj)))

  # ----------------------------------------------------------
  # Build new Seurat object from combined spliced/unspliced counts
  # ----------------------------------------------------------

  new_seurat <- CreateSeuratObject(counts = combined_counts)
  new_seurat <- AddMetaData(
    new_seurat,
    metadata = seurat_obj@meta.data[, c("orig.ident",
                                        "original_barcode",
                                        "genotype",
                                        "stage",
                                        "predicted_sex",
                                        "new_celltypes",
                                        "seurat_clusters_new_sample",
                                        "full_dataset_clusters_no_cc",
                                        "G2M.Score",
                                        "S.Score")]
  )

  # ----------------------------------------------------------
  # Normalise, find variable features, scale, PCA, UMAP
  # ----------------------------------------------------------

  re_embed_new_seurat <- new_seurat %>%
    NormalizeData(
      assay = "RNA",
      normalization.method = "LogNormalize",
      scale.factor = 10000,
      verbose = FALSE
    ) %>%
    FindVariableFeatures(
      assay = "RNA",
      selection.method = "vst",
      nfeatures = nfeatures,
      verbose = FALSE
    ) %>%
    ScaleData(
      assay = "RNA",
      verbose = FALSE
    ) %>%
    RunPCA(
      assay = "RNA",
      verbose = FALSE
    ) %>%
    FindNeighbors(
      reduction = "pca",
      dims = 1:n_dims,
      verbose = FALSE
    ) %>%
    RunUMAP(
      reduction = "pca",
      dims = 1:n_dims,
      verbose = FALSE
    )

  # ----------------------------------------------------------
  # Cluster resolution exploration (printed inline)
  # ----------------------------------------------------------

  for (res in c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6)) {

    clustered <- FindClusters(re_embed_new_seurat, resolution = res, verbose = FALSE)

    p1 <- DimPlot(
      clustered,
      group.by = paste0("RNA_snn_res.", res),
      label = TRUE,
      label.size = 6
    ) +
      labs(
        title = paste0(sample_ID, " - Resolution: ", res),
        caption = "Combined spliced/unspliced"
      )

    p2 <- FeaturePlot(
      clustered,
      features = c("G2M.Score", "S.Score"),
      reduction = "umap",
      blend = FALSE,
      label = TRUE,
      label.size = 6
    ) &
      scale_colour_gradientn(
        colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
        name = "Expression"
      )

    p2 <- p2 + plot_annotation(
      title = paste0(sample_ID, " - Resolution: ", res, " - Cell Cycle Scores"),
      caption = "Combined spliced/unspliced"
    )

    print(p1)
    print(p2)
  }

  # ----------------------------------------------------------
  # Transfer spliced/unspliced UMAP embeddings to original Seurat object
  # ----------------------------------------------------------

  umap_embed <- Embeddings(re_embed_new_seurat, reduction = "umap")
  umap_embed <- umap_embed[colnames(seurat_obj), ]

  seurat_obj[["umap_spliced"]] <- CreateDimReducObject(
    embeddings = umap_embed,
    key = "UMAPspliced_",
    assay = DefaultAssay(seurat_obj)
  )

  message("Finished: ", sample_ID)

  return(seurat_obj)
}
```

# Run loom processing on all samples

```{r}
# Collect loaded Seurat objects into a named list
samples <- mget(sample_names)

# Process all samples
processed_samples <- lapply(
  names(samples),
  function(sample_ID) {
    process_loom_sample(
      seurat_obj   = samples[[sample_ID]],
      sample_ID    = sample_ID,
      loom_colname = loom_colnames[[sample_ID]]
    )
  }
)

# Keep sample names
names(processed_samples) <- sample_names

# Quick check to confirm reductions were added
lapply(processed_samples, Reductions)
```

# Comparison UMAPs — PCA / sample UMAP / spliced UMAP

```{r}
for (sample_ID in sample_names) {

  obj <- processed_samples[[sample_ID]]

  plot_1 <- DimPlot(obj,
                    reduction = "umap_original",
                    group.by = "new_celltypes",
                    cols = celltype_cols,
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = umap_original",
         subtitle = sample_ID,
         caption = "Gene level counts")

  plot_2 <- DimPlot(obj,
                    reduction = "umap_no_cc_sample",
                    group.by = "new_celltypes",
                    cols = celltype_cols,
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = umap_no_cc_sample",
         subtitle = sample_ID,
         caption = "Gene level counts")

  plot_3 <- DimPlot(obj,
                    reduction = "umap_spliced",
                    group.by = "new_celltypes",
                    cols = celltype_cols,
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = UMAP spliced_no_cc_sample",
         subtitle = sample_ID,
         caption = "Combined spliced/unspliced counts")

  prow <- plot_grid(
    plot_1 + theme(legend.position = "none"),
    plot_2 + theme(legend.position = "none"),
    plot_3 + theme(legend.position = "none"),
    align = "vh",
    hjust = -1,
    nrow = 1,
    rel_widths = c(1.5, 1.5, 1.5)
  )

  legend <- get_legend(
    plot_1 +
      guides(color = guide_legend(nrow = 1)) +
      theme(legend.position = "bottom")
  )

  print(
    plot_grid(prow, legend, ncol = 1, rel_heights = c(1, .1))
  )
  }
```

# Write out the processed objects as .h5ad for scVelo

```{r}
for (sample in sample_names) {

  obj <- processed_samples[[sample]]

  Idents(obj) <- "seurat_clusters_new_sample"
  Idents(obj) <- factor(Idents(obj))
  DefaultAssay(obj) <- "RNA"

  # Convert Assay5 → Assay3 (legacy) so sceasy can handle it
  obj[["RNA"]] <- as(obj[["RNA"]], "Assay")
  cat("Assay class:", class(obj[["RNA"]]), "\n")

  as.anndata(obj,
             file_path = file.path(out_dir, sample),
             file_name = paste0(sample, "_spliced_unspliced.h5ad"))
  }
```

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "10.3_session_info.txt"))
```

