---
title: "Subset to Endocardium and Mesenchyme and re-embed"
author: "Melanie Smith"
date: "2026-09-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes in my sample level Seurat objects (post re-embed), subsets each to Endocardium and Mesenchyme cells only, then re-embeds from scratch on the subset.

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)

# Plotting
library(patchwork)
library(RColorBrewer)
```

# Set project directories

```{r}
# set project directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "11-rna_velocity")
in_dir  <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

# Set sample-level output directories
for (sample in sample_names) {
  dir.create(
    file.path(out_dir, sample),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes  <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes  <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# User defined functions

```{r}
# Helper function (put this at the top of your script)
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

# Load required Seurat objects
- These sample level Seurat objects have been re-embedded following removal of the cell-cycle-associated genes (step 10.2).

```{r}
for (sample in sample_names) {
  assign(
    sample,
    read_rds(
      file = file.path(in_dir, sample, paste0(sample, "_no_cc_re_embed.rds"))
    )
    )
  }
```

# Cell type counts (QC before subsetting)
- Check we have enough cells in both the Endocardium and Mesenchyme populations before subsetting.

```{r}
# ============================================================
# Print cell type counts per sample
# ============================================================

for (sample_ID in sample_names) {

  obj <- get(sample_ID)

  message("Cell type counts: ", sample_ID)
  print(table(obj$new_celltypes))
  }
```

# Subset to Endocardium and Mesenchyme

```{r}
# ============================================================
# Subset each sample to Endocardium and Mesenchyme cells only
# ============================================================

subset_samples <- list()

for (sample_ID in sample_names) {

  obj <- get(sample_ID)

  message("Subsetting: ", sample_ID)
  message("Dimensions before subsetting: ", paste(dim(obj), collapse = " x "))

  obj_subset <- subset(
    obj,
    subset = new_celltypes %in% c("Endocardium", "Mesenchyme")
  )

  message("Dimensions after subsetting: ", paste(dim(obj_subset), collapse = " x "))

  subset_samples[[sample_ID]] <- obj_subset
  }
```

# Re-embedding function
- perform re-embedding for the subset (Endocardium and Mesenchyme) Seurat objects

```{r}
# ============================================================
# Function to re-embed one subset sample
# ============================================================

re_embed_subset <- function(seurat_obj,
                             sample_ID,
                             n_dims = 30,
                             nfeatures = 3000,
                             resolution = 0.5) {

  message("Processing: ", sample_ID)

  # ----------------------------------------------------------
  # Set default assay and reset idents
  # ----------------------------------------------------------

  DefaultAssay(seurat_obj) <- "RNA"
  Idents(seurat_obj) <- seurat_obj$new_celltypes

  # ----------------------------------------------------------
  # Recalculate variable features
  # ----------------------------------------------------------

  re_embed_seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures = nfeatures,
    verbose = FALSE
  )


  # ----------------------------------------------------------
  # Scale data and calculate new PCA
  # ----------------------------------------------------------

  re_embed_seurat_obj <- ScaleData(
    re_embed_seurat_obj,
    features = rownames(re_embed_seurat_obj),
    verbose = FALSE
  )

  re_embed_seurat_obj <- RunPCA(
    re_embed_seurat_obj,
    reduction.name = "pca_endo_mesen",
    reduction.key = "PCEndoMesen_",
    verbose = FALSE
  )


  # ----------------------------------------------------------
  # Calculate new neighbour graph
  # ----------------------------------------------------------

  re_embed_seurat_obj <- FindNeighbors(
    re_embed_seurat_obj,
    reduction = "pca_endo_mesen",
    dims = 1:n_dims,
    verbose = FALSE
  )


  # ----------------------------------------------------------
  # Resolution sweep (inspection only, not stored permanently)
  # ----------------------------------------------------------

  for (res in c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6)) {

    sweep_clusters <- FindClusters(
      re_embed_seurat_obj,
      resolution = res,
      verbose = FALSE
    )

    message(
      "Resolution ", res, ": ",
      length(unique(sweep_clusters$seurat_clusters)), " clusters"
    )
    }


  # ----------------------------------------------------------
  # Calculate new clusters at the supplied resolution
  # ----------------------------------------------------------

  re_embed_seurat_obj <- FindClusters(
    re_embed_seurat_obj,
    resolution = resolution,
    cluster.name = "seurat_clusters_endo_mesen",
    verbose = FALSE
  )


  # ----------------------------------------------------------
  # Calculate new UMAP
  # ----------------------------------------------------------

  re_embed_seurat_obj <- RunUMAP(
    re_embed_seurat_obj,
    reduction = "pca_endo_mesen",
    dims = 1:n_dims,
    reduction.name = "umap_endo_mesen",
    reduction.key = "UMAPEndoMesen_",
    verbose = FALSE
  )


  message("Finished: ", sample_ID)

  return(re_embed_seurat_obj)
  }
```

# Run re-embedding on all subset samples

```{r}
# ============================================================
# Run re-embedding on all subset samples
# ============================================================

re_embed_subset_samples <- lapply(
  names(subset_samples),
  function(sample_ID) {

    re_embed_subset(
      seurat_obj = subset_samples[[sample_ID]],
      sample_ID = sample_ID
    )

  }
)

# Keep sample names
names(re_embed_subset_samples) <- sample_names

# Quick check to make check reductions
lapply(
  re_embed_subset_samples,
  Reductions
  )

# Check cluster numbers
lapply(
  re_embed_subset_samples,
  function(x) {
    length(unique(x$seurat_clusters_endo_mesen))
    }
  )
```

# UMAP coloured by cell type

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_celltype <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "new_celltypes",
      cols = celltype_cols,
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (cell types)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_celltype)

  save_if(umap_celltype,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".new_celltype.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by new cluster

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_newCluster <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "seurat_clusters_endo_mesen",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (new clusters)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_newCluster)

  save_if(umap_newCluster,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".new_clusters.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by cell cycle Phase

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_phase <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "Phase",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (Phase)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_phase)

  save_if(umap_phase,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".phase.png")),
          width = 8,
          height = 8)
  }
```

# Endocardial lineage markers

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  endocardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Ecscr", "Cdh5"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Endocardial lineage markers"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(endocardial_markers)

  save_if(endocardial_markers,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".endocardial_genes.png")),
          width = 8,
          height = 8)
  }
```

# Mesenchymal lineage markers

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  mesenchymal_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Postn", "Cthrc1"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Mesenchymal lineage markers"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(mesenchymal_markers)

  save_if(mesenchymal_markers,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".mesenchymal_genes.png")),
          width = 8,
          height = 8)
  }
```

# Cell cycle scores

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  cellCycle_feature <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("G2M.Score", "S.Score"),
    ncol = 1,                      # vertical panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Cell Cycle"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(cellCycle_feature)

  save_if(cellCycle_feature,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".cellCycle.png")),
          width = 8,
          height = 8)
  }
```

# Write out the re-embedded subset objects

```{r}
for (sample in sample_names) {
  write_rds(
    x = re_embed_subset_samples[[sample]],
    file = file.path(out_dir, sample, paste0(sample, "_endo_mesen_no_cc_re_embed", ".rds"))
  )

  message("Saved: ", sample)
  }
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "11.1_session_info.txt"))
```
