---
title: "Subset to Endocardium and Mesenchyme and re-embed"
author: "Melanie Smith"
date: "2026-09-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes in my sample level Seurat objects (post re-embed), subsets each to Endocardium and Mesenchyme cells only, then re-embeds from scratch on the subset.

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)
library(clustree)

# Plotting
library(patchwork)
library(RColorBrewer)
```

# Set project directories

```{r}
# set project directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "11-rna_velocity")
in_dir  <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

# Set sample-level output directories
for (sample in sample_names) {
  dir.create(
    file.path(out_dir, sample),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes  <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes  <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# User defined functions

```{r}
# Helper function (put this at the top of your script)
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

# Load required Seurat objects
- These sample level Seurat objects have been re-embedded following removal of the cell-cycle-associated genes (step 10.2).  
- Seurat objects retain all cell types at this point

```{r}
for (sample in sample_names) {
  assign(
    sample,
    read_rds(
      file = file.path(in_dir, sample, paste0(sample, "_no_cc_re_embed.rds"))
    )
    )
  }
```

# Cell type counts (QC before subsetting)
- Check we have enough cells in both the Endocardium and Mesenchyme populations before subsetting.

```{r}
# ============================================================
# Print cell type counts per sample
# ============================================================

for (sample_ID in sample_names) {

  obj <- get(sample_ID)

  message("Cell type counts: ", sample_ID)
  print(table(obj$new_celltypes))
  }
```

# Subset to Endocardium and Mesenchyme

```{r}
# ============================================================
# Subset each sample to Endocardium and Mesenchyme cells only
# ============================================================

subset_samples <- list()

for (sample_ID in sample_names) {

  obj <- get(sample_ID)

  message("Subsetting: ", sample_ID)
  message("Dimensions before subsetting: ", paste(dim(obj), collapse = " x "))

  obj_subset <- subset(
    obj,
    subset = new_celltypes %in% c("Endocardium", "Mesenchyme")
  )

  message("Dimensions after subsetting: ", paste(dim(obj_subset), collapse = " x "))

  subset_samples[[sample_ID]] <- obj_subset
  }
```

# Re-embedding function
- perform re-embedding for the subset (Endocardium and Mesenchyme) Seurat objects

```{r}
# ============================================================
# Function to re-embed one subset sample
# ============================================================

re_embed_subset <- function(seurat_obj,
                             sample_ID,
                             n_dims = 30,
                             nfeatures = 3000) {

  message("Processing: ", sample_ID)

  # ----------------------------------------------------------
  # Set default assay and reset idents
  # ----------------------------------------------------------
  DefaultAssay(seurat_obj) <- "RNA"
  Idents(seurat_obj) <- seurat_obj$new_celltypes

  # ----------------------------------------------------------
  # Recalculate variable features
  # ----------------------------------------------------------
  re_embed_seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures        = nfeatures,
    verbose          = FALSE
  )

  # ----------------------------------------------------------
  # Scale data and calculate new PCA
  # ----------------------------------------------------------
  re_embed_seurat_obj <- ScaleData(
    re_embed_seurat_obj,
    features = rownames(re_embed_seurat_obj),
    verbose  = FALSE
  )

  re_embed_seurat_obj <- RunPCA(
    re_embed_seurat_obj,
    reduction.name = "pca_endo_mesen",
    reduction.key  = "PCEndoMesen_",
    verbose        = FALSE
  )

  # ----------------------------------------------------------
  # Calculate neighbour graph
  # ----------------------------------------------------------
  re_embed_seurat_obj <- FindNeighbors(
    re_embed_seurat_obj,
    reduction = "pca_endo_mesen",
    dims      = 1:n_dims,
    verbose   = FALSE
  )

  # ----------------------------------------------------------
  # Calculate UMAP
  # ----------------------------------------------------------
  re_embed_seurat_obj <- RunUMAP(
    re_embed_seurat_obj,
    reduction      = "pca_endo_mesen",
    dims           = 1:n_dims,
    reduction.name = "umap_endo_mesen",
    reduction.key  = "UMAPEndoMesen_",
    verbose        = FALSE
  )

  message("Finished: ", sample_ID)
  return(re_embed_seurat_obj)
}
```

# Run re-embedding on all subset samples

```{r}
# ============================================================
# Run re-embedding on all subset samples
# ============================================================

re_embed_subset_samples <- lapply(
  names(subset_samples),
  function(sample_ID) {

    re_embed_subset(
      seurat_obj = subset_samples[[sample_ID]],
      sample_ID = sample_ID
    )

  }
)

# Keep sample names
names(re_embed_subset_samples) <- sample_names

# Quick check to make check reductions
lapply(
  re_embed_subset_samples,
  Reductions
  )
```

# EndMT Module Scoring

## Define EndMT gene sets
- Five gene sets capturing different aspects of endocardial-to-mesenchymal transition (EndMT), each preceded by a comment block giving the gene set name, supporting references, and biological context.

```{r endmt-gene-sets}
# EndMT gene set 1: Endocardial identity (pre-EndMT)
# Marks cells that retain endocardial character before transition begins.
# Key markers: Nfatc1 and Npr3 are the most specific endocardial (non-coronary) markers.
# Ref: Timmerman et al. (2004) J Clin Invest 114:755-763;
#      de Lange et al. (2004) Dev Biol;
#      Luxan et al. (2013) Nat Med.
genes_EndMT_endocardial <- c("Nfatc1", "Npr3", "Cdh5", "Pecam1", "Tie1",
                              "Gata2", "Klf2", "Klf4", "Sox17", "Hey1", "Hey2")

# EndMT gene set 2: EndMT induction / transition state
# Marks cells actively undergoing endocardial-to-mesenchymal transition.
# Has2 (hyaluronan synthase 2) and Versican mark cardiac jelly — highly specific
# to the AV canal and outflow tract at E11.5-E12.5.
# Bmp2, Tgfb2, Tgfb3 are the canonical EndMT-inducing ligands in the AV canal.
# Ref: Timmerman et al. (2004) J Clin Invest 114:755-763;
#      Townsend et al. (2011) Dev Biol 356:208-220;
#      de Vlaming et al. (2012) Differentiation.
genes_EndMT_transition <- c("Snai1", "Snai2", "Twist1", "Twist2", "Zeb1", "Zeb2",
                             "Tbx20", "Has2", "Vcan", "Notch1", "Notch2",
                             "Bmp2", "Tgfb2", "Tgfb3")

# EndMT gene set 3: Valvular mesenchyme (post-EndMT)
# Marks cells that have completed transition and become valvular interstitial
# cell precursors. Most directly validated by Hulin et al. (2019) scRNA-seq
# of mouse valve development at E12.5 — the closest published reference to
# our timepoints.
# Ref: Lincoln et al. (2006) Dev Biol 294:292-302;
#      Levay et al. (2008) Dev Biol 322:189-200;
#      Hulin et al. (2019) Circ Res 124:1328-1342.
genes_EndMT_mesenchyme <- c("Postn", "Col1a1", "Col3a1", "Vim", "Fn1",
                             "Tcf21", "Pdgfra", "Sox9", "Msx1", "Msx2",
                             "Twist1", "Acta2")

# EndMT gene set 4: Notch signaling axis (EndMT-driving)
# Score separately from general Notch targets to distinguish EndMT-driving
# Notch activity from trabeculation/compaction Notch activity.
# Notch1 loss of function abolishes EndMT in the mouse AV canal.
# Ref: Timmerman et al. (2004) J Clin Invest;
#      Luxan et al. (2013) Nat Med 19:1443-1451.
genes_EndMT_Notch <- c("Notch1", "Notch2", "Jag1", "Dll4", "Hey1", "Hey2",
                        "Hes1", "Nrg1", "Erbb2", "Erbb4", "Snai2")

# EndMT gene set 5: TGF-beta / BMP signaling axis
# Eng (endoglin) is a TGF-beta co-receptor expressed specifically on endocardium
# and is a key upstream regulator of AV canal EndMT.
# Ref: Townsend et al. (2011) Dev Biol 356:208-220;
#      Mercado-Pimentel & Runyan (2007) Dev Biol 304:264-275.
genes_EndMT_TGFb <- c("Tgfb2", "Tgfb3", "Bmp2", "Bmp4", "Smad1", "Smad2",
                       "Smad3", "Smad5", "Id1", "Id2", "Alk2", "Alk3",
                       "Alk5", "Eng")
```

## Run module scoring per sample
- For each sample, run `AddModuleScore()` once per EndMT gene set (`nbin = 24`, `ctrl = 5`), storing the resulting `EndMT_*1` metadata columns back into `re_embed_subset_samples`. Each call is wrapped in `tryCatch()` so a scoring failure warns with the sample ID and gene set name rather than halting the loop.

```{r endmt-module-scoring}
gene_sets <- list(
  EndMT_endocardial = genes_EndMT_endocardial,
  EndMT_transition  = genes_EndMT_transition,
  EndMT_mesenchyme  = genes_EndMT_mesenchyme,
  EndMT_Notch       = genes_EndMT_Notch,
  EndMT_TGFb        = genes_EndMT_TGFb
)

for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  for (set_name in names(gene_sets)) {

    obj <- tryCatch({
      AddModuleScore(
        object   = obj,
        features = list(gene_sets[[set_name]]),
        name     = set_name,
        nbin     = 24,
        ctrl     = 5
      )
    }, error = function(e) {
      warning(sample_ID, " — ", set_name, " scoring failed: ", conditionMessage(e))
      obj
    })
  }

  re_embed_subset_samples[[sample_ID]] <- obj
}
```

## Confirm EndMT score columns were added
- Print the metadata column names for one sample to confirm all five `EndMT_*1` score columns are present.

```{r endmt-confirm-columns}
colnames(re_embed_subset_samples[["e11_control"]][[]])
```

# Metabolic State Module Scoring

## Define metabolic state gene sets


```{r metabolic-state-gene-sets}
# Metabolic gene set 1: Glycolysis
# At E11.5-E12.5 cardiomyocytes are primarily glycolytic — high scores are
# expected in immature and proliferating CM populations. Scoring this alongside
# OxPhos allows the glycolytic-to-oxidative metabolic transition to be mapped
# across the endocardial/mesenchymal subset.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_GLYCOLYSIS (official mouse HALLMARK
# collection — mouse gene symbols, no liftover required).
# Ref: Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140;
#      Liberzon et al. (2015) Cell Syst 1:417-425.
genes_Metab_glycolysis <- c("Hk2", "Pkm", "Ldha", "Pgk1", "Eno1",
                             "Pfkl", "Slc2a1", "Aldoa", "Pfkfb3", "Gapdh")

# Metabolic gene set 2: Oxidative phosphorylation
# Expected to be low in immature/transitioning cells at these timepoints and
# higher in more mature or post-EndMT mesenchymal populations. Contrasting
# with glycolysis score gives a metabolic maturation axis across the subset.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_OXIDATIVE_PHOSPHORYLATION.
# Ref: Liberzon et al. (2015) Cell Syst 1:417-425;
#      Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140.
genes_Metab_oxphos <- c("Ndufb3", "Sdhb", "Uqcrc1", "Cox5a", "Atp5f1b",
                         "Cycs", "Sdha", "Uqcrq", "Cox6a1", "Ndufa4")

# Metabolic gene set 3: Hypoxia response
# The embryonic heart is physiologically hypoxic at E11.5-E12.5. Hif1a-driven
# hypoxia signalling is expected in endocardial and transitioning cells.
# Cited2 and Epas1 (HIF2a) are particularly relevant to cardiac development
# and help distinguish cardiac-specific hypoxia response from a generic stress
# signature. High hypoxia score in post-EndMT mesenchyme may indicate cells
# in the avascular cardiac jelly microenvironment.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_HYPOXIA.
# Ref: Liberzon et al. (2015) Cell Syst 1:417-425;
#      Semenza (2012) Science 339:62-66.
genes_Metab_hypoxia <- c("Hif1a", "Vegfa", "Slc2a1", "Pgk1", "Ldha",
                          "Bnip3", "Egln1", "Pdk1", "Cited2", "Epas1")

# Metabolic gene set 4: Fatty acid oxidation
# FAO is the dominant energy source in mature cardiomyocytes but is low at
# embryonic stages. In the endocardial/mesenchymal subset a high FAO score
# would suggest more advanced metabolic maturation. Ppara is the master
# transcriptional regulator of the FAO gene programme in the heart.
# Ref: Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140;
#      Vega et al. (2000) Mol Cell Biol 20:1868-1876.
genes_Metab_FAO <- c("Ppara", "Cpt1b", "Acadm", "Hadha", "Hadhb",
                      "Acadvl", "Acaa2", "Echs1", "Slc25a20", "Fabp3")
```

## Run module scoring per sample
- For each sample, run `AddModuleScore()` once per metabolic state gene set (`nbin = 24`, `ctrl = 5`), storing the resulting `Metab_*1` metadata columns back into `re_embed_subset_samples`. Each call is wrapped in `tryCatch()` so a scoring failure warns with the sample ID and gene set name rather than halting the loop.

```{r metabolic-state-module-scoring}
gene_sets <- list(
  Metab_glycolysis = genes_Metab_glycolysis,
  Metab_oxphos     = genes_Metab_oxphos,
  Metab_hypoxia    = genes_Metab_hypoxia,
  Metab_FAO        = genes_Metab_FAO
)

for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  for (set_name in names(gene_sets)) {

    obj <- tryCatch({
      AddModuleScore(
        object   = obj,
        features = list(gene_sets[[set_name]]),
        name     = set_name,
        nbin     = 24,
        ctrl     = 5
      )
    }, error = function(e) {
      warning(sample_ID, " — ", set_name, " scoring failed: ", conditionMessage(e))
      obj
    })
  }

  re_embed_subset_samples[[sample_ID]] <- obj
}
```

## Confirm metabolic state score columns were added
- Print the metadata column names for one sample to confirm all four `Metab_*1` score columns are present alongside the existing `EndMT_*1` columns.

```{r metabolic-state-confirm-columns}
colnames(re_embed_subset_samples[["e11_control"]][[]])
```

# Resolution sweep
- Run this chunk and inspect the plots for each sample before proceeding.
- Choose a resolution per sample based on the plots below.
- Enter your choices in the **next chunk** before running it.

```{r resolution-sweep}
# ============================================================
# Resolution sweep — store all resolutions in object metadata,
# plot UMAPs for each, then visualise stability with clustree.
# Inspect all plots before running the supervised clustering chunk.
# ============================================================

for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  # ----------------------------------------------------------
  # Run FindClusters at each resolution and store in metadata
  # ----------------------------------------------------------
  for (res in seq(0.1, 1.0, by = 0.1)) {

    col_name <- paste0("res_", res)

    obj <- FindClusters(
      obj,
      resolution   = res,
      cluster.name = col_name,
      verbose      = FALSE
    )

    n_clusters <- length(unique(obj[[col_name]][[1]]))
    message(sample_ID, " — resolution ", res, ": ", n_clusters, " clusters")
  }

  # Store the updated object back into the list so metadata is retained
  re_embed_subset_samples[[sample_ID]] <- obj

  # ----------------------------------------------------------
  # Plot UMAP for each resolution inline
  # ----------------------------------------------------------
  for (res in seq(0.1, 1.0, by = 0.1)) {

    col_name   <- paste0("res_", res)
    n_clusters <- length(unique(obj[[col_name]][[1]]))

    print(
      DimPlot(
        obj,
        reduction  = "umap_endo_mesen",
        group.by   = col_name,
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = paste0(sample_ID, " \u2014 resolution: ", res),
          subtitle = paste0(n_clusters, " clusters"),
          caption  = "Endocardium + Mesenchyme subset"
        ) +
        NoLegend()
    )
  }

  # ----------------------------------------------------------
  # Clustree plot — visualise cluster stability across resolutions
  # ----------------------------------------------------------
  print(
    clustree(
      obj,
      node_colour = "sc3_stability",
      # node_colour_aggr = "mean",
      prefix = "res_"
    ) +
      labs(
        title   = paste0(sample_ID, " \u2014 clustree", " \u2014 sc3_stability"),
        caption = "Endocardium + Mesenchyme subset — node size = cluster size, edge width = cells moving between clusters"
      )
  )
}
```

```{r clustree-overlay}
# ============================================================
# Clustree overlay — resolution sweep plotted on UMAP coordinates
# Reuses the res_0.1 ... res_1 columns stored by the resolution
# sweep chunk above. Inspect these plots before proceeding.
# ============================================================

for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  # Add UMAP coordinates to metadata (required by clustree_overlay)
  umap_coords <- Embeddings(obj, reduction = "umap_endo_mesen")
  colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
  obj <- AddMetaData(obj, metadata = as.data.frame(umap_coords))

  print(
    clustree_overlay(
      obj,
      prefix  = "res_",
      x_value = "UMAP_1",
      y_value = "UMAP_2"
    ) +
      labs(
        title   = paste0(sample_ID, " — clustree overlay on UMAP"),
        caption = "Endocardium + Mesenchyme subset"
      )
  )
}
```

```{r clustree-exploration}

 # ----------------------------------------------------------
 # Clustree plot — visualise signatures
 # ----------------------------------------------------------

sample_ID <- "e11_ko"
sig_genes <- c("Ecscr", "Cdh5", "Postn", "Cthrc1")    

  for (gene in sig_genes) {
    print(
      clustree(
        re_embed_subset_samples[[sample_ID]],
        node_colour = gene,
        node_colour_aggr = "mean",
        prefix = "res_"
        ) +
        labs(
          title   = paste0(sample_ID, " \u2014 clustree", " \u2014", gene, " avg exp."),
          caption = "Endocardium + Mesenchyme subset — node size = cluster size, edge width = cells moving between clusters"
          )
    )
    }


DimPlot(
        re_embed_subset_samples[[sample_ID]],
        reduction  = "umap_endo_mesen",
        group.by   = "new_celltypes",
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = paste0(sample_ID, " \u2014 new_celltypes"),
          caption  = "Endocardium + Mesenchyme subset"
        ) +
        NoLegend()

      signatures <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1", "EndMT_Notch1", "EndMT_TGFb1")

      for (sig in signatures) {
        print(
      FeaturePlot(
        re_embed_subset_samples[[sample_ID]],
        reduction  = "umap_endo_mesen",
        features = sig,
        label      = TRUE,
        label.size = 6
      ) +
        scale_colour_gradientn(
          colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
          name = "Expression"
        ) +
        labs(
          title    = paste0(sample_ID, " \u2014 ", sig),
          caption  = "Endocardium + Mesenchyme subset"
        )
        )
      }

```

# Clustree: EndMT Module Scores
- For each sample, plot a clustree coloured by each EndMT module score (mean per cluster) and export it as a PNG. The resolution prefix is detected automatically from the metadata column names rather than hardcoded, and `clustree()` is passed the metadata data frame directly rather than the Seurat object.

```{r clustree-endmt-scores}
endmt_scores <- c("EndMT_endocardial1", "EndMT_transition1",
                  "EndMT_mesenchyme1", "EndMT_Notch1", "EndMT_TGFb1")

n_saved <- 0

for (sample_ID in sample_names) {

  score_col <- NA

  tryCatch({

    seu     <- re_embed_subset_samples[[sample_ID]]
    meta_df <- seu@meta.data

    # Detect the clustering resolution prefix dynamically from metadata
    # column names (never hardcoded)
    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    dir.create(file.path(out_dir, sample_ID, "clustree"),
               recursive = TRUE, showWarnings = FALSE)

    for (score_col in endmt_scores) {

      score_name <- sub("1$", "", score_col)

      p <- clustree(meta_df,
                    prefix           = resolution_prefix,
                    node_colour      = score_col,
                    node_colour_aggr = "mean",
                    node_size        = "size") +
        ggtitle(paste(sample_ID, score_col))

      ggsave(
        filename = file.path(out_dir, sample_ID, "clustree",
                              paste0(sample_ID, "_", score_name,
                                     "_endo_mesen_no_cc_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )

      n_saved <- n_saved + 1
    }

  }, error = function(e) {
    warning(sample_ID, " — ", score_col, " clustree failed: ", conditionMessage(e))
  })
}

message("Saved ", n_saved, " EndMT clustree PNGs under ", out_dir, " (per-sample 'clustree' subdirectories)")
```

# Clustree: Metabolic Module Scores
- For each sample, plot a clustree coloured by each metabolic module score (mean per cluster) and export it as a PNG. Same structure as the EndMT clustree section above — dynamic resolution prefix detection, `tryCatch()` per sample, and `clustree()` called on the metadata data frame directly.

```{r clustree-metabolic-scores}
metab_scores <- c("Metab_glycolysis1", "Metab_oxphos1",
                  "Metab_hypoxia1", "Metab_FAO1")

n_saved <- 0

for (sample_ID in sample_names) {

  score_col <- NA

  tryCatch({

    seu     <- re_embed_subset_samples[[sample_ID]]
    meta_df <- seu@meta.data

    # Detect the clustering resolution prefix dynamically from metadata
    # column names (never hardcoded)
    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    dir.create(file.path(out_dir, sample_ID, "clustree"),
               recursive = TRUE, showWarnings = FALSE)

    for (score_col in metab_scores) {

      score_name <- sub("1$", "", score_col)

      p <- clustree(meta_df,
                    prefix           = resolution_prefix,
                    node_colour      = score_col,
                    node_colour_aggr = "mean",
                    node_size        = "size") +
        ggtitle(paste(sample_ID, score_col))

      ggsave(
        filename = file.path(out_dir, sample_ID, "clustree",
                              paste0(sample_ID, "_", score_name,
                                     "_endo_mesen_no_cc_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )

      n_saved <- n_saved + 1
    }

  }, error = function(e) {
    warning(sample_ID, " — ", score_col, " clustree failed: ", conditionMessage(e))
  })
}

message("Saved ", n_saved, " metabolic clustree PNGs under ", out_dir, " (per-sample 'clustree' subdirectories)")
```

# Final clustering — supervised resolution choice
- Inspect the resolution sweep plots above before running this chunk.
- Enter your per-sample resolution choices in the `chosen_resolutions` vector below.
- Then run this chunk to apply `FindClusters` at the chosen resolution for each sample.

```{r}
# ============================================================
# Enter your chosen resolution per sample after inspecting
# the sweep plots above. Run this chunk only after reviewing.
# ============================================================

chosen_resolutions <- c(
  "e11_control" = 0.4,
  "e11_ko"      = 0.4,
  "e12_control" = 0.4,
  "e12_ko"      = 0.4
  )

for (sample_ID in sample_names) {

  res <- chosen_resolutions[[sample_ID]]
  message("Applying resolution ", res, " to ", sample_ID)

  re_embed_subset_samples[[sample_ID]] <- FindClusters(
    re_embed_subset_samples[[sample_ID]],
    resolution   = res,
    cluster.name = "seurat_clusters_endo_mesen",
    verbose      = FALSE
  )

  n_clusters <- length(unique(re_embed_subset_samples[[sample_ID]]$seurat_clusters_endo_mesen))
  message("  ", sample_ID, ": ", n_clusters, " clusters at resolution ", res)
  }

```

# UMAP coloured by cell type

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_celltype <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "new_celltypes",
      cols = celltype_cols,
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (cell types)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_celltype)

  save_if(umap_celltype,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".new_celltype.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by new cluster

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_newCluster <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "seurat_clusters_endo_mesen",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (new clusters)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_newCluster)

  save_if(umap_newCluster,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".new_clusters.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by cell cycle Phase

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  umap_phase <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "Phase",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(sample_ID, " - endo/mesen re-embed (Phase)"),
        caption = sample_ID
      ) +
      NoLegend()

  print(umap_phase)

  save_if(umap_phase,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".phase.png")),
          width = 8,
          height = 8)
  }
```

# Endocardial lineage markers

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  endocardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Ecscr", "Cdh5"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Endocardial lineage markers"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(endocardial_markers)

  save_if(endocardial_markers,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".endocardial_genes.png")),
          width = 8,
          height = 8)
  }
```

# Mesenchymal lineage markers

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  mesenchymal_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Postn", "Cthrc1"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Mesenchymal lineage markers"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(mesenchymal_markers)

  save_if(mesenchymal_markers,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".mesenchymal_genes.png")),
          width = 8,
          height = 8)
  }
```

# Cell cycle scores

```{r}
for (sample_ID in sample_names) {

  obj <- re_embed_subset_samples[[sample_ID]]

  cellCycle_feature <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("G2M.Score", "S.Score"),
    ncol = 1,                      # vertical panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(sample_ID, " Cell Cycle"),
    caption = paste0("endo/mesen - re_embed - sample level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(cellCycle_feature)

  save_if(cellCycle_feature,
          filename = file.path(out_dir, sample_ID, paste0("11.1-umap.re_embed.", sample_ID, ".cellCycle.png")),
          width = 8,
          height = 8)
  }
```

# Write out the re-embedded subset objects

```{r}
for (sample in sample_names) {
  write_rds(
    x = re_embed_subset_samples[[sample]],
    file = file.path(out_dir, sample, paste0(sample, "_endo_mesen_no_cc_re_embed", ".rds"))
  )

  message("Saved: ", sample)
  }
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "11.1_session_info.txt"))
```
