---
title: "Import Velocyto spliced/unspliced counts for Endocardium and Mesenchyme subset"
author: "Melanie Smith"
date: "2026-09-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

- This script imports the Velocyto spliced/unspliced counts for the Endocardium/Mesenchyme subset created in 11.1, subsets them to the cells and genes retained in that subset, re-embeds, and exports the combined object as an .h5ad for scVelo/PAGA.

# Load Libraries

```{r}
reticulate::py_require(c("anndata==0.10.9", "pandas<3.0"))

library(tidyverse)
library(Seurat)
library(Matrix)
library(data.table)
library(ggokabeito)
library(cowplot)
library(patchwork)
library(reticulate)
library(sceasy)
library(anndata)
library(scCustomize)
library(clustree)

```

# Set project directories and parameters

```{r}
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "11-rna_velocity")
in_dir  <- file.path(project_repo_dir, "outDir", "10-rna_velocity")

sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

# Named vector mapping each sample to its Velocyto loom column prefix
loom_colnames <- c(
  "e11_control" = "possorted_genome_bam_5Q1UF:",
  "e11_ko"      = "possorted_genome_bam_LZRTJ:",
  "e12_control" = "possorted_genome_bam_YIL7N:",
  "e12_ko"      = "possorted_genome_bam_0YX73:"
)
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# Helper functions

```{r}
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
save_plots <- TRUE
```

# Load Endocardium and Mesenchyme subset Seurat objects
- These are the per-sample objects saved by 11.1
- They contain gene-level counts, metadata, and the `umap_endo_mesen` and `pca_endo_mesen` reductions
- Cell barcodes here define which cells are retained from the loom (spliced/unspliced) data

```{r}
for (sample in sample_names) {
  assign(
    sample,
    read_rds(
      file = file.path(out_dir, sample, paste0(sample, "_endo_mesen_no_cc_re_embed.rds"))
    )
    )
  }

# ------------------------------------------------------------
# Path verified against 11.1's write_rds() call ("Write out the
# re-embedded subset objects" chunk) - the file.path pattern above
# already matches it exactly. The loop above only creates loose
# global variables (e11_control, e11_ko, ...), but tasks further
# down this script expect a named list, matching the structure
# used throughout 11.1: re_embed_subset_samples[[sample_ID]].
# Built here from the same objects (no re-read from disk); the
# loose globals are left in place since the "Run loom processing"
# chunks below still collect them via mget().
# ------------------------------------------------------------
re_embed_subset_samples <- mget(sample_names)
```

# User defined functions

```{r resolution-sweep}
# ============================================================
# Function to import loom counts and re-embed one subset sample
# ============================================================

process_loom_endo_mesen <- function(seurat_obj,
                                     sample_ID,
                                     loom_colname,
                                     n_dims = 30,
                                     nfeatures = 3000) {

  message("Processing: ", sample_ID)

  # ----------------------------------------------------------
  # Validate input and set defaults
  # ----------------------------------------------------------

  stopifnot(inherits(seurat_obj, "Seurat"))
  DefaultAssay(seurat_obj) <- "RNA"
  Idents(seurat_obj) <- seurat_obj$new_celltypes

  # ----------------------------------------------------------
  # Import spliced counts
  # ----------------------------------------------------------

  velocyto_dir     <- file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID, "velocyto_output")
  spliced_obj_file <- file.path(velocyto_dir, "spliced_counts.tsv.gz")

  dt_spliced     <- fread(spliced_obj_file)
  barcodes       <- dt_spliced[[1]]
  clean_barcodes <- gsub(loom_colname, "", barcodes)
  clean_barcodes <- gsub("x$", "", clean_barcodes)
  spliced_mat    <- as.matrix(dt_spliced[, -1])
  storage.mode(spliced_mat) <- "double"
  rownames(spliced_mat) <- clean_barcodes
  spliced_mat    <- Matrix(t(spliced_mat), sparse = TRUE)
  spliced_assay  <- CreateAssayObject(counts = spliced_mat)

  # ----------------------------------------------------------
  # Import unspliced counts
  # ----------------------------------------------------------

  unspliced_obj_file <- file.path(velocyto_dir, "unspliced_counts.tsv.gz")

  dt_unspliced     <- fread(unspliced_obj_file)
  barcodes         <- dt_unspliced[[1]]
  clean_barcodes   <- gsub(loom_colname, "", barcodes)
  clean_barcodes   <- gsub("x$", "", clean_barcodes)
  unspliced_mat    <- as.matrix(dt_unspliced[, -1])
  storage.mode(unspliced_mat) <- "double"
  rownames(unspliced_mat) <- clean_barcodes
  unspliced_mat    <- Matrix(t(unspliced_mat), sparse = TRUE)
  unspliced_assay  <- CreateAssayObject(counts = unspliced_mat)

  # ----------------------------------------------------------
  # Detect barcode suffix from the Seurat object
  # ----------------------------------------------------------

  # seurat_obj here is already subset to Endocardium/Mesenchyme cells
  # only, so the first loom barcode is often not among the retained
  # cells (unlike the full-sample object used in 10.3). Try successive
  # barcodes until one matches, rather than assuming the first one does.
  example_bc <- character(0)
  matched_bc <- character(0)
  for (bc in colnames(spliced_assay)) {
    matched_bc <- grep(paste0("^", bc), colnames(seurat_obj), value = TRUE)
    if (length(matched_bc) >= 1) {
      example_bc <- bc
      break
    }
  }
  stopifnot(length(matched_bc) >= 1)
  barcode_suffix <- sub(paste0("^", example_bc), "", matched_bc[1])
  message("  Detected barcode suffix: '", barcode_suffix, "'")

  # ----------------------------------------------------------
  # Subset to high-quality cells and non-filtered genes
  # ----------------------------------------------------------

  colnames(spliced_assay)   <- paste0(colnames(spliced_assay),   barcode_suffix)
  colnames(unspliced_assay) <- paste0(colnames(unspliced_assay), barcode_suffix)

  target_genes <- rownames(seurat_obj)
  target_cells <- colnames(seurat_obj)

  spliced_counts   <- spliced_assay@counts[target_genes, target_cells]
  unspliced_counts <- unspliced_assay@counts[target_genes, target_cells]

  message("  Retained ", length(target_cells), " cells and ", length(target_genes), " genes after subsetting to Endocardium/Mesenchyme cells")

  rm(dt_spliced, spliced_mat, spliced_assay,
     dt_unspliced, unspliced_mat, unspliced_assay)
  gc()

  stopifnot(identical(dim(spliced_counts), dim(unspliced_counts)))

  rownames(spliced_counts)   <- paste0(rownames(spliced_counts),   "-spliced")
  rownames(unspliced_counts) <- paste0(rownames(unspliced_counts), "-unspliced")

  combined_counts <- rbind(spliced_counts, unspliced_counts)

  stopifnot(identical(colnames(combined_counts), colnames(seurat_obj)))

  # ----------------------------------------------------------
  # Build new Seurat object from combined spliced/unspliced counts
  # ----------------------------------------------------------

  new_seurat <- CreateSeuratObject(counts = combined_counts)
  new_seurat <- AddMetaData(
    new_seurat,
    metadata = seurat_obj@meta.data[, c("orig.ident",
                                        "original_barcode",
                                        "genotype",
                                        "stage",
                                        "predicted_sex",
                                        "new_celltypes",
                                        "seurat_clusters_endo_mesen",
                                        "full_dataset_clusters_no_cc",
                                        "G2M.Score",
                                        "S.Score")]
  )

  # ----------------------------------------------------------
  # Normalise, find variable features, scale, PCA, UMAP
  # ----------------------------------------------------------

  re_embed_new_seurat <- new_seurat %>%
    NormalizeData(
      assay = "RNA",
      normalization.method = "LogNormalize",
      scale.factor = 10000,
      verbose = FALSE
    ) %>%
    FindVariableFeatures(
      assay = "RNA",
      selection.method = "vst",
      nfeatures = nfeatures,
      verbose = FALSE
    ) %>%
    ScaleData(
      assay = "RNA",
      verbose = FALSE
    ) %>%
    RunPCA(
      assay = "RNA",
      verbose = FALSE
    ) %>%
    FindNeighbors(
      reduction = "pca",
      dims = 1:n_dims,
      graph.name = c("spliced_endo_mesen_nn", "spliced_endo_mesen_snn"),
      verbose = FALSE
    ) %>%
    RunUMAP(
      reduction = "pca",
      dims = 1:n_dims,
      verbose = FALSE
    )

  # ----------------------------------------------------------
  # Resolution sweep — inspect before proceeding
  # ----------------------------------------------------------
  # Run clustering once per resolution, storing each result in the
  # Seurat object metadata (Seurat auto-creates one column per
  # resolution, named "<graph.name>_res.<resolution>"), then reuse
  # those columns for both the per-resolution UMAPs and the clustree
  # comparison below - avoids clustering twice at the same resolution.
  #
  # Inspect these plots before running the next chunk.
  # Enter your chosen resolution per sample in the
  # 'chosen_resolutions' vector in the chunk that follows.

  resolution_range <- seq(0.1, 1.0, by = 0.1)

  re_embed_new_seurat <- FindClusters(
    re_embed_new_seurat,
    graph.name = "spliced_endo_mesen_snn",
    resolution = resolution_range,
    verbose    = FALSE
  )

  spliced_resolution_sweep_dir <- file.path(out_dir, sample_ID, "resolution_sweep")
  dir.create(spliced_resolution_sweep_dir, recursive = TRUE, showWarnings = FALSE)

  for (res in resolution_range) {

    col_name   <- paste0("spliced_endo_mesen_snn_res.", res)
    n_clusters <- length(unique(re_embed_new_seurat[[col_name]][[1]]))
    message("  ", sample_ID, " — resolution ", res, ": ", n_clusters, " clusters")

    res_plot <- DimPlot(
        re_embed_new_seurat,
        reduction  = "umap",
        group.by   = col_name,
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = paste0(sample_ID, " — resolution: ", res),
          subtitle = paste0(n_clusters, " clusters"),
          caption  = "Spliced/unspliced — Endocardium + Mesenchyme"
        ) +
        NoLegend()

    print(res_plot)

    ggsave(
      filename = file.path(spliced_resolution_sweep_dir,
                            paste0(sample_ID, "_resolution_", res, "_endo_mesen_spliced_unspliced_re_embed", ".png")),
      plot   = res_plot,
      width  = 8,
      height = 8,
      dpi    = 300
    )
  }

  # ----------------------------------------------------------
  # Resolution sweep — clustree comparison
  # ----------------------------------------------------------
  # Visualise how clusters split/merge across resolutions with
  # clustree(), reusing the resolution columns computed above.
  # Coloured by sc3_stability; saved as a PNG using the same
  # naming/ggsave convention established in 11.1's clustree exports
  # (clustree-endmt-scores / clustree-metabolic-scores).

  clustree_plot <- clustree(
    re_embed_new_seurat,
    prefix      = "spliced_endo_mesen_snn_res.",
    node_colour = "sc3_stability"
  ) +
    labs(
      title   = paste0(sample_ID, " — clustree — sc3_stability"),
      caption = "Spliced/unspliced — Endocardium + Mesenchyme"
    )

  print(clustree_plot)

  spliced_clustree_dir <- file.path(out_dir, sample_ID, "clustree")
  dir.create(spliced_clustree_dir, recursive = TRUE, showWarnings = FALSE)

  ggsave(
    filename = file.path(spliced_clustree_dir,
                          paste0(sample_ID, "_sc3_stability_endo_mesen_spliced_unspliced_re_embed", ".png")),
    plot   = clustree_plot,
    width  = 10,
    height = 12,
    dpi    = 300
  )

  # ----------------------------------------------------------
  # Transfer spliced/unspliced neighbour graph to original
  # Seurat object, so downstream clustering can point at the
  # spliced/unspliced graph rather than the gene-level graph
  # ----------------------------------------------------------

  cell_order <- colnames(seurat_obj)

  nn_mat  <- as(re_embed_new_seurat[["spliced_endo_mesen_nn"]],  "CsparseMatrix")
  snn_mat <- as(re_embed_new_seurat[["spliced_endo_mesen_snn"]], "CsparseMatrix")

  nn_mat  <- nn_mat[cell_order, cell_order]
  snn_mat <- snn_mat[cell_order, cell_order]

  seurat_obj[["spliced_endo_mesen_nn"]]  <- as.Graph(nn_mat)
  seurat_obj[["spliced_endo_mesen_snn"]] <- as.Graph(snn_mat)

  # ----------------------------------------------------------
  # Transfer spliced/unspliced UMAP embeddings to original
  # Seurat object
  # ----------------------------------------------------------

  umap_embed <- Embeddings(re_embed_new_seurat, reduction = "umap")
  umap_embed <- umap_embed[colnames(seurat_obj), ]

  seurat_obj[["umap_spliced_endo_mesen"]] <- CreateDimReducObject(
    embeddings = umap_embed,
    key = "UMAPspliced_",
    assay = DefaultAssay(seurat_obj)
  )

  message("Finished: ", sample_ID)

  return(seurat_obj)
}
```

# Run loom processing — Step 1: embedding and resolution sweep
- Run `process_loom_endo_mesen()` across all samples.
- Inspect the resolution sweep plots in the output before proceeding to the next chunk.

```{r}
# Collect loaded Seurat objects into a named list
samples <- mget(sample_names)

# Process all samples
processed_samples <- lapply(
  names(samples),
  function(sample_ID) {
    process_loom_endo_mesen(
      seurat_obj   = samples[[sample_ID]],
      sample_ID    = sample_ID,
      loom_colname = loom_colnames[[sample_ID]]
    )
  }
)

# Keep sample names
names(processed_samples) <- sample_names

# Quick check to confirm reductions were added
lapply(processed_samples, Reductions)
```

# Run loom processing — Step 2: supervised clustering
- Inspect the resolution sweep plots above before running this chunk.
- Enter the per-sample resolution choices in `chosen_resolutions` below.

```{r}
# ============================================================
# Enter chosen resolutions after inspecting sweep plots above.
# Run this chunk only after reviewing.
# ============================================================

chosen_resolutions <- c(
  "e11_control" = 0.3,
  "e11_ko"      = 0.3,
  "e12_control" = 0.3,
  "e12_ko"      = 0.3
)

for (sample_ID in sample_names) {

  res <- chosen_resolutions[[sample_ID]]
  message("Applying resolution ", res, " to ", sample_ID)

  processed_samples[[sample_ID]][["spliced_clusters_endo_mesen"]] <- FindClusters(
    processed_samples[[sample_ID]],
    graph.name   = "spliced_endo_mesen_snn",
    resolution   = res,
    verbose      = FALSE
  )$seurat_clusters

  n_clusters <- length(unique(processed_samples[[sample_ID]]$spliced_clusters_endo_mesen))
  message("  ", sample_ID, ": ", n_clusters, " clusters at resolution ", res)
}
```

# Comparison UMAPs — gene-level vs spliced/unspliced embeddings

```{r}
for (sample_ID in sample_names) {

  obj <- processed_samples[[sample_ID]]

  plot_1 <- DimPlot(obj,
                    reduction = "umap_endo_mesen",
                    group.by = "new_celltypes",
                    cols = celltype_cols,
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = umap_endo_mesen",
         subtitle = sample_ID,
         caption = "Gene level counts")

  plot_2 <- DimPlot(obj,
                    reduction = "umap_spliced_endo_mesen",
                    group.by = "new_celltypes",
                    cols = celltype_cols,
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = umap_spliced_endo_mesen",
         subtitle = sample_ID,
         caption = "Combined spliced/unspliced counts")

  plot_3 <- DimPlot(obj,
                    reduction = "umap_spliced_endo_mesen",
                    group.by = "spliced_clusters_endo_mesen",
                    label = TRUE,
                    label.size = 6,
                    pt.size = 1) +
    labs(title = "Reduction = umap_spliced_endo_mesen (clusters)",
         subtitle = sample_ID,
         caption = "Combined spliced/unspliced counts")

  prow <- plot_grid(
    plot_1 + theme(legend.position = "none"),
    plot_2 + theme(legend.position = "none"),
    plot_3 + theme(legend.position = "none"),
    align = "vh",
    hjust = -1,
    nrow = 1,
    rel_widths = c(1.5, 1.5, 1.5)
  )

  legend <- get_legend(
    plot_1 +
      guides(color = guide_legend(nrow = 1)) +
      theme(legend.position = "bottom")
  )

  print(
    plot_grid(prow, legend, ncol = 1, rel_heights = c(1, .1))
  )
  }
```

# Write out processed objects as .h5ad for scVelo/PAGA

```{r}
for (sample in sample_names) {

  obj <- processed_samples[[sample]]

  Idents(obj) <- "spliced_clusters_endo_mesen"
  Idents(obj) <- factor(Idents(obj))
  DefaultAssay(obj) <- "RNA"

  obj[["RNA"]] <- as(obj[["RNA"]], "Assay")
  cat("Assay class:", class(obj[["RNA"]]), "\n")

  as.anndata(obj,
             file_path = file.path(out_dir, sample),
             file_name = paste0(sample, "_endo_mesen_spliced_unspliced.h5ad"))
}
```

# Subset Loom Layers to 11.1 Cells and Genes
- Reads each sample's Velocyto `.loom` file directly (not the pre-extracted `spliced_counts.tsv.gz`/`unspliced_counts.tsv.gz` used above) so the `ambiguous` layer is available — `run_spliced_unspliced_export.py` reads all three loom layers via `sc.read_loom()` but only ever writes out `spliced` and `unspliced`, so `ambiguous` was never exported to TSV. Subsets all three layers to the cells and genes retained in the 11.1 Endocardium/Mesenchyme subset.

```{r subset-loom-to-11point1-cells}
loom_spliced_subset   <- list()
loom_unspliced_subset <- list()
loom_ambiguous_subset <- list()

for (sample_ID in sample_names) {

  tryCatch({

    message("Subsetting loom layers: ", sample_ID)

    velocyto_dir <- file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID, "velocyto_output")
    loom_file    <- file.path(velocyto_dir, paste0(sub(":$", "", loom_colnames[[sample_ID]]), ".loom"))

    lfile <- loomR::connect(filename = loom_file, mode = "r", skip.validate = TRUE)

    # loomR returns layers oriented cells (rows) x genes (columns)
    loom_barcodes <- lfile$col.attrs$CellID[]
    loom_genes    <- lfile$row.attrs$Gene[]

    n_cells_original <- length(loom_barcodes)
    n_genes_original <- length(loom_genes)

    # ------------------------------------------------------------
    # Clean loom barcodes: strip this sample's loom prefix and the
    # trailing "x" (same convention already used by
    # process_loom_endo_mesen() above), then detect and append the
    # Seurat barcode suffix by matching one cleaned barcode against
    # the 11.1 subset object's colnames.
    # ------------------------------------------------------------
    clean_barcodes <- gsub(loom_colnames[[sample_ID]], "", loom_barcodes)
    clean_barcodes <- gsub("x$", "", clean_barcodes)

    retained_cells <- colnames(re_embed_subset_samples[[sample_ID]])
    retained_genes <- rownames(re_embed_subset_samples[[sample_ID]])

    # retained_cells is already subset to Endocardium/Mesenchyme cells
    # only, so the first loom barcode is often not among them (same
    # issue fixed in process_loom_endo_mesen() above). Try successive
    # barcodes until one matches, rather than assuming the first one does.
    example_bc <- character(0)
    matched_bc <- character(0)
    for (bc in clean_barcodes) {
      matched_bc <- grep(paste0("^", bc), retained_cells, value = TRUE)
      if (length(matched_bc) >= 1) {
        example_bc <- bc
        break
      }
    }
    stopifnot(length(matched_bc) >= 1)
    barcode_suffix <- sub(paste0("^", example_bc), "", matched_bc[1])

    full_barcodes <- paste0(clean_barcodes, barcode_suffix)

    # ------------------------------------------------------------
    # Match retained (11.1) cells/genes into the loom's cell/gene order
    # ------------------------------------------------------------
    cell_match_idx <- match(retained_cells, full_barcodes)
    pct_matched    <- mean(!is.na(cell_match_idx))

    if (pct_matched < 0.8) {
      warning(sample_ID, " — only ", round(pct_matched * 100, 1),
              "% of 11.1 cells matched in the loom file (expected >= 80%)")
    }

    keep_cells_idx <- cell_match_idx[!is.na(cell_match_idx)]
    keep_cells     <- retained_cells[!is.na(cell_match_idx)]

    gene_match_idx <- match(retained_genes, loom_genes)
    keep_genes_idx <- gene_match_idx[!is.na(gene_match_idx)]
    keep_genes     <- retained_genes[!is.na(gene_match_idx)]

    # ------------------------------------------------------------
    # Read each layer, subset to retained cells/genes, transpose to
    # genes x cells (Seurat convention), wrap as sparse, then discard
    # the dense intermediate before moving to the next layer
    # ------------------------------------------------------------
    read_layer_subset <- function(layer_name) {
      full_mat <- lfile[[paste0("layers/", layer_name)]][, ]
      sub_mat  <- t(full_mat[keep_cells_idx, keep_genes_idx])
      rownames(sub_mat) <- keep_genes
      colnames(sub_mat) <- keep_cells
      sub_mat <- Matrix::Matrix(sub_mat, sparse = TRUE)
      rm(full_mat)
      gc()
      sub_mat
    }

    loom_spliced_subset[[sample_ID]]   <- read_layer_subset("spliced")
    loom_unspliced_subset[[sample_ID]] <- read_layer_subset("unspliced")
    loom_ambiguous_subset[[sample_ID]] <- read_layer_subset("ambiguous")

    lfile$close_all()

    message(
      sample_ID,
      " — cells: ", n_cells_original, " -> ", length(keep_cells),
      " | genes: ", n_genes_original, " -> ", length(keep_genes)
    )

  }, error = function(e) {
    warning(sample_ID, " — loom subsetting failed: ", conditionMessage(e))
  })
}

message("Completed loom layer subsetting for ", length(loom_spliced_subset), " of ", length(sample_names), " samples")
```

# Create Spliced/Unspliced Seurat Objects and Re-embed
- For each sample, build a new Seurat object from the subset spliced counts, attach the unspliced/ambiguous layers as additional assays, transfer all 11.1 metadata (including EndMT/Metab module scores), then re-embed and sweep the same resolution sequence used in 11.1 (`seq(0.1, 1.0, by = 0.1)`, `res_` naming).

```{r create-seurat-spliced-reembed}
resolution_sequence <- seq(0.1, 1.0, by = 0.1)
n_dims_spliced       <- 30
nfeatures_spliced    <- 3000

spliced_subset_samples <- list()

for (sample_ID in sample_names) {

  tryCatch({

    message("Creating spliced/unspliced object: ", sample_ID)

    seu <- CreateSeuratObject(counts = loom_spliced_subset[[sample_ID]])

    seu[["unspliced"]] <- CreateAssayObject(counts = loom_unspliced_subset[[sample_ID]])
    seu[["ambiguous"]] <- CreateAssayObject(counts = loom_ambiguous_subset[[sample_ID]])

    # Transfer all metadata from the 11.1 subset object, matched by barcode
    source_meta <- re_embed_subset_samples[[sample_ID]]@meta.data[colnames(seu), ]
    seu <- AddMetaData(seu, metadata = source_meta)

    DefaultAssay(seu) <- "RNA"

    seu <- NormalizeData(seu, verbose = FALSE)
    seu <- FindVariableFeatures(seu, selection.method = "vst",
                                 nfeatures = nfeatures_spliced, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, reduction.name = "pca_spliced", reduction.key = "PCspliced_", verbose = FALSE)
    seu <- FindNeighbors(seu, reduction = "pca_spliced", dims = 1:n_dims_spliced, verbose = FALSE)

    for (res in resolution_sequence) {
      seu <- FindClusters(seu, resolution = res, cluster.name = paste0("res_", res), verbose = FALSE)
    }

    seu <- RunUMAP(seu, reduction = "pca_spliced", dims = 1:n_dims_spliced,
                   reduction.name = "umap_spliced", reduction.key = "UMAPspliced_", verbose = FALSE)

    spliced_subset_samples[[sample_ID]] <- seu

    message("Finished: ", sample_ID)

  }, error = function(e) {
    warning(sample_ID, " — spliced/unspliced object creation failed: ", conditionMessage(e))
  })
}

message("Created spliced/unspliced objects for ", length(spliced_subset_samples), " of ", length(sample_names), " samples")
```

# Spliced/Unspliced Resolution Sweep Plots
- Reproduces 11.1's `resolution-sweep` figures (per-resolution UMAP + stability clustree) for the spliced/unspliced embedding, using the `res_0.1`...`res_1` columns created above. `clustree()` is passed the metadata data frame directly, with the resolution prefix detected dynamically.

```{r spliced-resolution-sweep-plots}
resolution_sequence <- seq(0.1, 1.0, by = 0.1)

for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    for (res in resolution_sequence) {

      col_name   <- paste0("res_", res)
      n_clusters <- length(unique(seu[[col_name]][[1]]))

      p <- DimPlot(
        seu,
        reduction  = "umap_spliced",
        group.by   = col_name,
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = paste0(sample_ID, " — resolution: ", res),
          subtitle = paste0(n_clusters, " clusters"),
          caption  = "Spliced/unspliced — Endocardium + Mesenchyme subset"
        ) +
        NoLegend()

      print(p)

      ggsave(
        filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".res_", res, ".png")),
        plot = p, width = 8, height = 8, dpi = 300
      )
    }

    meta_df <- seu@meta.data
    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    p_tree <- clustree(
      meta_df,
      prefix      = resolution_prefix,
      node_colour = "sc3_stability"
    ) +
      labs(
        title   = paste0(sample_ID, " — clustree — sc3_stability"),
        caption = "Spliced/unspliced — Endocardium + Mesenchyme subset"
      )

    print(p_tree)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-clustree.spliced_unspliced_re_embed.", sample_ID, ".sc3_stability.png")),
      plot = p_tree, width = 10, height = 12, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced resolution sweep plotting failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced resolution sweep plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced UMAP by Cell Type
- Reproduces 11.1's "UMAP coloured by cell type" figure using the spliced/unspliced embedding.

```{r spliced-umap-celltype}
for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    p <- DimPlot(
        seu,
        reduction = "umap_spliced",
        group.by = "new_celltypes",
        cols = celltype_cols,
        label = TRUE,
        label.size = 6
        ) +
        labs(
          title = paste0(sample_ID, " - spliced/unspliced re-embed (cell types)"),
          caption = sample_ID
        ) +
        NoLegend()

    print(p)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".new_celltype.png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced cell type UMAP failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced UMAP-by-cell-type plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced UMAP by Cell Cycle Phase
- Reproduces 11.1's "UMAP coloured by cell cycle Phase" figure using the spliced/unspliced embedding.

```{r spliced-umap-phase}
for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    p <- DimPlot(
        seu,
        reduction = "umap_spliced",
        group.by = "Phase",
        label = TRUE,
        label.size = 6
        ) +
        labs(
          title = paste0(sample_ID, " - spliced/unspliced re-embed (Phase)"),
          caption = sample_ID
        ) +
        NoLegend()

    print(p)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".phase.png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced Phase UMAP failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced UMAP-by-Phase plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced Endocardial Lineage Markers
- Reproduces 11.1's "Endocardial lineage markers" FeaturePlot using the spliced/unspliced embedding.

```{r spliced-endocardial-markers}
for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    p <- FeaturePlot(
      object = seu,
      reduction = "umap_spliced",
      features = c("Ecscr", "Cdh5"),
      ncol = 1,
      order = TRUE
      ) &
      scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
      ) &
      plot_annotation(
      title = paste0(sample_ID, " Endocardial lineage markers"),
      caption = paste0("spliced/unspliced - re_embed - sample level")
      ) &
      theme_bw(base_size = 14) &
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)
        )

    print(p)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".endocardial_genes.png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced endocardial marker FeaturePlot failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced endocardial marker plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced Mesenchymal Lineage Markers
- Reproduces 11.1's "Mesenchymal lineage markers" FeaturePlot using the spliced/unspliced embedding.

```{r spliced-mesenchymal-markers}
for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    p <- FeaturePlot(
      object = seu,
      reduction = "umap_spliced",
      features = c("Postn", "Cthrc1"),
      ncol = 1,
      order = TRUE
      ) &
      scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
      ) &
      plot_annotation(
      title = paste0(sample_ID, " Mesenchymal lineage markers"),
      caption = paste0("spliced/unspliced - re_embed - sample level")
      ) &
      theme_bw(base_size = 14) &
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)
        )

    print(p)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".mesenchymal_genes.png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced mesenchymal marker FeaturePlot failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced mesenchymal marker plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced Cell Cycle Scores
- Reproduces 11.1's "Cell cycle scores" FeaturePlot using the spliced/unspliced embedding.

```{r spliced-cellcycle-scores}
for (sample_ID in sample_names) {

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    p <- FeaturePlot(
      object = seu,
      reduction = "umap_spliced",
      features = c("G2M.Score", "S.Score"),
      ncol = 1,
      order = TRUE
      ) &
      scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
      ) &
      plot_annotation(
      title = paste0(sample_ID, " Cell Cycle"),
      caption = paste0("spliced/unspliced - re_embed - sample level")
      ) &
      theme_bw(base_size = 14) &
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)
        )

    print(p)

    ggsave(
      filename = file.path(fig_dir, paste0("11.2-umap.spliced_unspliced_re_embed.", sample_ID, ".cellCycle.png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(sample_ID, " — spliced cell cycle FeaturePlot failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced cell cycle score plots for ", length(sample_names), " samples")
```

# Spliced/Unspliced Module Score FeaturePlots
- FeaturePlot for all nine EndMT/Metab module score columns, using the spliced/unspliced embedding.

```{r spliced-module-score-featureplots}
module_scores <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                    "EndMT_Notch1", "EndMT_TGFb1",
                    "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

for (sample_ID in sample_names) {

  score_col <- NA

  tryCatch({

    seu <- spliced_subset_samples[[sample_ID]]

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    for (score_col in module_scores) {

      p <- FeaturePlot(
        seu,
        reduction  = "umap_spliced",
        features   = score_col,
        label      = TRUE,
        label.size = 6
      ) +
        scale_colour_gradientn(
          colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
          name = "Expression"
        ) +
        labs(
          title   = paste0(sample_ID, " — ", score_col),
          caption = "Spliced/unspliced — Endocardium + Mesenchyme subset"
        )

      print(p)

      ggsave(
        filename = file.path(fig_dir, paste0("11.2-featureplot.spliced_unspliced_re_embed.", sample_ID, ".", score_col, ".png")),
        plot = p, width = 8, height = 8, dpi = 300
      )
    }

  }, error = function(e) {
    warning(sample_ID, " — ", score_col, " spliced module score FeaturePlot failed: ", conditionMessage(e))
  })
}

message("Completed spliced/unspliced module score FeaturePlots for ", length(sample_names), " samples")
```

# Spliced/Unspliced Clustree: Module Scores
- Clustree for all nine EndMT/Metab module scores, using the spliced/unspliced embedding. Same approach as 11.1's `clustree-endmt-scores`/`clustree-metabolic-scores`: `clustree()` receives the metadata data frame directly, and the resolution prefix is detected dynamically via `grep()`.

```{r spliced-clustree-module-scores}
module_scores <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                    "EndMT_Notch1", "EndMT_TGFb1",
                    "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

n_saved <- 0

for (sample_ID in sample_names) {

  score_col <- NA

  tryCatch({

    seu     <- spliced_subset_samples[[sample_ID]]
    meta_df <- seu@meta.data

    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    fig_dir <- file.path(out_dir, sample_ID, "spliced_unspliced")
    dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

    for (score_col in module_scores) {

      score_name <- sub("1$", "", score_col)

      p <- clustree(meta_df,
                    prefix           = resolution_prefix,
                    node_colour      = score_col,
                    node_colour_aggr = "mean",
                    node_size        = "size") +
        ggtitle(paste(sample_ID, score_col))

      print(p)

      ggsave(
        filename = file.path(fig_dir,
                              paste0(sample_ID, "_", score_name,
                                     "_endo_mesen_spliced_unspliced_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )

      n_saved <- n_saved + 1
    }

  }, error = function(e) {
    warning(sample_ID, " — ", score_col, " spliced clustree failed: ", conditionMessage(e))
  })
}

message("Saved ", n_saved, " spliced/unspliced module-score clustree PNGs under 'spliced_unspliced' subdirectories")
```

# Export Spliced/Unspliced Objects — RDS
- Save each spliced/unspliced Seurat object to RDS.

```{r export-spliced-seurat-rds}
for (sample_ID in sample_names) {

  tryCatch({

    write_rds(
      x = spliced_subset_samples[[sample_ID]],
      file = file.path(out_dir, sample_ID, paste0(sample_ID, "_spliced_endo_mesen_subset.RDS"))
    )

    message("Saved RDS: ", sample_ID)

  }, error = function(e) {
    warning(sample_ID, " — RDS export failed: ", conditionMessage(e))
  })
}

message("Completed RDS export for ", length(sample_names), " samples")
```

# Export Spliced/Unspliced Objects — .h5ad
- Follows the same `sceasy::as.anndata()` approach already used in "Write out processed objects as .h5ad for scVelo/PAGA" above. Uses `res_0.4` as the cluster identity (matching 11.1's `chosen_resolutions`, applied uniformly to all four samples).

```{r export-spliced-h5ad}
for (sample_ID in sample_names) {

  tryCatch({

    obj <- spliced_subset_samples[[sample_ID]]

    Idents(obj) <- "res_0.4"
    Idents(obj) <- factor(Idents(obj))
    DefaultAssay(obj) <- "RNA"

    obj[["RNA"]] <- as(obj[["RNA"]], "Assay")

    as.anndata(
      obj,
      file_path = file.path(out_dir, sample_ID),
      file_name = paste0(sample_ID, "_spliced_endo_mesen_subset.h5ad")
    )

    message("Saved h5ad: ", sample_ID)

  }, error = function(e) {
    warning(sample_ID, " — h5ad export failed: ", conditionMessage(e))
  })
}

message("Completed h5ad export for ", length(sample_names), " samples")
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "11.2_session_info.txt"))
```
