---
title: "Subset to Endocardium and Mesenchyme by genotype and re-embed"
author: "Melanie Smith"
date: "2026-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reproducibility
```

- This short markdown file takes the full merged Seurat object (all four samples, all cell types, cell-cycle genes retained), splits it by genotype into control (e11_control + e12_control) and knock-out (e11_ko + e12_ko) groups, subsets each to Endocardium and Mesenchyme cells only, then re-embeds from scratch on the subset.
- Cell-cycle-associated genes are deliberately retained in this workflow (unlike `11.x`), so RNA velocity results can be inspected before that removal step.
- Mixing the two timepoints (E11 + E12) within each genotype is intentional, to let a genuine developmental time axis emerge within each group for the RNA velocity analysis.

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)
library(clustree)

# Plotting
library(patchwork)
library(RColorBrewer)
```

# Set project directories

```{r}
# set project directories
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "12-rna_velocity")

# the starting object for this workflow - full merged object, all cell types,
# cell-cycle genes retained (i.e. before the removal step used in 10.x/11.x)
full_object_path <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds")

group_names <- c("control", "ko")

# Maps our short group names onto the actual values in the `genotype` metadata column
genotype_lookup <- c(control = "control", ko = "Mib1_KO")

# Set group-level output directories
for (group_ID in group_names) {
  dir.create(
    file.path(out_dir, group_ID),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes  <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes  <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# User defined functions

```{r}
# Helper function (put this at the top of your script)
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

# Load and split the full Seurat object by genotype
- This is the full merged object (all four samples, all cell types), with cell-cycle-associated genes retained.
- Splitting on `genotype` combines both timepoints within each arm: `"control"` = e11_control + e12_control, `"Mib1_KO"` = e11_ko + e12_ko (confirmed directly against the per-sample cell counts in `orig.ident`).

```{r}
full_object <- read_rds(full_object_path)

stopifnot(inherits(full_object, "Seurat"))

split_by_genotype <- SplitObject(full_object, split.by = "genotype")

subset_groups <- list()

for (group_ID in group_names) {
  subset_groups[[group_ID]] <- split_by_genotype[[genotype_lookup[[group_ID]]]]
}

rm(full_object, split_by_genotype)
gc()
```

# Cell type counts (QC before subsetting)
- Check we have enough cells in both the Endocardium and Mesenchyme populations before subsetting.

```{r}
# ============================================================
# Print cell type counts per group
# ============================================================

for (group_ID in group_names) {

  obj <- subset_groups[[group_ID]]

  message("Cell type counts: ", group_ID)
  print(table(obj$new_celltypes))
  }
```

# Subset to Endocardium and Mesenchyme

```{r}
# ============================================================
# Subset each genotype group to Endocardium and Mesenchyme cells only
# ============================================================

for (group_ID in group_names) {

  obj <- subset_groups[[group_ID]]

  message("Subsetting: ", group_ID)
  message("Dimensions before subsetting: ", paste(dim(obj), collapse = " x "))

  obj_subset <- subset(
    obj,
    subset = new_celltypes %in% c("Endocardium", "Mesenchyme")
  )

  message("Dimensions after subsetting: ", paste(dim(obj_subset), collapse = " x "))

  subset_groups[[group_ID]] <- obj_subset
  }
```

# Re-embedding function
- perform re-embedding for the subset (Endocardium and Mesenchyme) Seurat objects

```{r}
# ============================================================
# Function to re-embed one subset group
# ============================================================

re_embed_subset <- function(seurat_obj,
                             group_ID,
                             n_dims = 30,
                             nfeatures = 3000) {

  message("Processing: ", group_ID)

  # ----------------------------------------------------------
  # Set default assay and reset idents
  # ----------------------------------------------------------
  DefaultAssay(seurat_obj) <- "RNA"
  Idents(seurat_obj) <- seurat_obj$new_celltypes

  # ----------------------------------------------------------
  # Recalculate variable features
  # ----------------------------------------------------------
  re_embed_seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures        = nfeatures,
    verbose          = FALSE
  )

  # ----------------------------------------------------------
  # Scale data and calculate new PCA
  # ----------------------------------------------------------
  re_embed_seurat_obj <- ScaleData(
    re_embed_seurat_obj,
    features = rownames(re_embed_seurat_obj),
    verbose  = FALSE
  )

  re_embed_seurat_obj <- RunPCA(
    re_embed_seurat_obj,
    reduction.name = "pca_endo_mesen",
    reduction.key  = "PCEndoMesen_",
    verbose        = FALSE
  )

  # ----------------------------------------------------------
  # Calculate neighbour graph
  # ----------------------------------------------------------
  re_embed_seurat_obj <- FindNeighbors(
    re_embed_seurat_obj,
    reduction = "pca_endo_mesen",
    dims      = 1:n_dims,
    verbose   = FALSE
  )

  # ----------------------------------------------------------
  # Calculate UMAP
  # ----------------------------------------------------------
  re_embed_seurat_obj <- RunUMAP(
    re_embed_seurat_obj,
    reduction      = "pca_endo_mesen",
    dims           = 1:n_dims,
    reduction.name = "umap_endo_mesen",
    reduction.key  = "UMAPEndoMesen_",
    verbose        = FALSE
  )

  message("Finished: ", group_ID)
  return(re_embed_seurat_obj)
}
```

# Run re-embedding on all subset groups

```{r}
# ============================================================
# Run re-embedding on all subset groups
# ============================================================

re_embed_subset_groups <- lapply(
  names(subset_groups),
  function(group_ID) {

    re_embed_subset(
      seurat_obj = subset_groups[[group_ID]],
      group_ID = group_ID
    )

  }
)

# Keep group names
names(re_embed_subset_groups) <- group_names

# Quick check to make check reductions
lapply(
  re_embed_subset_groups,
  Reductions
  )
```

# EndMT Module Scoring

## Define EndMT gene sets
- Five gene sets capturing different aspects of endocardial-to-mesenchymal transition (EndMT), each preceded by a comment block giving the gene set name, supporting references, and biological context.

```{r endmt-gene-sets}
# EndMT gene set 1: Endocardial identity (pre-EndMT)
# Marks cells that retain endocardial character before transition begins.
# Key markers: Nfatc1 and Npr3 are the most specific endocardial (non-coronary) markers.
# Ref: Timmerman et al. (2004) J Clin Invest 114:755-763;
#      de Lange et al. (2004) Dev Biol;
#      Luxan et al. (2013) Nat Med.
genes_EndMT_endocardial <- c("Nfatc1", "Npr3", "Cdh5", "Pecam1", "Tie1",
                              "Gata2", "Klf2", "Klf4", "Sox17", "Hey1", "Hey2")

# EndMT gene set 2: EndMT induction / transition state
# Marks cells actively undergoing endocardial-to-mesenchymal transition.
# Has2 (hyaluronan synthase 2) and Versican mark cardiac jelly — highly specific
# to the AV canal and outflow tract at E11.5-E12.5.
# Bmp2, Tgfb2, Tgfb3 are the canonical EndMT-inducing ligands in the AV canal.
# Ref: Timmerman et al. (2004) J Clin Invest 114:755-763;
#      Townsend et al. (2011) Dev Biol 356:208-220;
#      de Vlaming et al. (2012) Differentiation.
genes_EndMT_transition <- c("Snai1", "Snai2", "Twist1", "Twist2", "Zeb1", "Zeb2",
                             "Tbx20", "Has2", "Vcan", "Notch1", "Notch2",
                             "Bmp2", "Tgfb2", "Tgfb3")

# EndMT gene set 3: Valvular mesenchyme (post-EndMT)
# Marks cells that have completed transition and become valvular interstitial
# cell precursors. Most directly validated by Hulin et al. (2019) scRNA-seq
# of mouse valve development at E12.5 — the closest published reference to
# our timepoints.
# Ref: Lincoln et al. (2006) Dev Biol 294:292-302;
#      Levay et al. (2008) Dev Biol 322:189-200;
#      Hulin et al. (2019) Circ Res 124:1328-1342.
genes_EndMT_mesenchyme <- c("Postn", "Col1a1", "Col3a1", "Vim", "Fn1",
                             "Tcf21", "Pdgfra", "Sox9", "Msx1", "Msx2",
                             "Twist1", "Acta2")

# EndMT gene set 4: Notch signaling axis (EndMT-driving)
# Score separately from general Notch targets to distinguish EndMT-driving
# Notch activity from trabeculation/compaction Notch activity.
# Notch1 loss of function abolishes EndMT in the mouse AV canal.
# Ref: Timmerman et al. (2004) J Clin Invest;
#      Luxan et al. (2013) Nat Med 19:1443-1451.
genes_EndMT_Notch <- c("Notch1", "Notch2", "Jag1", "Dll4", "Hey1", "Hey2",
                        "Hes1", "Nrg1", "Erbb2", "Erbb4", "Snai2")

# EndMT gene set 5: TGF-beta / BMP signaling axis
# Eng (endoglin) is a TGF-beta co-receptor expressed specifically on endocardium
# and is a key upstream regulator of AV canal EndMT.
# Ref: Townsend et al. (2011) Dev Biol 356:208-220;
#      Mercado-Pimentel & Runyan (2007) Dev Biol 304:264-275.
genes_EndMT_TGFb <- c("Tgfb2", "Tgfb3", "Bmp2", "Bmp4", "Smad1", "Smad2",
                       "Smad3", "Smad5", "Id1", "Id2", "Alk2", "Alk3",
                       "Alk5", "Eng")
```

## Run module scoring per group
- For each group, run `AddModuleScore()` once per EndMT gene set (`nbin = 24`, `ctrl = 5`), storing the resulting `EndMT_*1` metadata columns back into `re_embed_subset_groups`. Each call is wrapped in `tryCatch()` so a scoring failure warns with the group ID and gene set name rather than halting the loop.

```{r endmt-module-scoring}
gene_sets <- list(
  EndMT_endocardial = genes_EndMT_endocardial,
  EndMT_transition  = genes_EndMT_transition,
  EndMT_mesenchyme  = genes_EndMT_mesenchyme,
  EndMT_Notch       = genes_EndMT_Notch,
  EndMT_TGFb        = genes_EndMT_TGFb
)

for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  for (set_name in names(gene_sets)) {

    obj <- tryCatch({
      AddModuleScore(
        object   = obj,
        features = list(gene_sets[[set_name]]),
        name     = set_name,
        nbin     = 24,
        ctrl     = 5
      )
    }, error = function(e) {
      warning(group_ID, " — ", set_name, " scoring failed: ", conditionMessage(e))
      obj
    })
  }

  re_embed_subset_groups[[group_ID]] <- obj
}
```

## Confirm EndMT score columns were added
- Print the metadata column names for one group to confirm all five `EndMT_*1` score columns are present.

```{r endmt-confirm-columns}
colnames(re_embed_subset_groups[["control"]][[]])
```

# Metabolic State Module Scoring

## Define metabolic state gene sets


```{r metabolic-state-gene-sets}
# Metabolic gene set 1: Glycolysis
# At E11.5-E12.5 cardiomyocytes are primarily glycolytic — high scores are
# expected in immature and proliferating CM populations. Scoring this alongside
# OxPhos allows the glycolytic-to-oxidative metabolic transition to be mapped
# across the endocardial/mesenchymal subset.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_GLYCOLYSIS (official mouse HALLMARK
# collection — mouse gene symbols, no liftover required).
# Ref: Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140;
#      Liberzon et al. (2015) Cell Syst 1:417-425.
genes_Metab_glycolysis <- c("Hk2", "Pkm", "Ldha", "Pgk1", "Eno1",
                             "Pfkl", "Slc2a1", "Aldoa", "Pfkfb3", "Gapdh")

# Metabolic gene set 2: Oxidative phosphorylation
# Expected to be low in immature/transitioning cells at these timepoints and
# higher in more mature or post-EndMT mesenchymal populations. Contrasting
# with glycolysis score gives a metabolic maturation axis across the subset.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_OXIDATIVE_PHOSPHORYLATION.
# Ref: Liberzon et al. (2015) Cell Syst 1:417-425;
#      Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140.
genes_Metab_oxphos <- c("Ndufb3", "Sdhb", "Uqcrc1", "Cox5a", "Atp5f1b",
                         "Cycs", "Sdha", "Uqcrq", "Cox6a1", "Ndufa4")

# Metabolic gene set 3: Hypoxia response
# The embryonic heart is physiologically hypoxic at E11.5-E12.5. Hif1a-driven
# hypoxia signalling is expected in endocardial and transitioning cells.
# Cited2 and Epas1 (HIF2a) are particularly relevant to cardiac development
# and help distinguish cardiac-specific hypoxia response from a generic stress
# signature. High hypoxia score in post-EndMT mesenchyme may indicate cells
# in the avascular cardiac jelly microenvironment.
# Source: MSigDB mh.all.v2023.2.Mm HALLMARK_HYPOXIA.
# Ref: Liberzon et al. (2015) Cell Syst 1:417-425;
#      Semenza (2012) Science 339:62-66.
genes_Metab_hypoxia <- c("Hif1a", "Vegfa", "Slc2a1", "Pgk1", "Ldha",
                          "Bnip3", "Egln1", "Pdk1", "Cited2", "Epas1")

# Metabolic gene set 4: Fatty acid oxidation
# FAO is the dominant energy source in mature cardiomyocytes but is low at
# embryonic stages. In the endocardial/mesenchymal subset a high FAO score
# would suggest more advanced metabolic maturation. Ppara is the master
# transcriptional regulator of the FAO gene programme in the heart.
# Ref: Lopaschuk & Jaswal (2010) J Cardiovasc Pharmacol 56:130-140;
#      Vega et al. (2000) Mol Cell Biol 20:1868-1876.
genes_Metab_FAO <- c("Ppara", "Cpt1b", "Acadm", "Hadha", "Hadhb",
                      "Acadvl", "Acaa2", "Echs1", "Slc25a20", "Fabp3")
```

## Run module scoring per group
- For each group, run `AddModuleScore()` once per metabolic state gene set (`nbin = 24`, `ctrl = 5`), storing the resulting `Metab_*1` metadata columns back into `re_embed_subset_groups`. Each call is wrapped in `tryCatch()` so a scoring failure warns with the group ID and gene set name rather than halting the loop.

```{r metabolic-state-module-scoring}
gene_sets <- list(
  Metab_glycolysis = genes_Metab_glycolysis,
  Metab_oxphos     = genes_Metab_oxphos,
  Metab_hypoxia    = genes_Metab_hypoxia,
  Metab_FAO        = genes_Metab_FAO
)

for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  for (set_name in names(gene_sets)) {

    obj <- tryCatch({
      AddModuleScore(
        object   = obj,
        features = list(gene_sets[[set_name]]),
        name     = set_name,
        nbin     = 24,
        ctrl     = 5
      )
    }, error = function(e) {
      warning(group_ID, " — ", set_name, " scoring failed: ", conditionMessage(e))
      obj
    })
  }

  re_embed_subset_groups[[group_ID]] <- obj
}
```

## Confirm metabolic state score columns were added
- Print the metadata column names for one group to confirm all four `Metab_*1` score columns are present alongside the existing `EndMT_*1` columns.

```{r metabolic-state-confirm-columns}
colnames(re_embed_subset_groups[["control"]][[]])
```

# UMAP by EndMT and Metabolic Signature
- For each group, plot a UMAP coloured by each EndMT and metabolic module score, and export it as a PNG under a `gene_level/umap_signatures` subfolder (sibling to `gene_level/clustree`).

```{r umap-signatures}
signatures <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                "EndMT_Notch1", "EndMT_TGFb1",
                "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

for (group_ID in group_names) {

  sig <- NA

  tryCatch({

    umap_signatures_dir <- file.path(out_dir, group_ID, "gene_level", "umap_signatures")
    dir.create(umap_signatures_dir, recursive = TRUE, showWarnings = FALSE)

    for (sig in signatures) {

      sig_name <- sub("1$", "", sig)

      p <- FeaturePlot(
        re_embed_subset_groups[[group_ID]],
        reduction  = "umap_endo_mesen",
        features = sig,
        label      = TRUE,
        label.size = 6
      ) +
        scale_colour_gradientn(
          colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
          name = "Expression"
        ) +
        labs(
          title    = paste0(group_ID, " — ", sig),
          caption  = "Endocardium + Mesenchyme subset — gene level"
        )

      print(p)

      ggsave(
        filename = file.path(umap_signatures_dir,
                              paste0(group_ID, "_", sig_name, "_endo_mesen_re_embed", ".png")),
        plot = p, width = 8, height = 8, dpi = 300
      )
    }

  }, error = function(e) {
    warning(group_ID, " — ", sig, " UMAP signature plot failed: ", conditionMessage(e))
  })
}
```

# Resolution sweep
- Run this chunk and inspect the plots for each group before proceeding.
- Choose a resolution per group based on the plots below.
- Enter your choices in the **next chunk** before running it.

```{r resolution-sweep}
# ============================================================
# Resolution sweep — store all resolutions in object metadata,
# plot UMAPs for each, then visualise stability with clustree.
# Inspect all plots before running the supervised clustering chunk.
# Per-resolution UMAPs are saved under a 'gene_level/resolution_sweep'
# subfolder, the stability clustree under 'gene_level/clustree' (sibling
# folders, same convention as the EndMT/metabolic clustree exports above),
# and a summary table of cluster counts per resolution is saved as a CSV.
# ============================================================

resolution_summary <- data.frame()

for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  resolution_sweep_dir <- file.path(out_dir, group_ID, "gene_level", "resolution_sweep")
  dir.create(resolution_sweep_dir, recursive = TRUE, showWarnings = FALSE)

  clustree_dir <- file.path(out_dir, group_ID, "gene_level", "clustree")
  dir.create(clustree_dir, recursive = TRUE, showWarnings = FALSE)

  # ----------------------------------------------------------
  # Run FindClusters at each resolution and store in metadata
  # ----------------------------------------------------------
  for (res in seq(0.1, 1.0, by = 0.1)) {

    col_name <- paste0("res_", res)

    obj <- FindClusters(
      obj,
      resolution   = res,
      cluster.name = col_name,
      verbose      = FALSE
    )

    n_clusters <- length(unique(obj[[col_name]][[1]]))
    message(group_ID, " — resolution ", res, ": ", n_clusters, " clusters")

    resolution_summary <- rbind(
      resolution_summary,
      data.frame(group_ID = group_ID, resolution = res, n_clusters = n_clusters)
    )
  }

  # Store the updated object back into the list so metadata is retained
  re_embed_subset_groups[[group_ID]] <- obj

  # ----------------------------------------------------------
  # Plot UMAP for each resolution inline, and save to disk
  # ----------------------------------------------------------
  for (res in seq(0.1, 1.0, by = 0.1)) {

    col_name   <- paste0("res_", res)
    n_clusters <- length(unique(obj[[col_name]][[1]]))

    res_plot <- DimPlot(
        obj,
        reduction  = "umap_endo_mesen",
        group.by   = col_name,
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = paste0(group_ID, " — resolution: ", res),
          subtitle = paste0(n_clusters, " clusters"),
          caption  = "Endocardium + Mesenchyme subset — gene level"
        ) +
        NoLegend()

    print(res_plot)

    ggsave(
      filename = file.path(resolution_sweep_dir,
                            paste0(group_ID, "_resolution_", res, "_endo_mesen_re_embed", ".png")),
      plot   = res_plot,
      width  = 8,
      height = 8,
      dpi    = 300
    )
  }

  # ----------------------------------------------------------
  # Clustree plot — visualise cluster stability across resolutions
  # ----------------------------------------------------------
  stability_plot <- clustree(
      obj,
      node_colour = "sc3_stability",
      # node_colour_aggr = "mean",
      prefix = "res_"
    ) +
      labs(
        title   = paste0(group_ID, " — clustree", " — sc3_stability"),
        caption = "Endocardium + Mesenchyme subset — gene level — node size = cluster size, edge width = cells moving between clusters"
      )

  print(stability_plot)

  ggsave(
    filename = file.path(clustree_dir,
                          paste0(group_ID, "_sc3_stability_endo_mesen_re_embed", ".png")),
    plot   = stability_plot,
    width  = 10,
    height = 12,
    dpi    = 300
  )
}

# ----------------------------------------------------------
# Save the resolution sweep summary table
# ----------------------------------------------------------
print(resolution_summary)

write.csv(
  resolution_summary,
  file.path(out_dir, "gene_level_resolution_sweep_summary.csv"),
  row.names = FALSE
)
```

```{r clustree-overlay}
# ============================================================
# Clustree overlay — resolution sweep plotted on UMAP coordinates
# Reuses the res_0.1 ... res_1 columns stored by the resolution
# sweep chunk above. Inspect these plots before proceeding.
# ============================================================

for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  # Add UMAP coordinates to metadata (required by clustree_overlay)
  umap_coords <- Embeddings(obj, reduction = "umap_endo_mesen")
  colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
  obj <- AddMetaData(obj, metadata = as.data.frame(umap_coords))

  print(
    clustree_overlay(
      obj,
      prefix  = "res_",
      x_value = "UMAP_1",
      y_value = "UMAP_2"
    ) +
      labs(
        title   = paste0(group_ID, " — clustree overlay on UMAP"),
        caption = "Endocardium + Mesenchyme subset — gene level"
      )
  )
}
```

```{r clustree-exploration}

 # ----------------------------------------------------------
 # Clustree plot — visualise signatures
 # ----------------------------------------------------------

sig_genes <- c("Ecscr", "Cdh5", "Postn", "Cthrc1")

for (group_ID in group_names) {

  gene <- NA

  tryCatch({

    clustree_dir <- file.path(out_dir, group_ID, "gene_level", "clustree")
    dir.create(clustree_dir, recursive = TRUE, showWarnings = FALSE)

    for (gene in sig_genes) {

      p <- clustree(
        re_embed_subset_groups[[group_ID]],
        node_colour = gene,
        node_colour_aggr = "mean",
        prefix = "res_"
        ) +
        labs(
          title   = paste0(group_ID, " — clustree", " —", gene, " avg exp."),
          caption = "Endocardium + Mesenchyme subset — gene level"
          )

      print(p)

      ggsave(
        filename = file.path(clustree_dir,
                              paste0(group_ID, "_", gene, "_endo_mesen_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )
    }

  }, error = function(e) {
    warning(group_ID, " — ", gene, " clustree exploration failed: ", conditionMessage(e))
  })
}

```

# UMAP by Cell Type (pre-final-clustering)
- The `new_celltypes` UMAP originally generated inside `clustree-exploration`, moved into its own chunk, looped across both groups, and saved to `gene_level/umap_signatures`.

```{r umap-new-celltypes-exploration}
for (group_ID in group_names) {

  tryCatch({

    umap_signatures_dir <- file.path(out_dir, group_ID, "gene_level", "umap_signatures")
    dir.create(umap_signatures_dir, recursive = TRUE, showWarnings = FALSE)

    p <- DimPlot(
            re_embed_subset_groups[[group_ID]],
            reduction  = "umap_endo_mesen",
            group.by   = "new_celltypes",
            label      = TRUE,
            label.size = 6
          ) +
            labs(
              title    = paste0(group_ID, " — new_celltypes"),
              caption  = "Endocardium + Mesenchyme subset — gene level"
            ) +
            NoLegend()

    print(p)

    ggsave(
      filename = file.path(umap_signatures_dir,
                            paste0(group_ID, "_new_celltypes_endo_mesen_re_embed", ".png")),
      plot = p, width = 8, height = 8, dpi = 300
    )

  }, error = function(e) {
    warning(group_ID, " — new_celltypes UMAP failed: ", conditionMessage(e))
  })
}
```

# Clustree: EndMT Module Scores
- For each group, plot a clustree coloured by each EndMT module score (mean per cluster) and export it as a PNG. The resolution prefix is detected automatically from the metadata column names rather than hardcoded, and `clustree()` is passed the metadata data frame directly rather than the Seurat object.

```{r clustree-endmt-scores}
endmt_scores <- c("EndMT_endocardial1", "EndMT_transition1",
                  "EndMT_mesenchyme1", "EndMT_Notch1", "EndMT_TGFb1")

n_saved <- 0

for (group_ID in group_names) {

  score_col <- NA

  tryCatch({

    seu     <- re_embed_subset_groups[[group_ID]]
    meta_df <- seu@meta.data

    # Detect the clustering resolution prefix dynamically from metadata
    # column names (never hardcoded)
    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    dir.create(file.path(out_dir, group_ID, "gene_level", "clustree"),
               recursive = TRUE, showWarnings = FALSE)

    for (score_col in endmt_scores) {

      score_name <- sub("1$", "", score_col)

      p <- clustree(meta_df,
                    prefix           = resolution_prefix,
                    node_colour      = score_col,
                    node_colour_aggr = "mean",
                    node_size        = "size") +
        ggtitle(paste(group_ID, score_col))

      ggsave(
        filename = file.path(out_dir, group_ID, "gene_level", "clustree",
                              paste0(group_ID, "_", score_name,
                                     "_endo_mesen_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )

      n_saved <- n_saved + 1
    }

  }, error = function(e) {
    warning(group_ID, " — ", score_col, " clustree failed: ", conditionMessage(e))
  })
}

message("Saved ", n_saved, " EndMT clustree PNGs under ", out_dir, " (per-group 'clustree' subdirectories)")
```

# Clustree: Metabolic Module Scores
- For each group, plot a clustree coloured by each metabolic module score (mean per cluster) and export it as a PNG. Same structure as the EndMT clustree section above — dynamic resolution prefix detection, `tryCatch()` per group, and `clustree()` called on the metadata data frame directly.

```{r clustree-metabolic-scores}
metab_scores <- c("Metab_glycolysis1", "Metab_oxphos1",
                  "Metab_hypoxia1", "Metab_FAO1")

n_saved <- 0

for (group_ID in group_names) {

  score_col <- NA

  tryCatch({

    seu     <- re_embed_subset_groups[[group_ID]]
    meta_df <- seu@meta.data

    # Detect the clustering resolution prefix dynamically from metadata
    # column names (never hardcoded)
    resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
    resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

    dir.create(file.path(out_dir, group_ID, "gene_level", "clustree"),
               recursive = TRUE, showWarnings = FALSE)

    for (score_col in metab_scores) {

      score_name <- sub("1$", "", score_col)

      p <- clustree(meta_df,
                    prefix           = resolution_prefix,
                    node_colour      = score_col,
                    node_colour_aggr = "mean",
                    node_size        = "size") +
        ggtitle(paste(group_ID, score_col))

      ggsave(
        filename = file.path(out_dir, group_ID, "gene_level", "clustree",
                              paste0(group_ID, "_", score_name,
                                     "_endo_mesen_re_embed", ".png")),
        plot   = p,
        width  = 10,
        height = 12,
        dpi    = 300
      )

      n_saved <- n_saved + 1
    }

  }, error = function(e) {
    warning(group_ID, " — ", score_col, " clustree failed: ", conditionMessage(e))
  })
}

message("Saved ", n_saved, " metabolic clustree PNGs under ", out_dir, " (per-group 'clustree' subdirectories)")
```

# Final clustering — supervised resolution choice
- Inspect the resolution sweep plots above before running this chunk.
- Enter your per-group resolution choices in the `chosen_resolutions` vector below.
- Then run this chunk to apply `FindClusters` at the chosen resolution for each group.

```{r}
# ============================================================
# Enter your chosen resolution per group after inspecting
# the sweep plots above. Run this chunk only after reviewing.
# ============================================================

chosen_resolutions <- c(
  "control" = 0.4,
  "ko"      = 0.5
  )

for (group_ID in group_names) {

  res <- chosen_resolutions[[group_ID]]
  message("Applying resolution ", res, " to ", group_ID)

  re_embed_subset_groups[[group_ID]] <- FindClusters(
    re_embed_subset_groups[[group_ID]],
    resolution   = res,
    cluster.name = "seurat_clusters_endo_mesen",
    verbose      = FALSE
  )

  n_clusters <- length(unique(re_embed_subset_groups[[group_ID]]$seurat_clusters_endo_mesen))
  message("  ", group_ID, ": ", n_clusters, " clusters at resolution ", res)
  }

```

# UMAP coloured by cell type

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  umap_celltype <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "new_celltypes",
      cols = celltype_cols,
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(group_ID, " - endo/mesen re-embed (cell types)"),
        caption = paste0(group_ID, " — gene level")
      ) +
      NoLegend()

  print(umap_celltype)

  dir.create(file.path(out_dir, group_ID, "gene_level", "umap_signatures"), recursive = TRUE, showWarnings = FALSE)

  save_if(umap_celltype,
          filename = file.path(out_dir, group_ID, "gene_level", "umap_signatures", paste0("12.1-umap.re_embed.", group_ID, ".new_celltype.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by new cluster

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  umap_newCluster <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "seurat_clusters_endo_mesen",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(group_ID, " - endo/mesen re-embed (new clusters)"),
        caption = paste0(group_ID, " — gene level")
      ) +
      NoLegend()

  print(umap_newCluster)

  dir.create(file.path(out_dir, group_ID, "gene_level"), recursive = TRUE, showWarnings = FALSE)

  save_if(umap_newCluster,
          filename = file.path(out_dir, group_ID, "gene_level", paste0("12.1-umap.re_embed.", group_ID, ".new_clusters.png")),
          width = 8,
          height = 8)
  }
```

# UMAP coloured by cell cycle Phase

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  umap_phase <- DimPlot(
      obj,
      reduction = "umap_endo_mesen",
      group.by = "Phase",
      label = TRUE,
      label.size = 6
      ) +
      labs(
        title = paste0(group_ID, " - endo/mesen re-embed (Phase)"),
        caption = paste0(group_ID, " — gene level")
      ) +
      NoLegend()

  print(umap_phase)

  dir.create(file.path(out_dir, group_ID, "gene_level", "umap_signatures"), recursive = TRUE, showWarnings = FALSE)

  save_if(umap_phase,
          filename = file.path(out_dir, group_ID, "gene_level", "umap_signatures", paste0("12.1-umap.re_embed.", group_ID, ".phase.png")),
          width = 8,
          height = 8)
  }
```

# Endocardial lineage markers

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  endocardial_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Ecscr", "Cdh5"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(group_ID, " Endocardial lineage markers"),
    caption = paste0("endo/mesen - re_embed - group level - gene level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(endocardial_markers)

  dir.create(file.path(out_dir, group_ID, "gene_level", "umap_signatures"), recursive = TRUE, showWarnings = FALSE)

  save_if(endocardial_markers,
          filename = file.path(out_dir, group_ID, "gene_level", "umap_signatures", paste0("12.1-umap.re_embed.", group_ID, ".endocardial_genes.png")),
          width = 8,
          height = 8)
  }
```

# Mesenchymal lineage markers

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  mesenchymal_markers <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("Postn", "Cthrc1"),
    ncol = 1,                      # side-by-side panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(group_ID, " Mesenchymal lineage markers"),
    caption = paste0("endo/mesen - re_embed - group level - gene level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(mesenchymal_markers)

  dir.create(file.path(out_dir, group_ID, "gene_level", "umap_signatures"), recursive = TRUE, showWarnings = FALSE)

  save_if(mesenchymal_markers,
          filename = file.path(out_dir, group_ID, "gene_level", "umap_signatures", paste0("12.1-umap.re_embed.", group_ID, ".mesenchymal_genes.png")),
          width = 8,
          height = 8)
  }
```

# Cell cycle scores

```{r}
for (group_ID in group_names) {

  obj <- re_embed_subset_groups[[group_ID]]

  cellCycle_feature <- FeaturePlot(
    object = obj,
    reduction = "umap_endo_mesen",
    features = c("G2M.Score", "S.Score"),
    ncol = 1,                      # vertical panels
    order = TRUE                   # high-expression cells on top
    ) &
    scale_colour_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
    name = "Expression"
    ) &
    plot_annotation(
    title = paste0(group_ID, " Cell Cycle"),
    caption = paste0("endo/mesen - re_embed - group level - gene level")
    ) &
    theme_bw(base_size = 14) &
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
      )

  print(cellCycle_feature)

  dir.create(file.path(out_dir, group_ID, "gene_level", "umap_signatures"), recursive = TRUE, showWarnings = FALSE)

  save_if(cellCycle_feature,
          filename = file.path(out_dir, group_ID, "gene_level", "umap_signatures", paste0("12.1-umap.re_embed.", group_ID, ".cellCycle.png")),
          width = 8,
          height = 8)
  }
```

# Write out the re-embedded subset objects

```{r}
for (group_ID in group_names) {
  write_rds(
    x = re_embed_subset_groups[[group_ID]],
    file = file.path(out_dir, group_ID, paste0(group_ID, "_endo_mesen_re_embed", ".rds"))
  )

  message("Saved: ", group_ID)
  }
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "12.1_session_info.txt"))
```
