---
title: "Add EGFP transgene metadata, subset to Endocardium and Mesenchyme, re-embed"
author: "Melanie Smith"
date: "2026-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE)
set.seed(42) # set seed for reproducibility
```

- This is the first script in a new `14.x` workflow, requested by Sophie: go back to the final merged object (all four samples, all cell types, cell-cycle genes retained — `outDir/06-cluster_marker_genes/final_interesting_cells_0.6.rds`), add newly (and correctly) re-quantified EGFP transgene UMI counts as cell metadata, remove the old, incorrectly-quantified `pEGFP-1` feature from the count matrix, subset to Endocardium + Mesenchyme, and re-embed.
- Unlike `12.x`/`13.x`, this workflow does **not** split by genotype or sample — all four samples stay merged together in one object throughout, only cell type is subset. The intent is to use `EGFP_UMI` to help identify which cells are Neural-Crest-derived, as a positive control before proceeding to RNA velocity.
- Cell-cycle-associated genes are retained (matching `12.x`, not `13.x`), since this is a fresh investigation independent of the earlier no-cc work.

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)

# Single Cell
library(Seurat) # Seurat (relies on dplyr/tidyr, so load after tidyverse)
library(clustree)

# Plotting
library(patchwork)
library(RColorBrewer)
```

# Set project directories

```{r}
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "14-rna_velocity")
downloads_dir <- file.path(project_repo_dir, "downloads")

full_object_path <- file.path(project_repo_dir, "outDir", "06-cluster_marker_genes", "final_interesting_cells_0.6.rds")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes  <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes  <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# User defined functions

```{r}
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
# change to TRUE to save plots
save_plots <- TRUE
```

# Load the full merged Seurat object
- All four samples, all cell types, cell-cycle genes retained.

```{r}
full_object <- read_rds(full_object_path)

stopifnot(inherits(full_object, "Seurat"))

message("Dimensions of full merged object: ", paste(dim(full_object), collapse = " x "))
print(table(full_object$orig.ident))
```

# Add EGFP transgene metadata
- Reads the four newly re-quantified transgene UMI files (one per sample), filters to `EGFP` only (the files also contain `MIB1_neo` and `ZEG_stop_pA`, which we are not interested in here), and adds two new metadata columns: `EGFP_UMI` (raw UMI count, 0 for any cell not listed with an `EGFP` row) and `EGFP_positive` (`EGFP_UMI > 0`). The EGFP transgene is a little leaky, so single-UMI detections are treated as noise and zeroed - `EGFP_UMI` only takes a nonzero value for cells with 2 or more EGFP UMIs.
- Each sample uses a single, consistent barcode suffix in the merged object (confirmed directly against `orig.ident`, not assumed), so barcode matching here is a straightforward per-sample suffix append rather than the "try successive barcodes" logic needed for the `.loom` files elsewhere in this project.
- Not every `EGFP`-tagged barcode in the transgene files is present in this QC-filtered object (this object already excludes cells that failed earlier QC) - the per-sample match rate is reported below and is expected to be well under 100%.

```{r add-egfp-metadata}
sample_suffixes <- c(
  e11_control = "_1_1",
  e11_ko      = "_2_1",
  e12_control = "_1",
  e12_ko      = "_2"
)

egfp_umi <- setNames(rep(0L, ncol(full_object)), colnames(full_object))

for (sample_ID in names(sample_suffixes)) {

  tsv_path <- file.path(downloads_dir, paste0(sample_ID, "_transgene_umis_final.tsv"))

  tsv <- read.delim(tsv_path, header = FALSE,
                     col.names = c("barcode", "gene", "umi"))

  egfp_rows <- tsv[tsv$gene == "EGFP", ]

  full_barcodes <- paste0(egfp_rows$barcode, sample_suffixes[[sample_ID]])

  matched <- full_barcodes %in% names(egfp_umi)

  message(
    sample_ID, " — EGFP rows: ", nrow(egfp_rows),
    " | matched to object: ", sum(matched),
    " (", round(100 * mean(matched), 1), "%)"
  )

  egfp_umi[full_barcodes[matched]] <- egfp_rows$umi[matched]
}

# The EGFP transgene is a little leaky - treat single-UMI detections as noise
# rather than true signal, zeroing any cell with < 2 EGFP UMIs.
n_leaky <- sum(egfp_umi == 1)
message(n_leaky, " cells with exactly 1 EGFP UMI zeroed out as leaky signal")
egfp_umi[egfp_umi < 2] <- 0

full_object <- AddMetaData(full_object, metadata = egfp_umi, col.name = "EGFP_UMI")
full_object$EGFP_positive <- full_object$EGFP_UMI > 0
full_object$EGFP_UMI_log1p <- log1p(full_object$EGFP_UMI)

message("Total EGFP-positive cells: ", sum(full_object$EGFP_positive), " of ", ncol(full_object))
print(table(full_object$EGFP_positive, full_object$new_celltypes))
```

# Remove the pEGFP-1 feature
- The original `pEGFP-1` quantification (a feature already present in this object's count matrix from the standard reference/pipeline) was not done correctly and has now been superseded by the `EGFP_UMI` metadata added above. Removing it from the count matrix so it doesn't contribute to any downstream normalisation, variable feature selection, or clustering.

```{r remove-pegfp1-feature}
stopifnot("pEGFP-1" %in% rownames(full_object))

message("Dimensions before removing pEGFP-1: ", paste(dim(full_object), collapse = " x "))

full_object <- subset(full_object, features = setdiff(rownames(full_object), "pEGFP-1"))

message("Dimensions after removing pEGFP-1: ", paste(dim(full_object), collapse = " x "))
stopifnot(!("pEGFP-1" %in% rownames(full_object)))
```

# Positive control — EGFP on all cells, all cell types
- Reuses the existing `pca`/`umap` reduction already stored in the object rather than re-embedding - removing a single low-signal feature (`pEGFP-1`) out of thousands is very unlikely to meaningfully change the embedding, so re-embedding here would add cost without changing the result.

```{r positive-control-umap, fig.width=20, fig.height=6}
positive_control_dir <- file.path(out_dir, "positive_control")
dir.create(positive_control_dir, recursive = TRUE, showWarnings = FALSE)

p1 <- DimPlot(
    full_object,
    reduction = "umap",
    group.by = "new_celltypes",
    cols = celltype_cols,
    label = TRUE,
    label.size = 6
  ) +
  labs(
    title = "All cells — cell type",
    caption = "Full merged object — positive control"
  ) +
  NoLegend()

p2 <- DimPlot(
    full_object,
    reduction = "umap",
    group.by = "EGFP_positive",
    cols = c("TRUE" = "#4DAF4A", "FALSE" = "grey85")
  ) +
  labs(
    title = "All cells — EGFP positive",
    caption = "Full merged object — positive control"
  )

p3 <- FeaturePlot(
    full_object,
    reduction = "umap",
    features = "EGFP_UMI_log1p",
    order = TRUE
  ) +
  # Spectral has no neutral colour at the low end, so zero-value cells still
  # render as a solid, visible colour - misleading for a sparse feature like
  # this. grey90 lets zero-value cells fade into the background instead.
  scale_colour_gradientn(
    colours = c("grey90", "gold", "red"),
    name = "log1p(EGFP UMI)"
  ) +
  labs(
    title = "All cells — EGFP UMI",
    caption = "Full merged object — positive control"
  )

positive_control_plot <- p1 + p2 + p3
print(positive_control_plot)

save_if(positive_control_plot,
        filename = file.path(positive_control_dir, "14.1-umap.all_cells.celltype_egfp.png"),
        width = 20,
        height = 6)
```

## EGFP UMAP split by sample

```{r positive-control-umap-split-sample, fig.width=20, fig.height=6}
egfp_split_sample_plot <- FeaturePlot(
    full_object,
    reduction = "umap",
    features = "EGFP_UMI_log1p",
    split.by = "orig.ident",
    order = TRUE
  ) &
  # `FeaturePlot(..., split.by=...)` returns a multi-panel patchwork object -
  # `+` here would only modify the LAST panel, `&` applies to all of them
  scale_colour_gradientn(
    colours = c("grey90", "gold", "red"),
    name = "log1p(EGFP UMI)"
  ) &
  labs(
    caption = "Full merged object — positive control"
  )

print(egfp_split_sample_plot)

save_if(egfp_split_sample_plot,
        filename = file.path(positive_control_dir, "14.1-umap.all_cells.egfp_by_sample.png"),
        width = 20,
        height = 6)
```

# Cell type counts (QC before subsetting)

```{r}
print(table(full_object$new_celltypes))
```

# Subset to Endocardium and Mesenchyme

```{r}
message("Dimensions before subsetting: ", paste(dim(full_object), collapse = " x "))

subset_object <- subset(
  full_object,
  subset = new_celltypes %in% c("Endocardium", "Mesenchyme")
)

message("Dimensions after subsetting: ", paste(dim(subset_object), collapse = " x "))

rm(full_object)
gc()
```

# Re-embedding function
- Re-embed the Endocardium + Mesenchyme subset, mirroring `12.1`'s `re_embed_subset()` mechanics.

```{r}
re_embed_subset <- function(seurat_obj,
                             n_dims = 30,
                             nfeatures = 3000) {

  DefaultAssay(seurat_obj) <- "RNA"
  Idents(seurat_obj) <- seurat_obj$new_celltypes

  re_embed_seurat_obj <- FindVariableFeatures(
    seurat_obj,
    selection.method = "vst",
    nfeatures        = nfeatures,
    verbose          = FALSE
  )

  re_embed_seurat_obj <- ScaleData(
    re_embed_seurat_obj,
    features = rownames(re_embed_seurat_obj),
    verbose  = FALSE
  )

  re_embed_seurat_obj <- RunPCA(
    re_embed_seurat_obj,
    reduction.name = "pca_endo_mesen",
    reduction.key  = "PCEndoMesen_",
    verbose        = FALSE
  )

  re_embed_seurat_obj <- FindNeighbors(
    re_embed_seurat_obj,
    reduction = "pca_endo_mesen",
    dims      = 1:n_dims,
    verbose   = FALSE
  )

  re_embed_seurat_obj <- RunUMAP(
    re_embed_seurat_obj,
    reduction      = "pca_endo_mesen",
    dims           = 1:n_dims,
    reduction.name = "umap_endo_mesen",
    reduction.key  = "UMAPEndoMesen_",
    verbose        = FALSE
  )

  return(re_embed_seurat_obj)
}
```

```{r}
re_embed_subset_obj <- re_embed_subset(subset_object)

rm(subset_object)
gc()

Reductions(re_embed_subset_obj)
```

# EndMT Module Scoring

## Define EndMT gene sets

```{r endmt-gene-sets}
# EndMT gene set 1: Endocardial identity (pre-EndMT)
genes_EndMT_endocardial <- c("Nfatc1", "Npr3", "Cdh5", "Pecam1", "Tie1",
                              "Gata2", "Klf2", "Klf4", "Sox17", "Hey1", "Hey2")

# EndMT gene set 2: EndMT induction / transition state
genes_EndMT_transition <- c("Snai1", "Snai2", "Twist1", "Twist2", "Zeb1", "Zeb2",
                             "Tbx20", "Has2", "Vcan", "Notch1", "Notch2",
                             "Bmp2", "Tgfb2", "Tgfb3")

# EndMT gene set 3: Valvular mesenchyme (post-EndMT)
genes_EndMT_mesenchyme <- c("Postn", "Col1a1", "Col3a1", "Vim", "Fn1",
                             "Tcf21", "Pdgfra", "Sox9", "Msx1", "Msx2",
                             "Twist1", "Acta2")

# EndMT gene set 4: Notch signaling axis (EndMT-driving)
genes_EndMT_Notch <- c("Notch1", "Notch2", "Jag1", "Dll4", "Hey1", "Hey2",
                        "Hes1", "Nrg1", "Erbb2", "Erbb4", "Snai2")

# EndMT gene set 5: TGF-beta / BMP signaling axis
genes_EndMT_TGFb <- c("Tgfb2", "Tgfb3", "Bmp2", "Bmp4", "Smad1", "Smad2",
                       "Smad3", "Smad5", "Id1", "Id2", "Alk2", "Alk3",
                       "Alk5", "Eng")
```

## Run module scoring

```{r endmt-module-scoring}
gene_sets <- list(
  EndMT_endocardial = genes_EndMT_endocardial,
  EndMT_transition  = genes_EndMT_transition,
  EndMT_mesenchyme  = genes_EndMT_mesenchyme,
  EndMT_Notch       = genes_EndMT_Notch,
  EndMT_TGFb        = genes_EndMT_TGFb
)

for (set_name in names(gene_sets)) {

  re_embed_subset_obj <- tryCatch({
    AddModuleScore(
      object   = re_embed_subset_obj,
      features = list(gene_sets[[set_name]]),
      name     = set_name,
      nbin     = 24,
      ctrl     = 5
    )
  }, error = function(e) {
    warning(set_name, " scoring failed: ", conditionMessage(e))
    re_embed_subset_obj
  })
}
```

# Metabolic State Module Scoring

## Define metabolic state gene sets

```{r metabolic-state-gene-sets}
genes_Metab_glycolysis <- c("Hk2", "Pkm", "Ldha", "Pgk1", "Eno1",
                             "Pfkl", "Slc2a1", "Aldoa", "Pfkfb3", "Gapdh")

genes_Metab_oxphos <- c("Ndufb3", "Sdhb", "Uqcrc1", "Cox5a", "Atp5f1b",
                         "Cycs", "Sdha", "Uqcrq", "Cox6a1", "Ndufa4")

genes_Metab_hypoxia <- c("Hif1a", "Vegfa", "Slc2a1", "Pgk1", "Ldha",
                          "Bnip3", "Egln1", "Pdk1", "Cited2", "Epas1")

genes_Metab_FAO <- c("Ppara", "Cpt1b", "Acadm", "Hadha", "Hadhb",
                      "Acadvl", "Acaa2", "Echs1", "Slc25a20", "Fabp3")
```

## Run module scoring

```{r metabolic-state-module-scoring}
gene_sets <- list(
  Metab_glycolysis = genes_Metab_glycolysis,
  Metab_oxphos     = genes_Metab_oxphos,
  Metab_hypoxia    = genes_Metab_hypoxia,
  Metab_FAO        = genes_Metab_FAO
)

for (set_name in names(gene_sets)) {

  re_embed_subset_obj <- tryCatch({
    AddModuleScore(
      object   = re_embed_subset_obj,
      features = list(gene_sets[[set_name]]),
      name     = set_name,
      nbin     = 24,
      ctrl     = 5
    )
  }, error = function(e) {
    warning(set_name, " scoring failed: ", conditionMessage(e))
    re_embed_subset_obj
  })
}

colnames(re_embed_subset_obj[[]])
```

# UMAP by EndMT and Metabolic Signature

```{r umap-signatures}
signatures <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                "EndMT_Notch1", "EndMT_TGFb1",
                "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

umap_signatures_dir <- file.path(out_dir, "gene_level", "umap_signatures")
dir.create(umap_signatures_dir, recursive = TRUE, showWarnings = FALSE)

for (sig in signatures) {

  sig_name <- sub("1$", "", sig)

  p <- FeaturePlot(
    re_embed_subset_obj,
    reduction  = "umap_endo_mesen",
    features = sig,
    label      = TRUE,
    label.size = 6
  ) +
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) +
    labs(
      title    = sig,
      caption  = "Endocardium + Mesenchyme subset — gene level"
    )

  print(p)

  ggsave(
    filename = file.path(umap_signatures_dir, paste0(sig_name, "_endo_mesen_re_embed.png")),
    plot = p, width = 8, height = 8, dpi = 300
  )
}
```

# Resolution sweep
- Run this chunk and inspect the plots before proceeding.
- Choose a resolution based on the plots below, then enter it in the **final clustering** chunk further down.

```{r resolution-sweep}
resolution_summary <- data.frame()

resolution_sweep_dir <- file.path(out_dir, "gene_level", "resolution_sweep")
dir.create(resolution_sweep_dir, recursive = TRUE, showWarnings = FALSE)

clustree_dir <- file.path(out_dir, "gene_level", "clustree")
dir.create(clustree_dir, recursive = TRUE, showWarnings = FALSE)

for (res in seq(0.1, 1.0, by = 0.1)) {

  col_name <- paste0("res_", res)

  re_embed_subset_obj <- FindClusters(
    re_embed_subset_obj,
    resolution   = res,
    cluster.name = col_name,
    verbose      = FALSE
  )

  n_clusters <- length(unique(re_embed_subset_obj[[col_name]][[1]]))
  message("resolution ", res, ": ", n_clusters, " clusters")

  resolution_summary <- rbind(
    resolution_summary,
    data.frame(resolution = res, n_clusters = n_clusters)
  )
}

for (res in seq(0.1, 1.0, by = 0.1)) {

  col_name   <- paste0("res_", res)
  n_clusters <- length(unique(re_embed_subset_obj[[col_name]][[1]]))

  res_plot <- DimPlot(
      re_embed_subset_obj,
      reduction  = "umap_endo_mesen",
      group.by   = col_name,
      label      = TRUE,
      label.size = 6
    ) +
      labs(
        title    = paste0("resolution: ", res),
        subtitle = paste0(n_clusters, " clusters"),
        caption  = "Endocardium + Mesenchyme subset — gene level"
      ) +
      NoLegend()

  print(res_plot)

  ggsave(
    filename = file.path(resolution_sweep_dir, paste0("resolution_", res, "_endo_mesen_re_embed.png")),
    plot   = res_plot,
    width  = 8,
    height = 8,
    dpi    = 300
  )
}

stability_plot <- clustree(
    re_embed_subset_obj,
    node_colour = "sc3_stability",
    prefix = "res_"
  ) +
    labs(
      title   = "clustree — sc3_stability",
      caption = "Endocardium + Mesenchyme subset — gene level"
    )

print(stability_plot)

ggsave(
  filename = file.path(clustree_dir, "sc3_stability_endo_mesen_re_embed.png"),
  plot   = stability_plot,
  width  = 10,
  height = 12,
  dpi    = 300
)

print(resolution_summary)

write.csv(
  resolution_summary,
  file.path(out_dir, "gene_level_resolution_sweep_summary.csv"),
  row.names = FALSE
)
```

```{r clustree-overlay}
umap_coords <- Embeddings(re_embed_subset_obj, reduction = "umap_endo_mesen")
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
re_embed_subset_obj <- AddMetaData(re_embed_subset_obj, metadata = as.data.frame(umap_coords))

print(
  clustree_overlay(
    re_embed_subset_obj,
    prefix  = "res_",
    x_value = "UMAP_1",
    y_value = "UMAP_2"
  ) +
    labs(
      title   = "clustree overlay on UMAP",
      caption = "Endocardium + Mesenchyme subset — gene level"
    )
)
```

```{r clustree-exploration}
sig_genes <- c("Ecscr", "Cdh5", "Postn", "Cthrc1")

for (gene in sig_genes) {

  p <- clustree(
    re_embed_subset_obj,
    node_colour = gene,
    node_colour_aggr = "mean",
    prefix = "res_"
    ) +
    labs(
      title   = paste0("clustree — ", gene, " avg exp."),
      caption = "Endocardium + Mesenchyme subset — gene level"
      )

  print(p)

  ggsave(
    filename = file.path(clustree_dir, paste0(gene, "_endo_mesen_re_embed.png")),
    plot   = p,
    width  = 10,
    height = 12,
    dpi    = 300
  )
}
```

# UMAP by Cell Type (pre-final-clustering)

```{r umap-new-celltypes-exploration}
p <- DimPlot(
        re_embed_subset_obj,
        reduction  = "umap_endo_mesen",
        group.by   = "new_celltypes",
        label      = TRUE,
        label.size = 6
      ) +
        labs(
          title    = "new_celltypes",
          caption  = "Endocardium + Mesenchyme subset — gene level"
        ) +
        NoLegend()

print(p)

ggsave(
  filename = file.path(umap_signatures_dir, "new_celltypes_endo_mesen_re_embed.png"),
  plot = p, width = 8, height = 8, dpi = 300
)
```

# Clustree: EndMT Module Scores

```{r clustree-endmt-scores}
endmt_scores <- c("EndMT_endocardial1", "EndMT_transition1",
                  "EndMT_mesenchyme1", "EndMT_Notch1", "EndMT_TGFb1")

n_saved <- 0

meta_df <- re_embed_subset_obj@meta.data
resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

for (score_col in endmt_scores) {

  score_name <- sub("1$", "", score_col)

  p <- clustree(meta_df,
                prefix           = resolution_prefix,
                node_colour      = score_col,
                node_colour_aggr = "mean",
                node_size        = "size") +
    ggtitle(score_col)

  ggsave(
    filename = file.path(clustree_dir, paste0(score_name, "_endo_mesen_re_embed.png")),
    plot   = p,
    width  = 10,
    height = 12,
    dpi    = 300
  )

  n_saved <- n_saved + 1
}

message("Saved ", n_saved, " EndMT clustree PNGs")
```

# Clustree: Metabolic Module Scores

```{r clustree-metabolic-scores}
metab_scores <- c("Metab_glycolysis1", "Metab_oxphos1",
                  "Metab_hypoxia1", "Metab_FAO1")

n_saved <- 0

for (score_col in metab_scores) {

  score_name <- sub("1$", "", score_col)

  p <- clustree(meta_df,
                prefix           = resolution_prefix,
                node_colour      = score_col,
                node_colour_aggr = "mean",
                node_size        = "size") +
    ggtitle(score_col)

  ggsave(
    filename = file.path(clustree_dir, paste0(score_name, "_endo_mesen_re_embed.png")),
    plot   = p,
    width  = 10,
    height = 12,
    dpi    = 300
  )

  n_saved <- n_saved + 1
}

message("Saved ", n_saved, " metabolic clustree PNGs")
```

# Final clustering — supervised resolution choice
- Inspect the resolution sweep plots above before running this chunk.

```{r}
chosen_resolution <- 0.4

re_embed_subset_obj <- FindClusters(
  re_embed_subset_obj,
  resolution   = chosen_resolution,
  cluster.name = "seurat_clusters_endo_mesen",
  verbose      = FALSE
)

n_clusters <- length(unique(re_embed_subset_obj$seurat_clusters_endo_mesen))
message("Applying resolution ", chosen_resolution, ": ", n_clusters, " clusters")
```

# UMAP coloured by cell type

```{r}
umap_celltype <- DimPlot(
    re_embed_subset_obj,
    reduction = "umap_endo_mesen",
    group.by = "new_celltypes",
    cols = celltype_cols,
    label = TRUE,
    label.size = 6
    ) +
    labs(
      title = "endo/mesen re-embed (cell types)",
      caption = "gene level"
    ) +
    NoLegend()

print(umap_celltype)

save_if(umap_celltype,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.new_celltype.png"),
        width = 8,
        height = 8)
```

# UMAP coloured by new cluster

```{r}
umap_newCluster <- DimPlot(
    re_embed_subset_obj,
    reduction = "umap_endo_mesen",
    group.by = "seurat_clusters_endo_mesen",
    label = TRUE,
    label.size = 6
    ) +
    labs(
      title = "endo/mesen re-embed (new clusters)",
      caption = "gene level"
    ) +
    NoLegend()

print(umap_newCluster)

save_if(umap_newCluster,
        filename = file.path(out_dir, "gene_level", "14.1-umap.re_embed.new_clusters.png"),
        width = 8,
        height = 8)
```

# UMAP coloured by cell cycle Phase

```{r}
umap_phase <- DimPlot(
    re_embed_subset_obj,
    reduction = "umap_endo_mesen",
    group.by = "Phase",
    label = TRUE,
    label.size = 6
    ) +
    labs(
      title = "endo/mesen re-embed (Phase)",
      caption = "gene level"
    ) +
    NoLegend()

print(umap_phase)

save_if(umap_phase,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.phase.png"),
        width = 8,
        height = 8)
```

# EGFP on the re-embedded subset
- Same `EGFP_UMI_log1p`/`EGFP_positive` metadata added earlier, now on the Endocardium + Mesenchyme-only embedding.

```{r}
umap_egfp_positive <- DimPlot(
    re_embed_subset_obj,
    reduction = "umap_endo_mesen",
    group.by = "EGFP_positive",
    cols = c("TRUE" = "#4DAF4A", "FALSE" = "grey85")
    ) +
    labs(
      title = "endo/mesen re-embed (EGFP positive)",
      caption = "gene level"
    )

print(umap_egfp_positive)

save_if(umap_egfp_positive,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.egfp_positive.png"),
        width = 8,
        height = 8)

umap_egfp_umi <- FeaturePlot(
    re_embed_subset_obj,
    reduction = "umap_endo_mesen",
    features = "EGFP_UMI_log1p",
    order = TRUE
    ) +
    # grey90 low end so zero-value cells fade into the background rather than
    # rendering as a solid, misleading colour (see positive-control chunk above)
    scale_colour_gradientn(
      colours = c("grey90", "gold", "red"),
      name = "log1p(EGFP UMI)"
    ) +
    labs(
      title = "endo/mesen re-embed (EGFP UMI)",
      caption = "gene level"
    )

print(umap_egfp_umi)

save_if(umap_egfp_umi,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.egfp_umi.png"),
        width = 8,
        height = 8)
```

# Endocardial lineage markers

```{r}
endocardial_markers <- FeaturePlot(
  object = re_embed_subset_obj,
  reduction = "umap_endo_mesen",
  features = c("Ecscr", "Cdh5"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Endocardial lineage markers",
  caption = "endo/mesen - re_embed - gene level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(endocardial_markers)

save_if(endocardial_markers,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.endocardial_genes.png"),
        width = 8,
        height = 8)
```

# Mesenchymal lineage markers

```{r}
mesenchymal_markers <- FeaturePlot(
  object = re_embed_subset_obj,
  reduction = "umap_endo_mesen",
  features = c("Postn", "Cthrc1"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Mesenchymal lineage markers",
  caption = "endo/mesen - re_embed - gene level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(mesenchymal_markers)

save_if(mesenchymal_markers,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.mesenchymal_genes.png"),
        width = 8,
        height = 8)
```

# Cell cycle scores

```{r}
cellCycle_feature <- FeaturePlot(
  object = re_embed_subset_obj,
  reduction = "umap_endo_mesen",
  features = c("G2M.Score", "S.Score"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Cell Cycle",
  caption = "endo/mesen - re_embed - gene level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(cellCycle_feature)

save_if(cellCycle_feature,
        filename = file.path(out_dir, "gene_level", "umap_signatures", "14.1-umap.re_embed.cellCycle.png"),
        width = 8,
        height = 8)
```

# Write out the re-embedded subset object

```{r}
write_rds(
  x = re_embed_subset_obj,
  file = file.path(out_dir, "all_samples_endo_mesen_re_embed.rds")
)

message("Saved: all_samples_endo_mesen_re_embed.rds")
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "14.1_session_info.txt"))
```
