---
title: "Import Velocyto spliced/unspliced counts for Endocardium and Mesenchyme (all samples merged)"
author: "Melanie Smith"
date: "2026-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE)
set.seed(42) # set seed for reproducibility
```

- This script imports the Velocyto spliced/unspliced/ambiguous counts for the Endocardium/Mesenchyme subset created in `14.1`, subsets them to the cells and genes retained in that subset, re-embeds, and exports the combined object as an `.h5ad` for scVelo/PAGA.
- Route B only (reads the raw `.loom` files directly), mirroring `12.2`/`13.2`. Unlike those scripts, all four samples are combined into **one single object** here, not split by genotype - matching `14.1`'s "all samples merged, no group split" design.
- Barcode suffixes are known deterministically for this object lineage (confirmed directly against `14.1`'s output, unchanged by subsetting), so matching here is a straightforward per-sample suffix append, same approach used for the EGFP transgene files in `14.1` - no "try successive barcodes" fallback needed.
- All `14.1` metadata is transferred across, including `EGFP_UMI`/`EGFP_positive`/`EGFP_UMI_log1p` and the EndMT/metabolic module scores.

# Load Libraries

```{r}
reticulate::py_require(c("anndata==0.10.9", "pandas<3.0"))

library(tidyverse)
library(Seurat)
library(Matrix)
library(ggokabeito)
library(cowplot)
library(patchwork)
library(reticulate)
library(sceasy)
library(anndata)
library(scCustomize)
library(clustree)
```

# Set project directories and parameters

```{r}
project_repo_dir <- "/home/melanie-smith/workDir/sophieWiszniak/20251104_sophieWiszniak_ncc_pa"
out_dir <- file.path(project_repo_dir, "outDir", "14-rna_velocity")

sample_names <- c("e11_control", "e11_ko", "e12_control", "e12_ko")

# Barcode suffixes confirmed directly against the 14.1 output object
sample_suffixes <- c(
  e11_control = "_1_1",
  e11_ko      = "_2_1",
  e12_control = "_1",
  e12_ko      = "_2"
  )

# Named vector mapping each sample to its Velocyto loom column prefix
loom_colnames <- c(
  "e11_control" = "possorted_genome_bam_5Q1UF:",
  "e11_ko"      = "possorted_genome_bam_LZRTJ:",
  "e12_control" = "possorted_genome_bam_YIL7N:",
  "e12_ko"      = "possorted_genome_bam_0YX73:"
  )
```

## Set up the cell-type colours and marker genes

```{r}
celltype_cols <- c("Myocardium"    = "#E41A1C",
                   "VSMC"          = "#984EA3",
                   "Endocardium"   = "#377EB8",
                   "Epicardium"    = "#FF7F00",
                   "Mesenchyme"    = "#4DAF4A",
                   "Other/Unknown" = "grey70")

endocardial_genes <- c("Ecscr", "Cdh5", "Tie1", "Rasip1", "Plxnd1")
mesenchymal_genes <- c("Postn", "Cthrc1", "Col1a1", "Fn1")
```

# Helper functions

```{r}
save_if <- function(plot_object, filename, width, height, ...) {
  if (save_plots) {
    ggsave(plot = plot_object, filename = filename,
           width = width, height = height, dpi = 300, ...)
  }
}
save_plots <- TRUE
```

# Load the Endocardium and Mesenchyme subset Seurat object
- This is the object saved by `14.1`. It contains gene-level counts, metadata (including `EGFP_UMI`/`EGFP_positive`), and the `umap_endo_mesen`/`pca_endo_mesen` reductions.
- Cell barcodes here define which cells are retained from the loom (spliced/unspliced/ambiguous) data.

```{r}
re_embed_subset_obj <- read_rds(
  file = file.path(out_dir, "all_samples_endo_mesen_re_embed.rds")
  )
```

# Subset Loom Layers to 14.1 Cells and Genes
- Loops over all four samples, reads each sample's Velocyto `.loom` file directly (so the `ambiguous` layer is available), matches barcodes against the cells from that sample retained in the `14.1` object (identified via `orig.ident`), then combines all four samples' matrices into one matrix per layer.

```{r subset-loom-to-14point1-cells}
retained_genes <- rownames(re_embed_subset_obj)

group_spliced_parts   <- list()
group_unspliced_parts <- list()
group_ambiguous_parts <- list()

for (sample_ID in sample_names) {

  velocyto_dir <- file.path(project_repo_dir, "outDir", "09-rna_velocity", sample_ID, "velocyto_output")
  loom_file    <- file.path(velocyto_dir, paste0(sub(":$", "", loom_colnames[[sample_ID]]), ".loom"))

  lfile <- loomR::connect(filename = loom_file, mode = "r", skip.validate = TRUE)

  # loomR returns layers oriented cells (rows) x genes (columns)
  loom_barcodes <- lfile$col.attrs$CellID[]
  loom_genes    <- lfile$row.attrs$Gene[]

  n_cells_original <- length(loom_barcodes)

  clean_barcodes <- gsub(loom_colnames[[sample_ID]], "", loom_barcodes)
  clean_barcodes <- gsub("x$", "", clean_barcodes)

  # sample_suffixes only captures what follows "-1" in the Seurat barcode
  # (that's how it was derived in 14.1, matching against TSV barcodes that
  # already included "-1"). The loom barcodes don't have "-1" after cleaning
  # (just the bare 16bp sequence), so it needs to be reinserted here.
  full_barcodes <- paste0(clean_barcodes, "-1", sample_suffixes[[sample_ID]])

  retained_cells_this_sample <- colnames(re_embed_subset_obj)[
    re_embed_subset_obj$orig.ident == sample_ID
  ]

  cell_match_idx <- match(retained_cells_this_sample, full_barcodes)
  pct_matched    <- mean(!is.na(cell_match_idx))

  if (pct_matched < 0.8) {
    warning(sample_ID, " — only ", round(pct_matched * 100, 1),
            "% of 14.1 cells matched in the loom file (expected >= 80%)")
  }

  keep_cells_idx <- cell_match_idx[!is.na(cell_match_idx)]
  keep_cells     <- retained_cells_this_sample[!is.na(cell_match_idx)]

  gene_match_idx <- match(retained_genes, loom_genes)
  keep_genes_idx <- gene_match_idx[!is.na(gene_match_idx)]
  keep_genes     <- retained_genes[!is.na(gene_match_idx)]

  # Read each layer, subset to retained cells/genes, transpose to genes x
  # cells (Seurat convention), wrap as sparse, then discard the dense
  # intermediate before moving to the next layer
  read_layer_subset <- function(layer_name) {
    full_mat <- lfile[[paste0("layers/", layer_name)]][, ]
    sub_mat  <- t(full_mat[keep_cells_idx, keep_genes_idx])
    rownames(sub_mat) <- keep_genes
    colnames(sub_mat) <- keep_cells
    sub_mat <- Matrix::Matrix(sub_mat, sparse = TRUE)
    rm(full_mat)
    gc()
    sub_mat
  }

  group_spliced_parts[[sample_ID]]   <- read_layer_subset("spliced")
  group_unspliced_parts[[sample_ID]] <- read_layer_subset("unspliced")
  group_ambiguous_parts[[sample_ID]] <- read_layer_subset("ambiguous")

  lfile$close_all()

  message(
    sample_ID,
    " — cells: ", n_cells_original, " -> ", length(keep_cells),
    " | genes matched: ", length(keep_genes),
    " (", round(100 * pct_matched, 1), "% of 14.1 cells matched)"
  )
  }

# Combine all four samples' matrices (same genes, different cells) into one matrix per layer
loom_spliced_subset   <- do.call(cbind, group_spliced_parts)
loom_unspliced_subset <- do.call(cbind, group_unspliced_parts)
loom_ambiguous_subset <- do.call(cbind, group_ambiguous_parts)

message("Combined cells: ", ncol(loom_spliced_subset))
```

# Create Spliced/Unspliced Seurat Object and Re-embed
- Build a new Seurat object from the combined spliced counts, attach the unspliced/ambiguous layers as additional assays, transfer all `14.1` metadata (including `EGFP_UMI`/`EGFP_positive` and the EndMT/Metab module scores), then re-embed and sweep the same resolution sequence used in `14.1` (`seq(0.1, 1.0, by = 0.1)`, `res_` naming). Normalisation, scaling, PCA, neighbours, clustering, and UMAP all run on the spliced assay only - unspliced/ambiguous are carried as separate assays for the later velocity model fit, not used to build this embedding.

```{r create-seurat-spliced-reembed}
resolution_sequence <- seq(0.1, 1.0, by = 0.1)
n_dims_spliced       <- 30
nfeatures_spliced    <- 3000

spliced_subset_obj <- CreateSeuratObject(counts = loom_spliced_subset)

spliced_subset_obj[["unspliced"]] <- CreateAssayObject(counts = loom_unspliced_subset)
spliced_subset_obj[["ambiguous"]] <- CreateAssayObject(counts = loom_ambiguous_subset)

# Transfer all metadata from the 14.1 subset object, matched by barcode
source_meta <- re_embed_subset_obj@meta.data[colnames(spliced_subset_obj), ]
spliced_subset_obj <- AddMetaData(spliced_subset_obj, metadata = source_meta)

DefaultAssay(spliced_subset_obj) <- "RNA"

spliced_subset_obj <- NormalizeData(spliced_subset_obj, verbose = FALSE)
spliced_subset_obj <- FindVariableFeatures(spliced_subset_obj, selection.method = "vst",
                                            nfeatures = nfeatures_spliced, verbose = FALSE)
spliced_subset_obj <- ScaleData(spliced_subset_obj, verbose = FALSE)
spliced_subset_obj <- RunPCA(spliced_subset_obj, reduction.name = "pca_spliced",
                              reduction.key = "PCspliced_", verbose = FALSE)
spliced_subset_obj <- FindNeighbors(spliced_subset_obj, reduction = "pca_spliced",
                                     dims = 1:n_dims_spliced, verbose = FALSE)

for (res in resolution_sequence) {
  spliced_subset_obj <- FindClusters(spliced_subset_obj, resolution = res,
                                      cluster.name = paste0("res_", res), verbose = FALSE)
  }

spliced_subset_obj <- RunUMAP(spliced_subset_obj, reduction = "pca_spliced", dims = 1:n_dims_spliced,
                               reduction.name = "umap_spliced", reduction.key = "UMAPspliced_", verbose = FALSE)

message("Finished re-embedding. n cells: ", ncol(spliced_subset_obj))
```

# Spliced/Unspliced Resolution Sweep Plots
- Reproduces `14.1`'s resolution-sweep figures (per-resolution UMAP + stability clustree) for the spliced/unspliced embedding.

```{r spliced-resolution-sweep-plots}
resolution_sweep_dir <- file.path(out_dir, "spliced_unspliced", "resolution_sweep")
dir.create(resolution_sweep_dir, recursive = TRUE, showWarnings = FALSE)

clustree_dir <- file.path(out_dir, "spliced_unspliced", "clustree")
dir.create(clustree_dir, recursive = TRUE, showWarnings = FALSE)

for (res in resolution_sequence) {

  col_name   <- paste0("res_", res)
  n_clusters <- length(unique(spliced_subset_obj[[col_name]][[1]]))

  p <- DimPlot(
    spliced_subset_obj,
    reduction  = "umap_spliced",
    group.by   = col_name,
    label      = TRUE,
    label.size = 6
  ) +
    labs(
      title    = paste0("resolution: ", res),
      subtitle = paste0(n_clusters, " clusters"),
      caption  = "Endocardium + Mesenchyme subset — spliced/unspliced level"
    ) +
    NoLegend()

  print(p)

  ggsave(
    filename = file.path(resolution_sweep_dir, paste0("14.2-umap.spliced_unspliced_re_embed.res_", res, ".png")),
    plot = p, width = 8, height = 8, dpi = 300
  )
  }

stability_plot <- clustree(
    spliced_subset_obj,
    node_colour = "sc3_stability",
    prefix = "res_"
  ) +
    labs(
      title   = "clustree — sc3_stability",
      caption = "Endocardium + Mesenchyme subset — spliced/unspliced level"
    )

print(stability_plot)

ggsave(
  filename = file.path(clustree_dir, "14.2-clustree.spliced_unspliced_re_embed.sc3_stability.png"),
  plot = stability_plot, width = 10, height = 12, dpi = 300
  )
```

# Spliced/Unspliced Clustree Exploration

```{r spliced-clustree-exploration}
sig_genes <- c("Ecscr", "Cdh5", "Postn", "Cthrc1")

for (gene in sig_genes) {

  p <- clustree(
    spliced_subset_obj,
    node_colour      = gene,
    node_colour_aggr = "mean",
    prefix = "res_"
    ) +
    labs(
      title   = paste0("clustree — ", gene, " avg exp."),
      caption = "Endocardium + Mesenchyme subset — spliced/unspliced level"
      )

  print(p)

  ggsave(
    filename = file.path(clustree_dir, paste0("14.2-clustree.spliced_unspliced_re_embed.", gene, ".png")),
    plot = p, width = 10, height = 12, dpi = 300
  )
  }
```

# Spliced/Unspliced UMAP by Cell Type

```{r spliced-umap-celltype}
umap_signatures_dir <- file.path(out_dir, "spliced_unspliced", "umap_signatures")
dir.create(umap_signatures_dir, recursive = TRUE, showWarnings = FALSE)

p <- DimPlot(
    spliced_subset_obj,
    reduction = "umap_spliced",
    group.by = "new_celltypes",
    cols = celltype_cols,
    label = TRUE,
    label.size = 6
    ) +
    labs(
      title = "spliced/unspliced re-embed (cell types)",
      caption = "spliced/unspliced level"
    ) +
    NoLegend()

print(p)

save_if(p,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.new_celltype.png"),
        width = 8,
        height = 8)
```

# Spliced/Unspliced UMAP by Cell Cycle Phase

```{r spliced-umap-phase}
p <- DimPlot(
    spliced_subset_obj,
    reduction = "umap_spliced",
    group.by = "Phase",
    label = TRUE,
    label.size = 6
    ) +
    labs(
      title = "spliced/unspliced re-embed (Phase)",
      caption = "spliced/unspliced level"
    ) +
    NoLegend()

print(p)

save_if(p,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.phase.png"),
        width = 8,
        height = 8)
```

# EGFP on the spliced/unspliced embedding

```{r spliced-umap-egfp}
umap_egfp_positive <- DimPlot(
    spliced_subset_obj,
    reduction = "umap_spliced",
    group.by = "EGFP_positive",
    cols = c("TRUE" = "#4DAF4A", "FALSE" = "grey85")
    ) +
    labs(
      title = "spliced/unspliced re-embed (EGFP positive)",
      caption = "spliced/unspliced level"
    )

print(umap_egfp_positive)

save_if(umap_egfp_positive,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.egfp_positive.png"),
        width = 8, height = 8)

umap_egfp_umi <- FeaturePlot(
    spliced_subset_obj,
    reduction = "umap_spliced",
    features = "EGFP_UMI_log1p",
    order = TRUE
    ) +
    scale_colour_gradientn(
      colours = c("grey90", "gold", "red"),
      name = "log1p(EGFP UMI)"
    ) +
    labs(
      title = "spliced/unspliced re-embed (EGFP UMI)",
      caption = "spliced/unspliced level"
    )

print(umap_egfp_umi)

save_if(umap_egfp_umi,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.egfp_umi.png"),
        width = 8, height = 8)
```

# Spliced/Unspliced Endocardial Lineage Markers

```{r spliced-endocardial-markers}
p <- FeaturePlot(
  object = spliced_subset_obj,
  reduction = "umap_spliced",
  features = c("Ecscr", "Cdh5"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Endocardial lineage markers",
  caption = "spliced/unspliced level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(p)

save_if(p,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.endocardial_genes.png"),
        width = 8, height = 8)
```

# Spliced/Unspliced Mesenchymal Lineage Markers

```{r spliced-mesenchymal-markers}
p <- FeaturePlot(
  object = spliced_subset_obj,
  reduction = "umap_spliced",
  features = c("Postn", "Cthrc1"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Mesenchymal lineage markers",
  caption = "spliced/unspliced level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(p)

save_if(p,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.mesenchymal_genes.png"),
        width = 8, height = 8)
```

# Spliced/Unspliced Cell Cycle Scores

```{r spliced-cellcycle-scores}
p <- FeaturePlot(
  object = spliced_subset_obj,
  reduction = "umap_spliced",
  features = c("G2M.Score", "S.Score"),
  ncol = 1,
  order = TRUE
  ) &
  scale_colour_gradientn(
  colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
  name = "Expression"
  ) &
  plot_annotation(
  title = "Cell Cycle",
  caption = "spliced/unspliced level"
  ) &
  theme_bw(base_size = 14) &
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
    )

print(p)

save_if(p,
        filename = file.path(umap_signatures_dir, "14.2-umap.spliced_unspliced_re_embed.cellCycle.png"),
        width = 8, height = 8)
```

# Spliced/Unspliced Module Score FeaturePlots

```{r spliced-module-score-featureplots}
module_scores <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                    "EndMT_Notch1", "EndMT_TGFb1",
                    "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

for (score_col in module_scores) {

  p <- FeaturePlot(
    spliced_subset_obj,
    reduction  = "umap_spliced",
    features   = score_col,
    label      = TRUE,
    label.size = 6
  ) +
    scale_colour_gradientn(
      colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
      name = "Expression"
    ) +
    labs(
      title   = score_col,
      caption = "Endocardium + Mesenchyme subset — spliced/unspliced level"
    )

  print(p)

  ggsave(
    filename = file.path(umap_signatures_dir, paste0("14.2-featureplot.spliced_unspliced_re_embed.", score_col, ".png")),
    plot = p, width = 8, height = 8, dpi = 300
  )
}
```

# Spliced/Unspliced Clustree: Module Scores

```{r spliced-clustree-module-scores}
module_scores <- c("EndMT_endocardial1", "EndMT_transition1", "EndMT_mesenchyme1",
                    "EndMT_Notch1", "EndMT_TGFb1",
                    "Metab_glycolysis1", "Metab_oxphos1", "Metab_hypoxia1", "Metab_FAO1")

meta_df <- spliced_subset_obj@meta.data
resolution_cols   <- grep("^res_[0-9]", colnames(meta_df), value = TRUE)
resolution_prefix <- unique(sub("[0-9.]+$", "", resolution_cols))

n_saved <- 0

for (score_col in module_scores) {

  score_name <- sub("1$", "", score_col)

  p <- clustree(meta_df,
                prefix           = resolution_prefix,
                node_colour      = score_col,
                node_colour_aggr = "mean",
                node_size        = "size") +
    ggtitle(score_col) +
    labs(caption = "Endocardium + Mesenchyme subset — spliced/unspliced level")

  print(p)

  ggsave(
    filename = file.path(clustree_dir, paste0(score_name, "_endo_mesen_spliced_unspliced_re_embed.png")),
    plot = p, width = 10, height = 12, dpi = 300
  )

  n_saved <- n_saved + 1
}

message("Saved ", n_saved, " spliced/unspliced module-score clustree PNGs")
```

# Export Spliced/Unspliced Object — RDS

```{r export-spliced-seurat-rds}
write_rds(
  x = spliced_subset_obj,
  file = file.path(out_dir, "all_samples_spliced_endo_mesen_subset.RDS")
)

message("Saved: all_samples_spliced_endo_mesen_subset.RDS")
```

# Export Spliced/Unspliced Object — .h5ad
- Follows the same `sceasy::as.anndata()` approach already established in `11.2`/`10.x`/`12.2`/`13.2`.

```{r export-spliced-h5ad}
chosen_resolution <- 0.4

res_col <- paste0("res_", chosen_resolution)

obj_export <- spliced_subset_obj

Idents(obj_export) <- res_col
Idents(obj_export) <- factor(Idents(obj_export))
DefaultAssay(obj_export) <- "RNA"

obj_export[["RNA"]] <- as(obj_export[["RNA"]], "Assay")

as.anndata(
  obj_export,
  file_path = out_dir,
  file_name = "all_samples_spliced_endo_mesen_subset.h5ad"
)

message("Saved h5ad: all_samples_spliced_endo_mesen_subset.h5ad (Idents = ", res_col, ")")
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "14.2_session_info.txt"))
```
